---
title: "Day 1 — Virtual species with **nicheR**"
subtitle: "ENM-Thailand workshop"
author: "Workshop materials"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: cosmo
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  dpi       = 100
)
```

# Overview

`nicheR` is an R package for building and visualising **ellipsoid-based ecological niches** in environmental space (E-space) and projecting them onto geographic space (G-space). Today we treat the niche as a hypothesis we control — a **virtual species** — so we never need any real occurrence data.

This is the ideal way to learn ENM mechanics, because we can compare every output back to a known truth.

Today we will:

1. Build an ellipsoid from a user-defined range of bioclimatic tolerances.
2. Explore the environmental space of Thailand in **3D** using BIO1, BIO12, and BIO15.
3. Project the ellipsoid onto Thailand bioclim rasters to get suitability and Mahalanobis distance.
4. Add a sampling-bias layer and sample virtual occurrence points using three sampling strategies (`centroid`, `random`, `edge`).

# Install and load packages

The chunk below installs everything Day 1 needs the first time you run it. Re-running is safe — already-installed packages are skipped.

```{r install, eval = FALSE}
# ---- CRAN dependencies -----------------------------------------------------
cran_pkgs <- c(
  "terra",
  "sf",
  "plotly",
  "viridis",
  "rmarkdown",
  "knitr",
  "remotes"
)
to_install <- cran_pkgs[!cran_pkgs %in% rownames(installed.packages())]
if (length(to_install)) install.packages(to_install, dependencies = TRUE)

# ---- nicheR from GitHub ----------------------------------------------------
if (!requireNamespace("nicheR", quietly = TRUE)) {
  remotes::install_github("castanedaM/nicheR", upgrade = "never")
}
```

```{r packages}
library(nicheR)
library(terra)
library(sf)
library(plotly)

# Reproducible random draws throughout the lesson
set.seed(2026)
```

# Data

For Day 1 we work with the WorldClim bioclim rasters and the Thailand polygon. These files are produced by Day 2a (`Day2a_Data_Download.Rmd`) — please knit Day 2a once before Day 1 so the files exist in `data/processed/`.

```{r load-data}
bio_path <- "data/processed/bioclim_thailand.tif"
tha_path <- "data/processed/thailand_boundary.gpkg"

bioclim <- rast(bio_path)
thailand <- st_read(tha_path, quiet = TRUE)

plot(thailand[2])

# Show layer names — bio1 … bio19
names(bioclim)

summary(bioclim)
```

For a clean teaching example we work with three bioclimatic variables. These also drive the 3-D environmental exploration in the next step:

- **bio1** — annual mean temperature (°C, WorldClim v2.1)
- **bio12** — annual precipitation (mm)
- **bio15** — precipitation seasonality (CV)

```{r pick-vars}
env3    <- bioclim[[c("bio1", "bio12", "bio15")]]
env3_df <- as.data.frame(env3, na.rm = TRUE)
plot(env3)
```

# Step 1 — Explore E-space in 3D (BIO1 × BIO12 × BIO15)

Before defining a niche, look at the *available* environmental conditions of Thailand in three dimensions. `BIO15` (precipitation seasonality, CV) captures within-year rainfall variability and adds an important third axis on top of mean temperature (`BIO1`) and total annual precipitation (`BIO12`).

```{r env-3d}
plotly::plot_ly(
  data   = env3_df,
  x      = ~bio1, y = ~bio12, z = ~bio15,
  type   = "scatter3d", mode = "markers",
  marker = list(size = 2, opacity = 0.5)
) |>
  plotly::layout(scene = list(
    xaxis = list(title = "BIO1 (°C)"),
    yaxis = list(title = "BIO12 (mm)"),
    zaxis = list(title = "BIO15 (CV)")
  ))
```

# Step 2 — Define a virtual species' niche

A `nicheR` ellipsoid is built from a **range data frame** that lists the minimum and maximum each environmental variable can take inside the species' niche. The package converts this into a centroid, a covariance matrix, semi-axes, and a niche volume.

Imagine a hypothetical tropical mammal that tolerates mean annual temperatures between **25 and 28 °C** and annual precipitation between **1200 and 2500 mm**.

```{r build-ellipsoid}
range_df <- data.frame(
  bio1  = c(25, 28),
  bio12 = c(1200, 2500),
  bio15 = c(70, 90)
)

ell <- build_ellipsoid(range = range_df)
ell
```

`ell` carries the centroid, covariance matrix, and dimensions of the niche. We can visualise it directly in E-space, with the available environmental conditions of Thailand as background.

```{r plot-ellipsoid}
# `env3_df` from Step 1 carries every pixel's (bio1, bio12, bio15) values —
# we use it as the background point cloud here too.
plot_ellipsoid(ell, background = env3_df)
```

# Step 3 — Predict suitability and Mahalanobis distance in G-space

`predict()` projects the ellipsoid onto a `SpatRaster` and returns both a suitability surface (1 at the centroid, 0 at the boundary, with `truncated = TRUE` clamping outside-niche pixels to zero) and a Mahalanobis distance surface.

```{r predict}
pred <- predict(ell,
                newdata = env3,
                include_suitability = TRUE,
                suitability_truncated = TRUE,
                include_mahalanobis = TRUE,
                mahalanobis_truncated = TRUE,
                keep_data = TRUE)

# pred is a multi-layer SpatRaster carrying "suitability" and "mahalanobis"
pred

plot(pred,
     col  = viridis::viridis(100))

# Binary potential distribution map
binary_rast <- (pred$suitability_trunc > 0) * 1

plot(binary_rast,
     axes = FALSE, box = TRUE,
     col  = c("#d4d4d4", "#0004d5"),
     main = "Potential Distribution (Binary)")
```

# Step 4 — Add a sampling-bias layer

Real-world surveys are biased toward accessible places. We can simulate that bias and apply it on top of suitability.

For this lesson we use species-richness and nighttime-lights rasters (written by `Day2a_Data_Download.Rmd`) as proxies for survey effort.

```{r bias-layer}
sp_rich_layer  <- rast("data/processed/sp_richness_TH.tif")
nighttime_layer <- rast("data/processed/nighttime_TH.tif")
bias_layer <- c(sp_rich_layer, nighttime_layer)
plot(bias_layer)

# Combine bias layers into a 0-1 composite surface
bias_prep <- prepare_bias(bias_surface = bias_layer,
                          effect_direction = c("direct", "inverse"))

plot(bias_prep$composite_surface)
```

```{r apply-bias, fig.height = 4.5}
biased_pred <- apply_bias(prepared_bias = bias_prep,
                          prediction = pred,
                          prediction_layer = "suitability",
                          effect_direction = "direct")

plot(biased_pred$suitability_biased,
     main = "Suitability × bias  =  realised sampling surface")
plot(pred$suitability)
```

# Step 5 — Generate virtual occurrence points with three sampling strategies

`nicheR::sample_data()` accepts three `sampling` strategies:

- `"centroid"` — points weighted toward the niche centroid (preferred conditions).
- `"random"` — points drawn uniformly from inside the niche.
- `"edge"` — points biased toward the niche periphery (marginal conditions).

We draw 50 virtual records under each strategy and compare them.

```{r sampling-three}
pts_centroid <- sample_data(n_occ = 50,
                            prediction = pred,
                            sampling = "centroid",
                            prediction_layer = "suitability",
                            seed = 123)

pts_random   <- sample_data(n_occ = 50,
                            prediction = pred,
                            sampling = "random",
                            prediction_layer = "suitability",
                            seed = 123)

pts_edge     <- sample_data(n_occ = 50,
                            prediction = pred,
                            sampling = "edge",
                            prediction_layer = "suitability",
                            seed = 123)

head(pts_centroid)
head(pts_random)
head(pts_edge)
```

A bias-weighted draw uses `sample_biased_data()` together with the realised sampling surface.

```{r biased-sample}
biased_pts <- sample_biased_data(
  n_occ            = 50,
  prediction       = biased_pred,
  prediction_layer = "suitability_biased_direct",
  seed             = 123
)

head(biased_pts)

occ_bias_env <- terra::extract(env3, biased_pts[, c("x", "y")])
```

```{r plot-samples, fig.height = 6, fig.width = 9}
par(mfrow = c(2, 2))

plot(pred[["suitability"]],
     main = "Centroid sample (n = 50)")
points(pts_centroid[, c("x", "y")], pch = 19, cex = 0.6, col = "red")

plot(pred[["suitability"]],
     main = "Random sample (n = 50)")
points(pts_random[, c("x", "y")], pch = 19, cex = 0.6, col = "red")

plot(pred[["suitability"]],
     main = "Edge sample (n = 50)")
points(pts_edge[, c("x", "y")], pch = 19, cex = 0.6, col = "red")

plot(biased_pred$suitability_biased,
     main = "Bias-weighted sample (n = 50)")
points(biased_pts[, c("x", "y")], pch = 19, cex = 0.6, col = "red")

par(mfrow = c(1, 1))
```

The three strategies populate the niche differently in E-space:

```{r samples-in-espace}
plot_ellipsoid(ell, background = env3_df)
points(pts_centroid[, c("bio1", "bio12")], col = "blue",      pch = 19, cex = 0.6)
points(pts_random[,   c("bio1", "bio12")], col = "darkgreen", pch = 17, cex = 0.6)
points(pts_edge[,     c("bio1", "bio12")], col = "purple",    pch = 15, cex = 0.6)
points(occ_bias_env[, c("bio1", "bio12")], col = "red",       pch = 4,  cex = 0.6)
legend("topright",
       legend = c("Centroid", "Random", "Edge", "Bias-weighted"),
       col    = c("blue", "darkgreen", "purple", "red"),
       pch    = c(19, 17, 15, 4),
       bty    = "n")
```

# Wrap-up

What we just did:

- Defined a niche **a priori** as a 2-D ellipsoid in bioclim space.
- Explored the available environmental space of Thailand in 3D using BIO1, BIO12, BIO15.
- Projected the ellipsoid to Thailand and inspected suitability + Mahalanobis distance.
- Added a sampling-bias surface and contrasted three sampling strategies (`centroid`, `random`, `edge`) plus a bias-weighted draw.
