---
title: "Day 1 — Virtual species with **nicheR**"
subtitle: "ENM-Thailand workshop"
author: "Workshop materials"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: cosmo
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  dpi       = 100
)
```

# Overview

`nicheR` is an R package for building and visualising **ellipsoid-based ecological niches** in environmental space (E-space) and projecting them onto geographic space (G-space). Today we treat the niche as a hypothesis we control — a **virtual species** — so we never need any real occurrence data.

This is the ideal way to learn ENM mechanics, because we can compare every output back to a known truth.

Today we will:

1. Build an ellipsoid from a user-defined range of bioclimatic tolerances.
2. Project the ellipsoid onto Thailand bioclim rasters to get suitability and Mahalanobis distance.
3. Add a sampling-bias layer and sample virtual occurrence points.
4. Simulate a virtual community and map richness.

> Prerequisite: you have run `source("scripts/00_install_packages.R")` and `source("scripts/00_download_data.R")` at least once.

# Packages and data

```{r packages}
library(nicheR)
library(terra)
library(sf)
library(dplyr)
library(ggplot2)
library(viridis)

# Reproducible random draws throughout the lesson
set.seed(2026)
```

We load the cropped WorldClim layers and Thailand polygon written by `00_download_data.R`.

```{r load-data}
bio_path <- "data/processed/bioclim_thailand.tif"
tha_path <- "data/processed/thailand_boundary.gpkg"

bioclim <- terra::rast(bio_path)
thailand <- sf::st_read(tha_path, quiet = TRUE)

# Show layer names — bio1 … bio19
names(bioclim)
```

For a clean teaching example we work with just two bioclimatic variables:

- **bio1** — annual mean temperature (°C × 10 in WorldClim units, but °C in v2.1)
- **bio12** — annual precipitation (mm)

```{r pick-vars}
env <- bioclim[[c("bio1", "bio12")]]
plot(env, main = c("BIO1 — annual mean T (°C)", "BIO12 — annual precip. (mm)"))
```

# Step 1 — Define a virtual species' niche

A `nicheR` ellipsoid is built from a **range data frame** that lists the minimum and maximum each environmental variable can take inside the species' niche. The package converts this into a centroid, a covariance matrix, semi-axes, and a niche volume.

Imagine a hypothetical tropical mammal that tolerates mean annual temperatures between **22 and 28 °C** and annual precipitation between **1500 and 2500 mm**.

```{r build-ellipsoid}
range_df <- data.frame(
  variable = c("bio1", "bio12"),
  min      = c(22,   1500),
  max      = c(28,   2500)
)

ell <- nicheR::ellipsoid(range_data = range_df)
ell
```

`ell` carries the centroid, covariance matrix, and dimensions of the niche. We can visualise it directly in E-space, with the available environmental conditions of Thailand as background.

```{r plot-ellipsoid}
# Pull every pixel's (bio1, bio12) values as the environmental background
env_df <- as.data.frame(env, na.rm = TRUE)

# Open an E-space canvas, add background points, then overlay the ellipsoid
nicheR::plot_ellipsoid(ell,
                       xlim = range(env_df$bio1),
                       ylim = range(env_df$bio12),
                       xlab = "BIO1 — annual mean T (°C)",
                       ylab = "BIO12 — annual precip. (mm)")
nicheR::add_data(env_df, vars = c("bio1", "bio12"),
                 col = adjustcolor("grey60", alpha.f = 0.3), pch = 16, cex = 0.4)
nicheR::add_ellipsoid(ell, col = "tomato", lwd = 2)
title(main = "Virtual niche relative to Thailand's environment")
```

# Step 2 — Predict suitability and Mahalanobis distance in G-space

`predict()` projects the ellipsoid onto a `SpatRaster` and returns both a suitability surface (1 at the centroid, 0 at the boundary, with `truncated = TRUE` clamping outside-niche pixels to zero) and a Mahalanobis distance surface.

```{r predict}
pred <- nicheR::predict(ell, data = env, truncated = TRUE)

# pred is a 2-layer SpatRaster: "suitability" and "mahalanobis"
print(pred)

plot(pred[["suitability"]],
     main = "Virtual species — predicted suitability across Thailand",
     col  = viridis::viridis(100))
plot(sf::st_geometry(thailand), add = TRUE, border = "grey20")
```

```{r predict-maha, fig.height = 4.5}
plot(pred[["mahalanobis"]],
     main = "Mahalanobis distance (NA outside the truncated niche)",
     col  = rev(viridis::magma(100)))
plot(sf::st_geometry(thailand), add = TRUE, border = "grey20")
```

# Step 3 — Add a sampling-bias layer

Real-world surveys are biased toward accessible places. We can simulate that bias and apply it on top of suitability.

For this lesson we use longitude as a proxy for accessibility — many roads in Thailand follow north–south corridors, so western pixels are slightly under-sampled.

```{r bias-layer}
# A toy bias raster: higher values mean higher *detection* probability
bias_layer <- env[[1]]
bias_layer[] <- terra::xFromCell(env, seq_len(terra::ncell(env)))
bias_layer <- terra::mask(bias_layer, env[[1]])
names(bias_layer) <- "longitude_bias"

# Combine bias layers (here, only one) into a 0-1 composite surface
bias_prep <- nicheR::prepare_bias(layers = list(longitude_bias = bias_layer),
                                  directions = c(longitude_bias = "direct"))

plot(bias_prep,
     main = "Composite sampling bias (0 = avoid, 1 = preferred)",
     col  = viridis::cividis(100))
```

```{r apply-bias, fig.height = 4.5}
biased_pred <- nicheR::apply_bias(prediction = pred[["suitability"]],
                                  bias       = bias_prep)

plot(biased_pred,
     main = "Suitability × bias  =  realised sampling surface",
     col  = viridis::viridis(100))
plot(sf::st_geometry(thailand), add = TRUE, border = "grey20")
```

# Step 4 — Generate virtual occurrence points

With the bias-weighted prediction in hand, draw a sample of virtual records. We compare an unbiased draw with a bias-weighted one to see the effect.

```{r sampling}
n <- 150

# Unbiased: pure niche sampling, random across the ellipsoid
unbiased_pts <- nicheR::sample_data(prediction = pred[["suitability"]],
                                    n          = n,
                                    strategy   = "random",
                                    method     = "suitability")

# Biased: weight by the realised sampling surface
biased_pts   <- nicheR::sample_biased_data(prediction = pred[["suitability"]],
                                           bias       = bias_prep,
                                           n          = n,
                                           strategy   = "random",
                                           method     = "suitability")

head(unbiased_pts)
```

```{r plot-samples, fig.height = 5.5}
par(mfrow = c(1, 2))
plot(pred[["suitability"]], col = viridis::viridis(100, alpha = 0.7),
     main = "Unbiased virtual sample (n=150)")
points(unbiased_pts[, c("x", "y")], pch = 19, cex = 0.7, col = "white")

plot(biased_pred, col = viridis::viridis(100, alpha = 0.7),
     main = "Bias-weighted virtual sample (n=150)")
points(biased_pts[, c("x", "y")], pch = 19, cex = 0.7, col = "white")
par(mfrow = c(1, 1))
```

We can also visualise both samples on top of the niche in E-space.

```{r samples-in-espace}
nicheR::plot_ellipsoid(ell,
                       xlim = range(env_df$bio1),
                       ylim = range(env_df$bio12),
                       xlab = "BIO1 (°C)", ylab = "BIO12 (mm)")
nicheR::add_data(env_df, vars = c("bio1", "bio12"),
                 col = adjustcolor("grey80", alpha.f = 0.3), pch = 16, cex = 0.4)
nicheR::add_ellipsoid(ell, col = "tomato", lwd = 2)
points(unbiased_pts[, c("bio1", "bio12")], col = "royalblue", pch = 19, cex = 0.6)
points(biased_pts  [, c("bio1", "bio12")], col = "darkorange", pch = 17, cex = 0.6)
legend("topright", legend = c("Unbiased", "Biased"),
       col = c("royalblue", "darkorange"), pch = c(19, 17), bty = "n")
```

# Step 5 — Simulate a virtual community

`nicheR` can generate a community of virtual species whose niches relate to each other in known ways. We build three nested ellipsoids — useful for teaching beta-diversity ideas — and map the resulting richness.

```{r community}
community <- nicheR::nested_ellipses(
  reference  = ell,
  n_species  = 3,
  shrinkage  = 0.6
)

# Project each species' niche to G-space and stack
comm_pred <- nicheR::predict(community, data = env, truncated = TRUE)

# Build a presence/absence stack at suitability >= 0.05
pa <- terra::app(comm_pred, function(x) as.integer(x >= 0.05))
names(pa) <- paste0("species_", seq_len(terra::nlyr(pa)))

richness <- sum(pa, na.rm = TRUE)
plot(richness,
     main = "Virtual community richness (3 nested species)",
     col  = viridis::turbo(4))
plot(sf::st_geometry(thailand), add = TRUE, border = "grey20")
```

# Wrap-up & exercises

What we just did:

- Defined a niche **a priori** as a 2-D ellipsoid in bioclim space.
- Projected it to Thailand and inspected suitability + Mahalanobis distance.
- Added a sampling-bias surface and contrasted biased vs. unbiased virtual samples.
- Built a nested virtual community and mapped richness.

Try on your own:

1. Add **bio7** (temperature annual range) as a third dimension and rebuild the ellipsoid.
2. Replace `nested_ellipses()` with `random_ellipses()` and compare richness patterns.
3. Increase `n` to 1000 and re-fit the ellipsoid on the *virtual* samples — does it recover the true niche?

```{r session-info}
sessionInfo()
```
