---
title: "Day 2 — Environmental thinning with **bean**"
subtitle: "ENM-Thailand workshop · Sambar deer (*Rusa unicolor*)"
author: "Workshop materials"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: cosmo
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  dpi       = 100
)
```

# Overview

`bean` reduces sampling bias by thinning occurrence points that are **clustered in environmental space (E-space)** instead of in geographic space. The intuition: most field campaigns oversample easily reachable environments. Thinning per environmental "pod" produces a more even coverage of the species' niche, and therefore a less biased fundamental-niche estimate.

Today we will:

1. Load the cleaned GBIF occurrences for *Rusa unicolor* in Thailand.
2. Prepare them for `bean` and visualise the bias in E-space.
3. Find an **objective** environmental grid resolution.
4. Thin the points with both methods (`thin_env_nd` stochastic, `thin_env_center` deterministic).
5. Fit an ellipsoid niche on the thinned points and map suitability back to G-space.

> Prerequisite: you have run the install + download scripts. The Day-0 script wrote the GBIF + bioclim files we use below.

# Packages and data

```{r packages}
library(bean)
library(terra)
library(sf)
library(dplyr)
library(readr)
library(ggplot2)
library(viridis)

set.seed(2026)
```

```{r load-data}
bioclim  <- terra::rast("data/processed/bioclim_thailand.tif")
thailand <- sf::st_read("data/processed/thailand_boundary.gpkg", quiet = TRUE)

sambar <- readr::read_csv("data/processed/sambar_clean_thailand.csv",
                          show_col_types = FALSE)

cat("Cleaned occurrence records in Thailand: ", nrow(sambar), "\n")
head(sambar)
```

For a clean ellipsoid we again use two informative bioclim variables. These also map well onto Sambar deer habitat (warm, wet evergreen and mixed forests).

```{r pick-vars}
env <- bioclim[[c("bio1", "bio12")]]
names(env) <- c("bio1", "bio12")
```

# Step 1 — Prepare occurrences

`prepare_bean()` drops rows with missing coordinates, extracts environmental values at each occurrence, and returns a tidy data frame ready for thinning.

```{r prepare}
sambar_xy <- sambar %>%
  dplyr::select(longitude = decimalLongitude,
                latitude  = decimalLatitude)

prepped <- bean::prepare_bean(
  occurrences = sambar_xy,
  env         = env,
  lon         = "longitude",
  lat         = "latitude"
)

head(prepped)
cat("Records retained after env extraction: ", nrow(prepped), "\n")
```

Look at the raw points in G-space and in E-space side by side. Notice the clustering in E-space — that's the bias `bean` will reduce.

```{r raw-bias, fig.width = 9, fig.height = 4.5}
par(mfrow = c(1, 2))

plot(env[[1]], col = viridis::viridis(100, alpha = 0.7),
     main = "G-space: Sambar deer (raw)")
plot(sf::st_geometry(thailand), add = TRUE, border = "grey20")
points(prepped[, c("longitude", "latitude")], pch = 21, bg = "tomato",
       col = "white", cex = 0.7)

plot(prepped$bio1, prepped$bio12,
     xlab = "BIO1 — annual mean T (°C)", ylab = "BIO12 — annual precip. (mm)",
     pch = 19, col = adjustcolor("tomato", alpha.f = 0.5),
     main = "E-space: Sambar deer (raw)")

par(mfrow = c(1, 1))
```

# Step 2 — Choose an objective environmental resolution

`find_env_resolution()` walks a grid of candidate resolutions, looks at the average nearest-neighbour distance among points after thinning, and identifies the "elbow" where dense artificial clustering ends.

```{r find-res}
res_search <- bean::find_env_resolution(
  data       = prepped,
  vars       = c("bio1", "bio12"),
  resolutions = seq(0.05, 1.0, by = 0.05)  # standardised E-space units
)

res_search$plot
chosen_res <- res_search$resolution
cat("Chosen environmental resolution: ", chosen_res, "\n")
```

# Step 3 — Apply thinning

`bean` ships two thinning strategies; we run both for comparison.

```{r thin}
# Stochastic: randomly retain one occurrence per "pod"
thin_nd <- bean::thin_env_nd(
  data       = prepped,
  vars       = c("bio1", "bio12"),
  resolution = chosen_res
)

# Deterministic: replace each occupied pod with its centre
thin_center <- bean::thin_env_center(
  data       = prepped,
  vars       = c("bio1", "bio12"),
  resolution = chosen_res
)

cat("Raw n               : ", nrow(prepped),    "\n")
cat("Thinned (stochastic): ", nrow(thin_nd),    "\n")
cat("Thinned (centroid)  : ", nrow(thin_center),"\n")
```

```{r thin-vs-raw, fig.width = 9, fig.height = 4.5}
par(mfrow = c(1, 2))
plot(prepped$bio1, prepped$bio12, pch = 19, cex = 0.5,
     col = adjustcolor("grey60", alpha.f = 0.6),
     xlab = "BIO1 (°C)", ylab = "BIO12 (mm)",
     main = "Before thinning")

plot(prepped$bio1, prepped$bio12, pch = 19, cex = 0.5,
     col = adjustcolor("grey60", alpha.f = 0.25),
     xlab = "BIO1 (°C)", ylab = "BIO12 (mm)",
     main = "After thinning (stochastic)")
points(thin_nd[, c("bio1", "bio12")], pch = 21, bg = "tomato",
       col = "white", cex = 1)
par(mfrow = c(1, 1))
```

# Step 4 — Fit the ellipsoid niche

`fit_ellipsoid()` formalises the niche by fitting a bivariate (or multivariate) ellipsoid around the thinned points. We compare the ellipsoid fit on raw vs. thinned data.

```{r fit-ellipsoid}
ell_raw      <- bean::fit_ellipsoid(thin_center, vars = c("bio1", "bio12"))
ell_thinned  <- bean::fit_ellipsoid(thin_nd,     vars = c("bio1", "bio12"))

ell_thinned
```

```{r ellipsoid-overlay, fig.height = 5.5}
plot(prepped$bio1, prepped$bio12, pch = 19, cex = 0.4,
     col = adjustcolor("grey70", alpha.f = 0.4),
     xlab = "BIO1 (°C)", ylab = "BIO12 (mm)",
     main = "Raw vs. thinned ellipsoid (Sambar deer)")
# Overlay both ellipsoids using bean's plot method
lines(ellipse::ellipse(ell_raw$cov,     centre = ell_raw$center),
      col = "grey30", lwd = 2, lty = 2)
lines(ellipse::ellipse(ell_thinned$cov, centre = ell_thinned$center),
      col = "tomato", lwd = 2)
legend("topright",
       legend = c("Raw", "Thinned"),
       col    = c("grey30", "tomato"),
       lty    = c(2, 1), lwd = 2, bty = "n")
```

> If `ellipse` is not installed, run `install.packages("ellipse")`. We use it here only for an overlay comparison; `bean` ships its own plotting helpers as well.

# Step 5 — Project the niche back to G-space

```{r predict}
pred_raw     <- predict(ell_raw,     env)
pred_thinned <- predict(ell_thinned, env)

par(mfrow = c(1, 2))
plot(pred_raw,     main = "Suitability — raw niche",
     col = viridis::viridis(100))
plot(sf::st_geometry(thailand), add = TRUE, border = "grey20")
plot(pred_thinned, main = "Suitability — thinned niche",
     col = viridis::viridis(100))
plot(sf::st_geometry(thailand), add = TRUE, border = "grey20")
par(mfrow = c(1, 1))
```

```{r save}
dir.create("outputs", showWarnings = FALSE)
terra::writeRaster(pred_thinned,
                   "outputs/Day2_suitability_sambar_thinned.tif",
                   overwrite = TRUE)
saveRDS(ell_thinned, "outputs/Day2_ellipsoid_thinned.rds")
```

# Wrap-up & exercises

Take-homes:

- Spatial bias in occurrence data **propagates into E-space** as dense clusters.
- `bean` thins in E-space so that the fitted niche is not pulled toward the densest sampled environments.
- The two thinning methods (`thin_env_nd` vs. `thin_env_center`) trade randomness for reproducibility — pick one and stick with it for a given analysis.

Try on your own:

1. Add **bio7** (temperature annual range) as a third variable to the niche.
2. Vary the `resolution` argument by ±50 % around the elbow value — how stable is the resulting ellipsoid?
3. Compute the Schoener's-D overlap between the raw and thinned predictions (e.g. with `ENMTools`) to quantify the bias correction.

```{r session-info}
sessionInfo()
```
