---
title: "Day 2b — Environmental thinning with **bean**"
subtitle: "ENM-Thailand workshop · Sambar deer (*Rusa unicolor*)"
author: "Workshop materials"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: cosmo
    df_print: paged
editor_options:
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  dpi       = 100
)
```

# Overview

`bean` reduces sampling bias by thinning occurrence points that are **clustered in environmental space (E-space)** instead of in geographic space. The intuition: most field campaigns oversample easily reachable environments. Thinning per environmental "pod" produces a more even coverage of the species' niche, and therefore a less biased fundamental-niche estimate.

Today we will:

1. Load the cleaned GBIF occurrences for *Rusa unicolor* in Thailand (written by `Day2a_Data_Download.Rmd`).
2. Prepare them for `bean` and visualise the bias in E-space.
3. Find an **objective** environmental grid resolution.
4. Thin the points with both methods (`thin_env_nd` stochastic, `thin_env_center` deterministic).
5. Fit an ellipsoid niche on the thinned points and map suitability back to G-space.

> Prerequisite: knit `Day2a_Data_Download.Rmd` first so `data/processed/` is populated.

# Install and load packages

```{r install, eval = FALSE}
cran_pkgs <- c(
  "terra",
  "sf",
  "dplyr",
  "readr",
  "rmarkdown",
  "knitr",
  "remotes"
)
to_install <- cran_pkgs[!cran_pkgs %in% rownames(installed.packages())]
if (length(to_install)) install.packages(to_install, dependencies = TRUE)

# ---- bean from GitHub ------------------------------------------------------
if (!requireNamespace("bean", quietly = TRUE)) {
  remotes::install_github("paanwaris/bean", upgrade = "never")
}
```

```{r packages}
library(bean)
library(terra)
library(sf)
library(dplyr)
library(readr)

set.seed(2026)
```

```{r load-data}
bioclim  <- rast("data/processed/bioclim_thailand.tif")
thailand <- st_read("data/processed/thailand_boundary.gpkg", quiet = TRUE)

sambar <- read_csv("data/processed/sambar_thailand.csv",
                   show_col_types = FALSE)

head(sambar)
```

For a clean ellipsoid we use two informative bioclim variables. These also map well onto Sambar deer habitat (warm, wet evergreen and mixed forests).

```{r pick-vars}
env <- bioclim[[c("bio1", "bio12", "bio15")]]
names(env) <- c("bio1", "bio12", "bio15")
```

# Step 1 — Prepare occurrences

`prepare_bean()` drops rows with missing coordinates, extracts environmental values at each occurrence, and returns a tidy data frame ready for thinning.

```{r prepare}
sambar_xy <- sambar %>%
  dplyr::select(longitude = decimalLongitude,
                latitude  = decimalLatitude)

prepped <- prepare_bean(data        = sambar_xy,
                        env_rasters = env,
                        longitude   = "longitude",
                        latitude    = "latitude",
                        transform   = "none")

nrow(prepped)
```

Look at the raw points in G-space and in E-space side by side. Notice the clustering in E-space — that's the bias `bean` will reduce.

```{r raw-bias, fig.width = 9, fig.height = 4.5}
par(mfrow = c(1, 2))

plot(st_geometry(thailand), border = "grey20", main = "G-space")
points(prepped[, c("longitude", "latitude")], pch = 21, bg = "tomato",
       col = "white", cex = 0.7)

plot(prepped$bio1, prepped$bio12,
     xlab = "BIO1 — annual mean T (°C)", ylab = "BIO12 — annual precip. (mm)",
     pch = 19, col = adjustcolor("tomato", alpha.f = 0.5),
     main = "E-space")

par(mfrow = c(1, 1))
```

# Step 2 — Choose an objective environmental resolution

`find_env_resolution()` walks a grid of candidate resolutions, looks at the average nearest-neighbour distance among points after thinning, and identifies the "elbow" where dense artificial clustering ends.

```{r find-res}
res_search <- find_env_resolution(data     = prepped,
                                  env_vars = c("bio1", "bio12", "bio15"))

plot(res_search)

chosen_res <- res_search$suggested_resolution
chosen_res
```

# Step 3 — Apply thinning

`bean` ships two thinning strategies; we run both for comparison.

```{r thin}
# Stochastic: randomly retain one occurrence per "pod"
thin_nd <- thin_env_nd(data            = prepped,
                       env_vars        = c("bio1", "bio12", "bio15"),
                       grid_resolution = chosen_res)

# Deterministic: replace each occupied pod with its centre
thin_center <- thin_env_center(data            = prepped,
                               env_vars        = c("bio1", "bio12", "bio15"),
                               grid_resolution = c(0.5, 50, 20))

thin_nd
thin_center
```

```{r thin-vs-raw, fig.width = 9, fig.height = 4.5}
plot_bean(
  original_data  = prepped,
  thinned_object = thin_nd,
  env_vars       = c("bio1", "bio12", "bio15")
)

plot_bean(
  original_data  = prepped,
  thinned_object = thin_center,
  env_vars       = c("bio1", "bio12", "bio15")
)
```

# Step 4 — Fit the ellipsoid niche

`fit_ellipsoid()` formalises the niche by fitting a bivariate (or multivariate) ellipsoid around the thinned points. We compare the ellipsoid fit on raw vs. thinned data.

```{r fit-ellipsoid}
ell_raw <- fit_ellipsoid(prepped,
                         env_vars = c("bio1", "bio12", "bio15"),
                         method   = "covmat",
                         level    = 0.95)

ell_thinned <- fit_ellipsoid(thin_nd$thinned_data,
                             env_vars = c("bio1", "bio12", "bio15"),
                             method   = "covmat",
                             level    = 0.95)

plot(ell_raw, dims = c(1, 2))
plot(ell_raw, dims = c(1, 2, 3))

plot(ell_thinned, dims = c(1, 2))
plot(ell_thinned, dims = c(1, 2, 3))
```

# Step 5 — Project the niche back to G-space

```{r predict}
pred_raw <- predict(ell_raw, env)
pred_thinned <- predict(ell_thinned, env)

par(mfrow = c(1, 2))
plot(pred_raw$suitability, main = "Suitability — raw data")
plot(pred_thinned$suitability, main = "Suitability — thinned niche")
par(mfrow = c(1, 1))
```

```{r save}
dir.create("outputs", showWarnings = FALSE)
terra::writeRaster(pred_thinned,
                   "outputs/Day2_suitability_sambar_thinned.tif",
                   overwrite = TRUE)
saveRDS(ell_thinned, "outputs/Day2_ellipsoid_thinned.rds")
```

# Wrap-up

Take-homes:

- Spatial bias in occurrence data **propagates into E-space** as dense clusters.
- `bean` thins in E-space so that the fitted niche is not pulled toward the densest sampled environments.
- The two thinning methods (`thin_env_nd` vs. `thin_env_center`) trade randomness for reproducibility — pick one and stick with it for a given analysis.