---
title: "Day 3 — Temporally-explicit SDMs with **TemporalModelR**"
subtitle: "ENM-Thailand workshop · Sambar deer (*Rusa unicolor*)"
author: "Workshop materials"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: cosmo
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  dpi       = 100
)
```

# Overview

`TemporalModelR` builds **temporally-explicit** species distribution models. The key idea: each occurrence is paired with the environmental conditions *at the time it was recorded*, rather than against a long-term average. Models can then project predictions as time-series surfaces and detect pixel-level temporal trajectories (always-suitable, increasing, decreasing, fluctuating, etc.).

> **Important note on data.** Truly temporally-resolved bioclimatic layers (one raster per year) are large and slow to download. To keep the workshop reproducible we work with the **static** WorldClim bioclim that we cropped on Day 0. We give every raster the *same* values across years, which means the **temporal signal in our toy example comes entirely from the year associated with each observation, not from the environment itself**. The pipeline you learn here is identical to the one you would use with truly time-varying layers (e.g. CHELSA-TS, ERA5, MODIS); you would just point `pattern` at one raster per year.

Today we will:

1. Build a stack of "annual" bioclim layers (10-min resolution, Thailand) — one copy per year.
2. Align them, rarefy Sambar deer occurrences in space *and* time, and extract environment.
3. Build spatiotemporal cross-validation folds and fold-stratified pseudoabsences.
4. Fit a `build_temporal_glm()` model and a `build_temporal_rf()` model.
5. Project predictions across multiple years and run the postprocessing pipeline (consensus, changepoints, regional summaries).

# Packages and data

```{r packages}
library(TemporalModelR)
library(terra)
library(sf)
library(dplyr)
library(readr)
library(ggplot2)
library(viridis)

set.seed(2026)
```

```{r load-data}
bioclim  <- terra::rast("data/processed/bioclim_thailand.tif")
thailand <- sf::st_read("data/processed/thailand_boundary.gpkg", quiet = TRUE)

sambar <- readr::read_csv("data/processed/sambar_clean_thailand.csv",
                          show_col_types = FALSE) %>%
  dplyr::filter(year >= 2000, year <= 2020) %>%
  dplyr::rename(longitude = decimalLongitude,
                latitude  = decimalLatitude)

cat("Occurrences after year filter:", nrow(sambar), "\n")
table(sambar$year)
```

# Step 1 — Build the temporally-explicit raster stack

`TemporalModelR` matches occurrence records to rasters via **time placeholders in filenames**. We pick three predictors (BIO1, BIO12, BIO15) and write one copy per year. In a real study these copies would actually differ year-by-year — here they are duplicates of the static WorldClim values purely to demonstrate the workflow.

```{r write-temporal-rasters}
dir.create("data/processed/temporal_rasters", showWarnings = FALSE)

years     <- 2000:2020
variables <- c("bio1", "bio12", "bio15")

for (v in variables) {
  layer <- bioclim[[v]]
  for (y in years) {
    out <- file.path("data/processed/temporal_rasters",
                     sprintf("%s_%d.tif", v, y))
    if (!file.exists(out)) {
      terra::writeRaster(layer, filename = out, overwrite = TRUE)
    }
  }
}

list.files("data/processed/temporal_rasters") |> head()
```

`patterns` tells `TemporalModelR` how to expand a placeholder. Below, the placeholder `<year>` is replaced with each year.

```{r patterns}
patterns <- list(
  bio1_year  = "data/processed/temporal_rasters/bio1_<year>.tif",
  bio12_year = "data/processed/temporal_rasters/bio12_<year>.tif",
  bio15_year = "data/processed/temporal_rasters/bio15_<year>.tif"
)

# time_cols maps placeholders to occurrence-table columns
time_cols <- list("<year>" = "year")
```

# Step 2 — Preprocessing

## Align rasters

`raster_align()` reprojects and resamples every raster to a common grid. Our copies are already aligned (they came from the same source), but we run the function to demonstrate the pattern.

```{r raster-align}
aligned <- TemporalModelR::raster_align(
  patterns     = patterns,
  time_values  = list("<year>" = years),
  reference    = bioclim[[1]]
)
```

## Spatio-temporal rarefaction

Keep at most one record per pixel **per year** (so a repeated visit to the same place in a later year is preserved as a temporally-independent observation).

```{r rarefy}
rarefied <- TemporalModelR::spatiotemporal_rarefaction(
  occurrences = sambar,
  raster_ref  = bioclim[[1]],
  time_cols   = list(year = "year"),
  lon = "longitude", lat = "latitude"
)

cat("Rarefied n: ", nrow(rarefied), "\n")
```

## Match each observation to the environment it experienced

```{r extract}
extracted <- TemporalModelR::temporally_explicit_extraction(
  occurrences = rarefied,
  patterns    = patterns,
  time_cols   = time_cols,
  lon = "longitude", lat = "latitude"
)

extracted$data |> head()
```

## Scale rasters (z-score)

```{r scale}
scaled_rasters <- TemporalModelR::scale_rasters(
  patterns     = patterns,
  time_values  = list("<year>" = years),
  scaling_pars = extracted$scaling
)
```

## Cross-validation folds & pseudoabsences

We use **spatiotemporal** folds (blocks split in both space and time) — the most defensible choice for temporal SDMs.

```{r folds}
partitioned <- TemporalModelR::spatiotemporal_partition(
  data    = extracted$data,
  method  = "spatiotemporal",
  k       = 4,
  lon     = "longitude", lat = "latitude",
  time_cols = "year"
)

absences <- TemporalModelR::generate_absences(
  partition_result = partitioned,
  raster_ref       = bioclim[[1]],
  n_per_presence   = 1,
  patterns         = patterns,
  time_cols        = time_cols
)
```

# Step 3 — Modeling

We compare two algorithms; both share the same input objects.

## GLM (with a quadratic term on temperature)

```{r glm}
glm_res <- TemporalModelR::build_temporal_glm(
  partition_result    = partitioned,
  pseudoabsence_result = absences,
  model_formula = presence ~ bio1_year + I(bio1_year^2) + bio12_year + bio15_year
)

glm_res$evaluation
```

## Random forest

```{r rf}
rf_res <- TemporalModelR::build_temporal_rf(
  partition_result    = partitioned,
  pseudoabsence_result = absences,
  model_vars = c("bio1_year", "bio12_year", "bio15_year"),
  ntree      = 500
)

rf_res$evaluation
```

# Step 4 — Project predictions through time

Generate one map per year for both models.

```{r predict}
time_steps <- data.frame(year = years)

glm_preds <- TemporalModelR::generate_spatiotemporal_predictions(
  model_result = glm_res,
  rasters      = scaled_rasters,
  time_steps   = time_steps,
  patterns     = patterns,
  time_cols    = time_cols,
  output_dir   = "outputs/glm_predictions"
)

rf_preds <- TemporalModelR::generate_spatiotemporal_predictions(
  model_result = rf_res,
  rasters      = scaled_rasters,
  time_steps   = time_steps,
  patterns     = patterns,
  time_cols    = time_cols,
  output_dir   = "outputs/rf_predictions"
)
```

Quick look at three years for the RF model:

```{r plot-years, fig.width = 9, fig.height = 4}
sample_years <- c(2000, 2010, 2020)
rf_stack <- terra::rast(file.path("outputs/rf_predictions",
                                  sprintf("fold1_year_%d.tif", sample_years)))

par(mfrow = c(1, 3))
for (i in seq_along(sample_years)) {
  plot(rf_stack[[i]],
       main = paste("RF — predicted suitability,", sample_years[i]),
       col  = viridis::viridis(100), range = c(0, 1))
  plot(sf::st_geometry(thailand), add = TRUE, border = "grey20")
}
par(mfrow = c(1, 1))
```

# Step 5 — Postprocessing

## Consensus surfaces across folds

```{r consensus}
consensus <- TemporalModelR::summarize_raster_outputs(
  prediction_result  = rf_preds,
  consensus_threshold = 3   # at least 3 of 4 folds must agree
)

plot(consensus$frequency,
     main = "Proportion of time steps each pixel was suitable (RF)",
     col  = viridis::magma(100))
plot(sf::st_geometry(thailand), add = TRUE, border = "grey20")
```

## Pixel-level temporal trajectories (changepoint analysis)

`analyze_temporal_patterns()` runs `fastcpd` per pixel and classifies trajectories as never-suitable, always-suitable, no pattern, increasing, decreasing, or fluctuating.

```{r changepoints}
patterns_out <- TemporalModelR::analyze_temporal_patterns(
  prediction_result  = rf_preds,
  consensus_result   = consensus,
  spatial_autocorr   = FALSE
)

plot(patterns_out$classification,
     main = "Temporal trajectory class per pixel",
     type = "classes")
```

## Aggregate by spatial unit (Thailand provinces)

```{r provinces, eval = FALSE}
provinces <- geodata::gadm(country = "THA", level = 1, path = "data/raw") |>
  sf::st_as_sf()

regional <- TemporalModelR::analyze_trends_by_spatial_unit(
  pattern_result = patterns_out,
  polygons       = provinces,
  id_field       = "NAME_1"
)

head(regional)
```

> The province step is set to `eval = FALSE` so the document still knits if you haven't fetched GADM level-1 yet. Set it to `TRUE` after running the line manually.

# Wrap-up & exercises

What we covered:

- Built a fully temporally-explicit pipeline: time-stamped occurrences ↔ time-stamped rasters.
- Used spatio-temporal cross-validation folds and fold-stratified pseudoabsences.
- Fit GLM and RF temporal models and projected through time.
- Classified pixel-level temporal trajectories and aggregated by region.

Try on your own:

1. Swap the static-clone rasters for **real** annual layers (e.g. CHELSA-TS bio1 1981–2020). The only thing that changes is the file path in `patterns`.
2. Fit a `build_temporal_gam()` and compare its E-space evaluation metrics to the GLM and RF.
3. Repeat the analysis with **monthly** time steps using `expand.grid(year = years, month = sprintf("%02d", 1:12))`.

```{r session-info}
sessionInfo()
```
