---
title: "Day 3 — Temporally-explicit SDMs with **TemporalModelR**"
subtitle: "ENM-Thailand workshop · Sambar deer (*Rusa unicolor*)"
author: "Workshop materials"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: cosmo
    df_print: paged
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  dpi       = 100
)
```

# Overview

`TemporalModelR` builds **temporally-explicit** species distribution models. Each occurrence is paired with the environmental conditions *at the time it was recorded*, rather than against a long-term average. Models can then project predictions as time-series surfaces and detect pixel-level temporal trajectories (always-suitable, increasing, decreasing, fluctuating, etc.).

> **Data.** Day 3 uses the pre-prepared annual rasters that already live in `temporal_rasters/` — one precipitation (`prec_YYYY.tif`) and one temperature (`temp_YYYY.tif`) file per year for 2000–2025. **We do not download new climate data here**; we simply read the local files and stack them. Day 2a produced the Sambar GBIF table and the Thailand polygon we use alongside them.

Today we will:

1. Read the annual precipitation and temperature rasters (2000–2025) from `temporal_rasters/` and stack them.
2. Align them to a common grid with `raster_align()`, rarefy occurrences with `spatiotemporal_rarefaction()`, and extract time-matched environment with `temporally_explicit_extraction()`.
3. Scale rasters, build spatiotemporal CV folds, and create fold-stratified pseudoabsences.
4. Fit `build_temporal_glm()` and `build_temporal_rf()` models and project predictions across years with `generate_spatiotemporal_predictions()`.
5. Postprocess with `summarize_raster_outputs()`, `analyze_temporal_patterns()`, and `analyze_trends_by_spatial_unit()`.

All function names and arguments below follow the official package docs:
<https://cjhughes926.github.io/TemporalModelR/articles/Preprocessing.html>

# Install and load packages

```{r install, eval = FALSE}
cran_pkgs <- c(
  "terra",
  "sf",
  "dplyr",
  "readr",
  "viridis",
  "randomForest",
  "fastcpd",
  "rmarkdown",
  "knitr",
  "remotes",
  "geodata"
  # "geodata"   # uncomment only if you enable the optional `provinces` chunk
)
to_install <- cran_pkgs[!cran_pkgs %in% rownames(installed.packages())]
if (length(to_install)) install.packages(to_install, dependencies = TRUE)

# ---- TemporalModelR from GitHub --------------------------------------------
if (!requireNamespace("TemporalModelR", quietly = TRUE)) {
  remotes::install_github("CJHughes926/TemporalModelR")
}
```

```{r packages}
library(TemporalModelR)
library(terra)
library(sf)
library(dplyr)
library(readr)

set.seed(2026)
```

# Inputs

```{r load-base}
thailand <- st_read("data/processed/thailand_boundary.gpkg", quiet = TRUE)

sambar <- read_csv("data/processed/sambar_thailand.csv",
                   show_col_types = FALSE) %>%
  filter(year >= 2010, year <= 2025) %>%
  rename(longitude = decimalLongitude,
                latitude  = decimalLatitude) %>%
  mutate(presence = 1L)

table(sambar$year)
```

# Step 1 — Preprocessing

## Align rasters

`raster_align()` reprojects/resamples every raster in `input_dir` against a reference and writes the aligned copies to `output_dir`.

```{r raster-align}
prec_ex <- rast("data/processed/temporal_split/prec_2025.tif")
temp_ex <- rast("data/processed/temporal_split/temp_2025.tif")
NDVI_ex <- rast("data/processed/temporal_split/NDVI_2025.tif")

par(mfrow = c(1, 3))
plot(prec_ex, main = "Precipitaion 2025")
plot(temp_ex, main = "Temperature 2025")
plot(NDVI_ex, main = "NDVI 2025")

split_dir <- "data/processed/temporal_split"
aligned_dir <- "data/processed/rasters_aligned"
dir.create(aligned_dir, showWarnings = FALSE, recursive = TRUE)

ref_raster <- NDVI_ex  # any one of the per-variable split rasters

raster_align(
  input_dir = split_dir,
  output_dir = aligned_dir,
  reference_raster = ref_raster,
  overwrite = TRUE
)

par(mfrow = c(1, 1))
```

## Spatio-temporal rarefaction

Keep at most one record per pixel **per year**, so a repeated visit to the same place in a later year is preserved as a temporally-independent observation.

```{r rarefy}
points_dir <- "data/processed/points"
dir.create(points_dir, showWarnings = FALSE, recursive = TRUE)

points_csv <- file.path(points_dir, "sambar_temporal.csv")
write_csv(sambar, points_csv)

study_crs <- st_crs(rast(ref_raster))

rare_out <- spatiotemporal_rarefaction(
  points_sp = points_csv,
  output_dir = file.path(points_dir, "rarefied"),
  reference_raster = ref_raster,
  time_cols = "year",
  xcol = "longitude",
  ycol = "latitude",
  points_crs = study_crs,
  output_prefix = "Sambar_annual"
)

rare_out$input_points
rare_out$spatial_points
rare_out$spatiotemporal_points
```

## Match each observation to the environment it experienced

`temporally_explicit_extraction()` walks the aligned raster directory using `variable_patterns`: clean names on the left, filename patterns with time placeholders (`YEAR`) on the right.

```{r extract}
ext_dir <- file.path(points_dir, "extracted")

ext_out <- temporally_explicit_extraction(
  points_sp = rare_out$files_created$spatiotemporal,
  raster_dir = aligned_dir,
  variable_patterns   = c(
    "prec" = "prec_YEAR",
    "temp" = "temp_YEAR",
    "NDVI" = "NDVI_YEAR"
  ),
  time_cols = "year",
  xcol = "X",
  ycol = "Y",
  points_crs = study_crs,
  output_dir = ext_dir,
  output_prefix = "sambar_extracted",
  save_raw = TRUE,
  save_scaled = TRUE,
  save_scaling_params = TRUE
)

head(ext_out$raw_values)
ext_out$scaling_params
```

## Scale rasters

`scale_rasters()` z-scores every raster in `input_dir` using the per-variable means and SDs from extraction, so the rasters and the points share the same transformation.

```{r scale}
scaled_dir <- "data/processed/rasters_scaled"
dir.create(scaled_dir, showWarnings = FALSE, recursive = TRUE)

scale_rasters(
  input_dir = aligned_dir,
  output_dir = scaled_dir,
  scaling_params_file = ext_out$files_created$scaling_params,
  variable_patterns = c(
    "prec" = "prec_YEAR",
    "temp" = "temp_YEAR",
    "NDVI" = "NDVI_YEAR"
  ),
  time_cols = "year",
  overwrite = TRUE
)
```

## Spatiotemporal partitioning

```{r partition}
par(mfrow = c(1, 3))
partition <- spatiotemporal_partition(
  reference_shapefile_path = thailand,
  points_file_path = ext_out$files_created$scaled,
  xcol = "x",
  ycol = "y",
  points_crs = study_crs,
  time_cols = "year",
  n_spatial_folds = 2,
  n_temporal_folds = 2,
  max_imbalance = 0.15,
  create_plot = TRUE
)

partition$summary
par(mfrow = c(1, 1))
```

## Generate pseudoabsences

```{r absences}
absences <- generate_absences(
  partition_result = partition,
  reference_shapefile_path = thailand,
  raster_dir = scaled_dir,
  variable_patterns = c(
    "prec" = "prec_YEAR",
    "temp" = "temp_YEAR",
    "NDVI" = "NDVI_YEAR"
  ),
  method = "buffer",
  buffer_distance = 50000,
  # 10 km — coarser pixels need a bigger buffer
  ratio = 2,
  time_cols = "year",
  create_plot = TRUE,
  plot_by_fold = TRUE,
  verbose = FALSE
)

absences$summary
```

# Step 2 — Modeling

## GLM (with a quadratic term on temperature)

```{r glm}
glm_res <- build_temporal_glm(
  partition_result = partition,
  pseudoabsence_result = absences,
  model_formula = ~ temp + I(temp^2) + prec + NDVI,
  link = "logit",
  threshold_method = "tss",
  output_dir = "outputs/GLM_Models",
  create_plot = FALSE,
  overwrite = TRUE,
  time_cols = "year",
  verbose = FALSE
)

glm_res$fold_test_metrics
```

## Random forest

```{r rf}
rf_res <- build_temporal_rf(
  partition_result = partition,
  pseudoabsence_result = absences,
  model_vars = c("temp", "prec", "NDVI"),
  rf_params = list(ntree = 500),
  threshold_method = "tss",
  output_dir = "outputs/RF_Models",
  create_plot = FALSE,
  overwrite = TRUE,
  time_cols = "year"
)

rf_res$fold_test_metrics
```

# Step 3 — Project predictions through time

```{r predict}
time_steps <- data.frame(year = c(2010:2025))

glm_preds <- generate_spatiotemporal_predictions(
  partition_result = partition,
  model_result = glm_res,
  pseudoabsence_result = absences,
  raster_dir = scaled_dir,
  variable_patterns = c("prec" = "prec_YEAR",
                        "temp" = "temp_YEAR",
                        "NDVI" = "NDVI_YEAR"), 
  time_cols = "year",
  time_steps = time_steps,
  output_dir = "outputs/glm_predictions",
  overwrite = TRUE
)

rf_preds <- generate_spatiotemporal_predictions(
  partition_result = partition,
  model_result = rf_res,
  pseudoabsence_result = absences,
  raster_dir = scaled_dir,
  variable_patterns = c("prec" = "prec_YEAR",
                        "temp" = "temp_YEAR",
                        "NDVI" = "NDVI_YEAR"), 
  time_cols = "year",
  time_steps = time_steps,
  output_dir = "outputs/rf_predictions",
  overwrite = TRUE
)
```

Quick look at three years for the RF model:

```{r plot-years, fig.width = 9, fig.height = 4}
sample_years <- c(2000, 2010, 2015, 2020, 2025)
rf_files <- list.files("outputs/rf_predictions",
                       pattern = paste0("Prediction_.*(",
                                        paste(sample_years, collapse = "|"),
                                        ").*\\.tif$"),
                       full.names = TRUE)

rf_stack <- terra::rast(rf_files[seq_len(min(5, length(rf_files)))])

par(mfrow = c(1, terra::nlyr(rf_stack)))
for (i in seq_len(terra::nlyr(rf_stack))) {
  plot(rf_stack[[i]],
       main = names(rf_stack)[i])
  plot(sf::st_geometry(thailand), add = TRUE, border = "grey20")
}
par(mfrow = c(1, 1))
```

# Step 4 — Postprocessing

## Consensus surfaces across folds

`summarize_raster_outputs()` reads the per-time-step vote-count rasters from the prediction directory and applies a consensus threshold.

```{r consensus}
consensus <- summarize_raster_outputs(
  predictions_dir = "outputs/rf_predictions",
  output_dir = "outputs/rf_consensus",
  overwrite = TRUE
)

plot(consensus$frequency_raster,
     main = "Proportion of time steps each pixel was suitable (RF)")
```

## Pixel-level temporal trajectories (changepoint analysis)

`analyze_temporal_patterns()` runs `fastcpd` per pixel and classifies trajectories as never-suitable, always-suitable, no-pattern, increasing, decreasing, or fluctuating.

```{r changepoints}
patterns_out <- analyze_temporal_patterns(
  binary_stack = consensus$consensus_stack,
  summary_raster = consensus$frequency_raster,
  time_steps = time_steps,
  output_dir = "outputs/rf_patterns",
  spatial_autocorrelation = FALSE,
  estimate_time = FALSE,
  overwrite = TRUE
)

plot(patterns_out$pattern,
     main = "Temporal trajectory class per pixel",
     type = "classes")
```

## Aggregate by spatial unit (Thailand provinces)

```{r provinces}
provinces <- geodata::gadm(country = "THA", level = 1, path = "data/raw") |>
  st_as_sf()

regional <- analyze_trends_by_spatial_unit(
  shapefile_path = provinces,
  name_field = "NAME_1",
  binary_stack = consensus$consensus_stack,
  pattern_raster = patterns_out$pattern,
  time_decrease_raster = patterns_out$time_decrease,
  time_increase_raster = patterns_out$time_increase,
  time_steps = time_steps,
  output_dir = "outputs/rf_regional",
  overwrite = TRUE,
  create_plot = TRUE
)

head(regional$overall_summary)
regional$plots$time_series
regional$plots$annual_change
regional$plots$total_change_by_unit
```

# Wrap-up & exercises

What we covered:

- Read pre-prepared annual precipitation and temperature rasters (2000–2025) directly from `temporal_rasters/` and stacked them — **no new downloads**.
- Built a fully temporally-explicit pipeline: time-stamped occurrences ↔ time-stamped rasters.
- Used spatio-temporal cross-validation folds and fold-stratified pseudoabsences.
- Fit GLM and RF temporal models and projected through time.
- Classified pixel-level temporal trajectories and aggregated by region.

```{r session-info}
sessionInfo()
```
