---
title: "Pre-workshop — Downloading and preparing the workshop data"
subtitle: "ENM-Thailand workshop · Sambar deer (*Rusa unicolor*)"
author: "Workshop materials"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: cosmo
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  dpi       = 100
)

# This is the Pre-workshop notebook itself, which actually downloads data from
# the internet. When the *public* pkgdown website is built we DO NOT want to
# re-hit GADM / WorldClim / GBIF every time — we just want the code to be
# visible on the website. The environment variable `PKGDOWN_BUILD` is set to
# "1" by `scripts/build_pkgdown_site.R`; we detect that here and switch chunk
# evaluation off when it is set.
.fetch_live <- !nzchar(Sys.getenv("PKGDOWN_BUILD"))
knitr::opts_chunk$set(eval = .fetch_live)
```

```{r data-status, echo = FALSE, results = "asis", eval = TRUE}
if (!.fetch_live) {
  cat(
    '<div class="alert alert-info" style="padding:1em;border-left:4px solid #1B6E3F;background:#E8F5E9;">',
    '<strong>Build mode:</strong> the live download code is <em>shown but ',
    'not executed</em> on the published website. Knit this file locally with ',
    'your GBIF credentials in <code>.Renviron</code> to actually populate ',
    '<code>data/processed/</code>.',
    '</div>\n\n'
  )
}
```

# Overview

> **Workshop navigation**
>
> - **Pre-workshop — Downloading the workshop data.** Get the Thailand boundary, WorldClim climate layers, and GBIF Sambar records before the workshop starts.
> - [**Step 1: nicheR** — Virtual species with nicheR.](https://paanwaris.github.io/ENM-Thailand/articles/Step1_nicheR.html) Build a virtual species niche and map it.
> - [**Step 2: bean** — Environmental thinning with bean.](https://paanwaris.github.io/ENM-Thailand/articles/Step2_bean.html) Reduce environmental sampling bias before fitting a real niche.
> - [**Step 3: TemporalModelR** — Temporally-explicit models with TemporalModelR.](https://paanwaris.github.io/ENM-Thailand/articles/Step3_TemporalModelR.html) Link species occurrences with environmental data through time.

This guide collects every dataset that the rest of the workshop will read. **Please knit it once before the workshop starts** — **Step 1**, **Step 2**, and **Step 3** all read from the `data/processed/` folder, so running this notebook ahead of time makes sure every workshop session will run smoothly.

By the end of this Pre-workshop you will have downloaded and prepared four datasets:

1.  **Thailand national boundary** (GADM v4.1) — a *vector* layer (polygons, lines, or points stored as coordinates) that draws the outline of Thailand on the map. We download it from **GADM**, the global database of administrative boundaries. The classic file format for vectors is the **shapefile** (`.shp` plus several side-car files), but we save ours as a **GeoPackage** (`.gpkg`), a single, modern, self-contained file. The `level` argument tells GADM how finely to subdivide the country: `level = 0` returns the whole country as a single polygon, `level = 1` returns provinces, `level = 2` returns districts, and so on. We only need the country outline, which is `level = 0`.

2.  **WorldClim v2.1 bioclim variables** (10 arc-minute, 19 layers, cropped + masked to Thailand) — the global standard climate dataset. *"10 arc-minute"* is the spatial resolution: one pixel is 10 minutes of arc on a side, which is about 18 km at the equator. (Finer resolutions exist: 5, 2.5, 0.5 arc-min but they are much larger files; 10 arc-min is plenty for a workshop demo.) The *"19 layers"* are the **BIO1–BIO19** *bioclimatic variables*: ecologically meaningful summaries of temperature and precipitation (e.g. BIO1 = annual mean temperature, BIO12 = annual precipitation, BIO15 = precipitation seasonality). They are derived from monthly climate normals and are the most commonly used predictors in ecological niche modelling.

3.  **Occurrence records** for Sambar deer (*Rusa unicolor*) in Thailand [**GBIF data**], strictly filtered to records whose `basisOfRecord` field is one of `HUMAN_OBSERVATION`, `OCCURRENCE`, or `OBSERVATION` — i.e. **field observations of live animals**. The `basisOfRecord` field in GBIF tells you *how a record was generated* (was the animal seen alive, photographed, trapped, a museum specimen, a fossil, etc.?). For ecological niche modelling we want records that represent the *contemporary, free-living distribution* of the species, so we exclude fossils, preserved specimens, and other non-field categories.

4.  **Sampling-bias proxies** (a species-richness layer and a night-time-lights layer) used in **Step 1: nicheR**. Sampling bias means that some places get visited and surveyed far more often than others, regardless of whether the species is actually more common there. To *model* that bias, we use two stand-in surfaces (proxies): species richness (the total number of recorded species in each pixel, a rough indicator of survey effort) and night-time lights (a satellite measure of human activity, a rough indicator of accessibility and urbanisation). Together these proxies tell us where field ecologists are most likely to have been.

# Install and load packages

The code below installs everything the Pre-workshop needs the first time you run it. Re-running is safe — already-installed packages are skipped.

```{r install, eval = FALSE}
cran_pkgs <- c(
  "terra",   
  "sf",        
  "geodata",   
  "rgbif",    
  "dplyr",    
  "readr",      
  "rmarkdown",
  "knitr",
  "remotes"
)
to_install <- cran_pkgs[!cran_pkgs %in% rownames(installed.packages())]
if (length(to_install)) install.packages(to_install, dependencies = TRUE)
```

```{r packages}
library(terra) # modern raster + vector spatial backend
library(sf) # tidy vector spatial data ("simple features")
library(geodata) # one-liner downloads for GADM, WorldClim, soils, etc.
library(rgbif) # programmatic access to GBIF from R
library(dplyr) # tidy data manipulation (select, filter, mutate, ...)
library(readr) # fast, friendly CSV reading and writing
```

```{r paths}
# Everything we download lives under `data/`. We split it into two top-level
# folders so the raw downloads stay separate from the clean, analysis-ready
# outputs the workshop notebooks will read.
raw_dir <- "data/raw"                          # untouched downloads
processed_dir <- "data/processed"                    # cleaned, analysis-ready

# `data/processed/` is organised into role-based sub-folders so each workshop
# Step knows where to find its inputs.
boundary_dir <- file.path(processed_dir, "boundary") # Thailand polygon
bioclim_dir <- file.path(processed_dir, "bioclim") # WorldClim layers
bias_dir <- file.path(processed_dir, "bias") # sampling-bias proxies (Step 1)
gbif_dir <- file.path(processed_dir, "gbif") # GBIF Sambar records

# Create every folder if it does not already exist (safe to re-run).
for (d in c(raw_dir, processed_dir,
            boundary_dir, bioclim_dir, bias_dir, gbif_dir)) {
  dir.create(d, recursive = TRUE, showWarnings = FALSE)
}
```

# Thailand boundary (GADM)

We download the Thailand national outline from GADM and save it as a GeoPackage so every later notebook can read it with a single `st_read()` call.

```{r tha-boundary}
# Download Thailand at `level = 0` (= national outline only).
# `geodata::gadm()` caches the raw download under `raw_dir`.
tha <- gadm(country = "THA", level = 0, path = raw_dir)

# Convert the SpatVector to an `sf` object — sf is the modern tidy class for
# vector spatial data and plays nicely with dplyr.
tha_sf <- st_as_sf(tha)

# Save as a GeoPackage (one self-contained file) into the boundary sub-folder.
st_write(tha_sf,
         file.path(boundary_dir, "thailand_boundary.gpkg"),
         delete_dsn = TRUE, quiet = TRUE)

plot(st_geometry(tha_sf), main = "Thailand boundary (GADM level 0)")
```

# Environmental variables

We download all 19 bioclim layers at 10 arc-min resolution, then crop and mask them so we keep only the pixels that fall inside Thailand.

```{r bioclim}
# Download the global WorldClim v2.1 bioclim stack (19 layers).
bio_global <- worldclim_global(var = "bio", res = 10, path = raw_dir)

# Restrict the global stack to Thailand:
#   crop() trims the raster to Thailand's bounding box (the rectangle)
#   mask() sets every pixel *outside* the country polygon to NA
bio_tha <- crop(bio_global, terra::ext(tha))
bio_tha <- mask(bio_tha, tha)

# Tidy the layer names so they read as `bio1` ... `bio19` instead of the
# verbose WorldClim default (`wc2.1_10m_bio_1`, ...). The vignettes refer to
# the variables by these short names.
names(bio_tha) <- paste0("bio", seq_len(terra::nlyr(bio_tha)))

# Save the 19-band raster as a single multi-band GeoTIFF.
writeRaster(
  bio_tha,
  filename = file.path(bioclim_dir, "bioclim_thailand.tif"),
  overwrite = TRUE
)

# Re-read from disk so what we plot is exactly what later notebooks will see.
bio_tha <- rast(file.path(bioclim_dir, "bioclim_thailand.tif"))

plot(bio_tha[[c("bio1", "bio12", "bio15")]],
     main = c("BIO1 — mean T (°C)",
              "BIO12 — annual precip. (mm)",
              "BIO15 — precip. seasonality (CV)"))
```

# Occurrence records for Sambar deer

Now we download every available occurrence record of Sambar deer (*Rusa unicolor*) in Thailand from **GBIF** (the *Global Biodiversity Information Facility*) — the largest open database of species records in the world.

GBIF records carry a `basisOfRecord` tag that tells you *how each record was generated*. For ecological niche modelling we want records that represent the **contemporary, free-living distribution** of the species, so we keep only field observations:

- **`HUMAN_OBSERVATION`** — a person directly observed the animal in the field (e.g. a wildlife survey or a citizen-science sighting). This is the gold standard for ENM.
- **`OBSERVATION`** — a generic observation record, less specific than `HUMAN_OBSERVATION`. Still a field record, so we keep it.
- **`OCCURRENCE`** — a catch-all "the species was here at this time" record. We include it because excluding it would discard otherwise usable field data.

We **exclude** these categories because they do not represent contemporary, wild, in-situ animals:

- `FOSSIL_SPECIMEN` — the animal lived long ago; not relevant to today's climate niche.
- `PRESERVED_SPECIMEN` — museum or herbarium material; locality information is often skewed toward roads, towns, or historical expeditions.
- `LIVING_SPECIMEN` — captive animals in zoos or breeding centres; their location says nothing about the wild niche.
- `MATERIAL_SAMPLE` / `MACHINE_OBSERVATION` — DNA samples or automated detections (e.g. camera traps, acoustic loggers). They are valuable but require dedicated processing pipelines and are excluded here for simplicity.

We will work through the GBIF pipeline in five short sub-steps below.

## Sub-step 1 — Set up your GBIF credentials

GBIF requires a **free account** to download data. If you do not yet have one, sign up at <https://www.gbif.org> — it takes one minute.

Then open the chunk below and replace the three bracketed placeholders (`[YOUR_GBIF_USERNAME]`, `[YOUR_EMAIL]`, `[YOUR_GBIF_PASSWORD]`) with your actual GBIF credentials. The `rgbif` package reads them from these three `options()` whenever you query the API.

> **Privacy tip.** Before you share this notebook (or commit it to a public Git repository), restore the bracketed placeholders so your password does not end up on the internet. The download chunks further down are smart enough to **skip themselves** as long as the placeholders are still in place — they will only run once you have pasted in real credentials.

```{r gbif-creds}
# ---- Fill in your gbif.org credentials -----------------------------------
options(gbif_user = "[YOUR_GBIF_USERNAME]")
options(gbif_email = "[YOUR_EMAIL]")
options(gbif_pwd = "[YOUR_GBIF_PASSWORD]")

# Safety check: do all three values look like real credentials, not the
# bracketed placeholders? If any value still starts with "[", we treat the
# credentials as "not set" and skip the download chunks below.
.have_gbif_creds <- !any(grepl("^\\[", c(
  getOption("gbif_user"),
  getOption("gbif_email"),
  getOption("gbif_pwd")
)))
```

## Sub-step 2 — Look Sambar deer up in the GBIF taxonomy

Scientific names can be ambiguous (synonyms, subspecies, typos). GBIF resolves every name to a numeric **usage key** that uniquely identifies one accepted taxon. We will use that key in the next sub-step so the download is taxonomically unambiguous.

```{r gbif-key, eval = .have_gbif_creds}
# Resolve "Rusa unicolor" to its accepted GBIF backbone usage key (an integer).
key <- name_backbone(name = "Rusa unicolor")$usageKey
key
```

## Sub-step 3 — Download field-observation records

We loop over the three `basisOfRecord` categories we kept, calling `rgbif::occ_search()` once per category and concatenating the results. We also ask GBIF to return only records that already have coordinates and that have no flagged geospatial issues — that saves us a cleaning round later.

```{r gbif-download, eval = .have_gbif_creds}
allowed_basis <- c("HUMAN_OBSERVATION", "OCCURRENCE", "OBSERVATION")

# Query GBIF once per basisOfRecord category. `lapply()` returns a list of
# data frames, one per category, which we then row-bind into a single table.
raw_list <- lapply(allowed_basis, function(b) {
  rgbif::occ_search(
    taxonKey = key, # unique taxon identifier from Sub-step 2
    country = "TH", # ISO-2 country code for Thailand
    basisOfRecord = b, # this category only
    hasCoordinate = TRUE, # drop records without lon/lat
    hasGeospatialIssue = FALSE, # drop records with known coord problems
    limit = 5000 # generous upper bound per category
  )$data
})

raw <- dplyr::bind_rows(raw_list)

# Belt-and-braces: re-apply the basisOfRecord filter on the combined table
# in case the API quietly returns something we did not ask for.
raw <- raw %>% filter(basisOfRecord %in% allowed_basis)

# Save the raw download to disk so we have a snapshot we can re-read later
# without re-hitting GBIF.
write_csv(raw, file.path(gbif_dir, "sambar_gbif.csv"))

nrow(raw)
```

## Sub-step 4 — Clean the table and keep only points inside Thailand

GBIF tags records with the **country** they were observed in, but the polygon we plot from is more precise. Coordinates labelled `country = "TH"` can still fall a few kilometres offshore (e.g. island sightings, GPS error). We filter spatially against our Thailand polygon to be sure every record actually lies on Thai land.

```{r gbif-clean, eval = .have_gbif_creds}
# Re-read the raw snapshot we just wrote.
sambar <- read_csv(file.path(gbif_dir, "sambar_gbif.csv"))

# Keep only the columns we will need downstream and drop rows with missing
# coordinates (we already asked GBIF for `hasCoordinate = TRUE`, but a final
# sanity check is cheap).
sambar <- raw %>%
  select(species,
         decimalLongitude, decimalLatitude,
         year, month,
         basisOfRecord, gbifID) %>%
  filter(!is.na(decimalLongitude),
         !is.na(decimalLatitude),
         basisOfRecord %in% allowed_basis)

# Turn the table into spatial points (CRS 4326 = WGS84 longitude/latitude).
# `remove = FALSE` keeps the numeric lon/lat columns alongside the geometry.
sambar_sf <- st_as_sf(sambar,
                      coords = c("decimalLongitude", "decimalLatitude"),
                      crs = 4326,
                      remove = FALSE)

# Test which records actually fall inside the Thailand polygon. `st_within()`
# returns a one-column logical matrix, so we pull out column 1.
inside  <- st_within(sambar_sf, tha_sf, sparse = FALSE)[, 1]
sambar  <- sambar[inside, ]
```

## Sub-step 5 — Save the analysis-ready dataset and visualise

We finally write the cleaned, in-Thailand-only table as `sambar_thailand.csv`. **Step 1**, **Step 2**, and **Step 3** all read from this single file.

```{r gbif-save, eval = .have_gbif_creds}
write_csv(sambar, file.path(gbif_dir, "sambar_thailand.csv"))

plot(st_geometry(tha_sf), border = "grey20",
     main = "Sambar deer occurrences")
points(sambar$decimalLongitude, sambar$decimalLatitude,
       pch = 21, bg = "tomato", col = "white", cex = 0.7)
```

# Shortcut — download the pre-built data from Zenodo

If you would rather skip the live downloads in the sections above, you can grab a pre-built copy of the entire `data/processed/` path. That is, the *cleaned*, *analysis-ready* outputs that every workshop notebook reads:

```         
data/processed/
├── boundary/   # Thailand polygon (GeoPackage)
├── bioclim/    # WorldClim bioclim stack for Thailand (GeoTIFF, 19 bands)
├── bias/       # Sampling-bias proxy rasters (species richness, night-lights)
├── gbif/       # Cleaned Sambar deer occurrence CSV
└── temporal_split/   # Per-year LST, precip, and NDVI rasters used by Step 3
```

The pre-built archive lives on **Zenodo**, an open research-data repository that mints a permanent DOI for every release:

> **DOI:** <https://zenodo.org/records/20344510>
>
> **File:** `processed.zip` — the full `data/processed/` tree, including `temporal_split/`.

```{r download-external, eval = FALSE}
# ---- 1. URL of the Zenodo archive -----------------------------------------
zenodo_url <- "https://zenodo.org/records/20344510/files/processed.zip?download=1"

# ---- 2. Where to put it ---------------------------------------------------
# The archive unpacks into `data/processed/...` directly, so we point `unzip`
# at `data/` and the folder structure lines up with the rest of the workshop.
target_dir <- "data"
dir.create(target_dir, showWarnings = FALSE, recursive = TRUE)

# ---- 3. Download to a temporary file and unpack ---------------------------
zip_path <- tempfile(fileext = ".zip")
message("Downloading from: ", zenodo_url)
utils::download.file(zenodo_url, destfile = zip_path, mode = "wb")

message("Unpacking into: ", normalizePath(target_dir))
utils::unzip(zip_path, exdir = target_dir, overwrite = TRUE)

# Clean up the temp file.
unlink(zip_path)
```

After running the code above, `data/processed/{boundary, bioclim, bias, gbif, temporal_split, ...}/` will all be populated and you can move straight on to **Step 1: nicheR**.

# References

- Castaneda-Guzman M, Hughes C, Paansri P, Cobos M (2026). *nicheR: Ellipsoid-Based Virtual Niches and Visualization*. R package version 0.1.0, <https://github.com/castanedaM/nicheR>.
- Chamberlain S, Barve V, Mcglinn D, Oldoni D, Desmet P, Geffert L, Ram K (2026). *rgbif: Interface to the Global Biodiversity Information Facility API*. R package version 3.8.5.2, <https://CRAN.R-project.org/package=rgbif>.
- Hijmans R (2026). *geodata: Access Geographic Data*. R package version 0.6-10, <https://rspatial.github.io/geodata/>.
- Hughes C, Castaneda-Guzman M, E. Escobar L (2026). *TemporalModelR: Temporally Explicit Species Distribution Modelling in R*. R package version 0.2.0, <https://github.com/CJHughes926/TemporalModelR>.
- Paansri P, Escobar L (2026). *bean: Data Thinning of Species Occurrences in Environmental Space*. R package version 0.2.1, <https://github.com/paanwaris/bean>.
