---
title: "Step 3 — Temporally-explicit SDMs with **TemporalModelR**"
subtitle: "ENM-Thailand workshop · Sambar deer (*Rusa unicolor*)"
author: "Workshop materials"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: cosmo
    df_print: paged
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  dpi       = 100
)
```

# Overview

> **Workshop navigation**
>
> - [**Pre-workshop** — Downloading the workshop data.](https://paanwaris.github.io/ENM-Thailand/articles/Pre-workshop.html) Get the Thailand boundary, WorldClim climate layers, and GBIF Sambar records before the workshop starts.
> - [**Step 1: nicheR** — Virtual species with nicheR.](https://paanwaris.github.io/ENM-Thailand/articles/Step1_nicheR.html) Build a virtual species niche and map it.
> - [**Step 2: bean** — Environmental thinning with bean.](https://paanwaris.github.io/ENM-Thailand/articles/Step2_bean.html) Reduce environmental sampling bias before fitting a real niche.
> - **Step 3: TemporalModelR — Temporally-explicit models with TemporalModelR.** Link species occurrences with environmental data through time.

`TemporalModelR` builds **temporally-explicit** species distribution models. Each occurrence is paired with the environmental conditions *at the time it was recorded*, rather than against a long-term average. Models can then project predictions as time-series surfaces and detect pixel-level temporal trajectories (always-suitable, never-suitable, increasing, decreasing, or fluctuating).

> **Data.** Step 3 uses the pre-prepared annual rasters that live under `data/processed/temporal_split/` — one precipitation (`prec_YYYY.tif`), one temperature (`temp_YYYY.tif`, derived from land-surface temperature), and one NDVI (`NDVI_YYYY.tif`) file per year. **No new climate data is downloaded here** — we simply read the local files and stack them. The `Pre-workshop.Rmd` notebook produced the Sambar deer GBIF table and the Thailand polygon we use alongside them.

In this Step we will:

1.  Read the per-year precipitation, temperature, and NDVI rasters and align them to a common grid.
2.  Build a temporal-mean E-space, thin the GBIF Sambar deer occurrences **once** with `bean` against it, then rarefy with `spatiotemporal_rarefaction()`, and pair each surviving record with the environment it actually experienced using `temporally_explicit_extraction()`.
3.  Z-score the rasters, summarise their temporal mean, build spatiotemporal CV folds, and create fold-stratified pseudoabsences.
4.  Fit all four `TemporalModelR` algorithms (`build_temporal_glm()`, `build_temporal_gam()`, `build_temporal_rf()`, `build_temporal_hv()`) against the same partition and project predictions across years with `generate_spatiotemporal_predictions()`.
5.  Postprocess **every model** with `summarize_raster_outputs()`, `analyze_temporal_patterns()`, and `analyze_trends_by_spatial_unit()` so consensus surfaces, trajectory classifications, and provincial summaries are directly comparable across algorithms.

All function names and arguments below follow the official package documentation: <https://cjhughes926.github.io/TemporalModelR/>

# Install and load packages

```{r install, eval = FALSE}
# ---- CRAN dependencies -----------------------------------------------------
cran_pkgs <- c(
  "terra",
  "sf",
  "dplyr",
  "readr",
  "viridis",
  "mgcv",          # backend for build_temporal_gam()
  "randomForest",  # backend for build_temporal_rf()
  "hypervolume",   # backend for build_temporal_hv() (presence-only)
  "fastcpd",
  "rmarkdown",
  "knitr",
  "remotes",
  "geodata",
  "bean"
)
to_install <- cran_pkgs[!cran_pkgs %in% rownames(installed.packages())]
if (length(to_install)) install.packages(to_install, dependencies = TRUE)

# ---- TemporalModelR from GitHub --------------------------------------------
if (!requireNamespace("TemporalModelR", quietly = TRUE)) {
  remotes::install_github("CJHughes926/TemporalModelR")
}
```

```{r packages}
library(TemporalModelR)  # temporally-explicit SDM pipeline
library(bean)            # per-year environmental thinning
library(terra)           # modern raster + vector spatial backend
library(sf)              # tidy vector spatial data ("simple features")
library(dplyr)           # tidy data manipulation
library(readr)           # fast, friendly CSV reading and writing

# Make every random draw in the rest of the file reproducible.
set.seed(2026)
```

# Inputs

For Step 3 we **fit and project the four models on the whole of Thailand** so the niche estimate sees the full range of climates Sambar deer actually use in the country. The 16-year modelling window runs from 2010 to 2025. Only the final, *regional* post-processing step is restricted to the **Dong Phayayen–Khao Yai Forest Complex (DPKY)** — a UNESCO World Heritage Site spanning roughly 6,150 km² of contiguous protected forest across five provinces in eastern Thailand and one of the most important refuges for wild Sambar deer in Thailand. Modelling Thailand-wide and reporting at DPKY-level gives us the best of both worlds: the model has enough data to estimate a credible niche, and the conservation-relevant interpretation focuses on the protected complex where management decisions are actually made.

We therefore load **two** polygons in this chunk: the country boundary `thailand` (used as the study-area reference for cross-validation folds and pseudoabsences below), and the DPKY polygon (held in reserve for the *Aggregate by spatial unit* step at the end of the document). The DPKY shapefile lives at `data/processed/DPKY/DPKY.shp` together with its sibling `.dbf`, `.prj`, `.shx`, and `.cpg` files, which together make up a complete ESRI shapefile.

```{r load-base}
# Load the country boundary polygon prepared in the Pre-workshop. This is
# the study-area reference used by spatiotemporal_partition() and
# generate_absences() further down — i.e. the polygon the modelling
# pipeline treats as the universe of possible locations.
# Arguments to st_read():
#   - First positional argument: path to the GeoPackage file.
#   - quiet = TRUE: suppresses the usual GDAL chatter so the rendered
#     HTML stays clean.
thailand <- st_read("data/processed/boundary/thailand_boundary.gpkg",
                    quiet = TRUE)

plot(thailand[2])

# Load the Dong Phayayen-Khao Yai Forest Complex polygon. We do not use
# DPKY for fitting or projecting — only for the final regional
# aggregation step, where we want pixel-level trajectories summarised at
# the level of the protected forest complex rather than at the level of
# every Thai province.
# Arguments to st_read():
#   - First positional argument: path to the .shp file. sf automatically
#     picks up the sibling .dbf / .prj / .shx files in the same folder.
#   - quiet = TRUE: same chatter-suppression as above.
# We pipe the result through st_transform() so the polygon shares the
# WGS84 (EPSG:4326) longitude/latitude CRS used by the GBIF records and
# every raster in the pipeline, guaranteeing no CRS-mismatch error when
# we later overlay it on the trajectory rasters.
dpky <- st_read("data/processed/DPKY/DPKY.shp", quiet = TRUE) |>
  st_transform(4326)

plot(dpky[11])

# Clean GBIF Sambar deer table from the Pre-workshop. We:
#   * restrict to 2010-2025 (the full 16-year range covered by every
#     per-year raster, and the window we will model below);
#   * rename the GBIF coordinate columns to the conventional `longitude` /
#     `latitude` so they match the names bean and TemporalModelR expect;
#   * add a `presence = 1L` column so the table is ready to be combined
#     with pseudoabsences (presence = 0) later on.
sambar <- read_csv("data/processed/gbif/sambar_thailand.csv",
                   show_col_types = FALSE) %>%
  filter(year >= 2010, year <= 2025) %>% # keep the 16-year modelled window
  rename(longitude = decimalLongitude,    # GBIF -> conventional name
         latitude  = decimalLatitude) %>% # GBIF -> conventional name
  mutate(presence = 1L)                   # every GBIF record is a presence

# How many records per year survive the 16-year temporal filter?
table(sambar$year)
```

# Preprocessing

## Align rasters

`raster_align()` reprojects and resamples every raster found in `input_dir` against a reference, then writes the aligned copies into `output_dir`. Because the three variables come from different sources, they do not share an extent, resolution, or CRS out of the box — we use one of the per-year NDVI files as the reference grid for everything.

```{r raster-align}
# Quick visual sanity-check of the three "current" layers.
prec_ex <- rast("data/processed/temporal_split/prec_2025.tif")
temp_ex <- rast("data/processed/temporal_split/temp_2025.tif")
NDVI_ex <- rast("data/processed/temporal_split/NDVI_2025.tif")

par(mfrow = c(1, 3))
plot(prec_ex, main = "Precipitation 2025")
plot(temp_ex, main = "Temperature 2025")
plot(NDVI_ex, main = "NDVI 2025")
par(mfrow = c(1, 1)) # reset the plot region back to a single panel

# Source directory of the per-variable per-year split rasters, and target
# directory where the aligned copies will be written.
split_dir <- "data/processed/temporal_split"
aligned_dir <- "data/processed/rasters_aligned"
dir.create(aligned_dir, showWarnings = FALSE, recursive = TRUE)

# Use NDVI as the reference grid — every per-year raster will be reprojected
# / resampled to match its extent, resolution, and CRS.
ref_raster <- NDVI_ex

raster_align(
  input_dir = split_dir, # where the per-year rasters currently live
  output_dir = aligned_dir, # where the aligned copies will be written
  reference_raster = ref_raster, # target grid (extent + resolution + CRS)
  overwrite = TRUE # replace any aligned files left from a prior run
)
```

## Build a temporal-mean E-space for thinning

Before any thinning, we build a single **temporal-mean** reference E-space that summarises the environment Sambar deer experienced across the entire 16-year **2010–2025** window. We compute the per-pixel mean across years for each of the three aligned, DPKY-masked variables, then stack the three mean surfaces into one 3-layer `SpatRaster`.

This temporal-mean stack will be the **reference environmental space** that `bean` uses to thin every year. Using a single, period-wide E-space (rather than the year-specific raster of each occurrence) keeps the thinning grid *consistent across years*: a 2014 record and a 2020 record that fall in the same temporal-mean pod are treated as environmentally redundant, even though their individual years' rasters may differ slightly.

```{r temporal-mean-env}
# List every aligned per-year raster, then split by variable name.
aligned_files <- list.files(aligned_dir, pattern = "\\.tif$", full.names = TRUE)

# Group the file list by variable name (prec_ / temp_ / NDVI_) so each
# group becomes a multi-layer SpatRaster spanning every year.
prec_files <- file.path(aligned_dir,
                        grep("prec_", basename(aligned_files), value = TRUE))
temp_files <- file.path(aligned_dir,
                        grep("temp_", basename(aligned_files), value = TRUE))
NDVI_files <- file.path(aligned_dir,
                        grep("NDVI_", basename(aligned_files), value = TRUE))

# Per-pixel mean across years for each variable — terra::mean() over a
# multi-layer SpatRaster collapses the time dimension into a single layer.
mean_prec <- terra::mean(rast(prec_files), na.rm = TRUE)
mean_temp <- terra::mean(rast(temp_files), na.rm = TRUE)
mean_NDVI <- terra::mean(rast(NDVI_files), na.rm = TRUE)

# Stack the three mean surfaces into one 3-layer reference E-space, and
# rename the layers to the variable names bean will look up.
env_ref <- c(mean_prec, mean_temp, mean_NDVI)
names(env_ref) <- c("mean_prec", "mean_temp", "mean_NDVI")

par(mfrow = c(1, 3))
plot(mean_prec, main = "Mean precipitation (2010–2025)")
plot(mean_temp, main = "Mean temperature (2010–2025)")
plot(mean_NDVI, main = "Mean NDVI (2010–2025)")
par(mfrow = c(1, 1)) # reset the plot region back to a single panel
```

Reading the panels: each map is the per-pixel average of the corresponding variable across the 16 modelled years, in the variable's natural units (mm for precipitation, °C-equivalent units for land-surface-temperature-derived `temp`, and a unitless 0–1 index for NDVI). Bright patches stand for places that were *consistently* wet, warm, or vegetated; dark patches stand for places that were *consistently* dry, cool, or sparse. This is the environmental landscape `bean` will use to define the thinning grid.

## Environmental thinning with `bean`

We reduce **environmental sampling bias** in the Sambar deer occurrences using `bean`, with the temporal-mean stack `env_ref` from the previous section as the reference E-space.

Because Step 3 treats every year as an independent observation along the time axis, we run `bean` **once per year** rather than once over the pooled occurrence table. Each year is thinned against only its own subset of records, so a heavily-sampled year cannot drag the thinning grid away from sparsely-sampled ones. The reference E-space `env_ref` stays *constant* across iterations: it is the period-wide environmental landscape every record is judged against, so the same env value means the same thing in 2018 as in 2024. Two records from *different* years that happen to fall in the same `env_ref` pod are both kept, because they carry independent information about the niche at two different points in time; only records that cluster *within the same year* are treated as environmentally redundant and thinned away.

We use the **`"sheather-jones"`** bandwidth selector inside `find_env_resolution()`. Sheather–Jones is a data-adaptive, plug-in selector (Sheather & Jones, 1991) that is robust to the non-Gaussian shapes typical of real occurrence data — it is the modern recommended default and is what the workshop uses everywhere `bean` appears. We pair it with the stochastic `thin_env_nd()` so each pod retains one of the original GBIF coordinates rather than a synthetic centroid, preserving the link back to the source record.

The loop below follows the same five steps inside every iteration:

1. **Pull the year's rows** out of `sambar`.
2. **Slim them to (longitude, latitude)** for bean — bean's pipeline only needs the coordinate pair.
3. **Run the bean pipeline** — `prepare_bean()` extracts the env values at each occurrence and z-scores them, `find_env_resolution()` picks per-variable Sheather–Jones bandwidths as the grid cell widths, and `thin_env_nd()` randomly retains one occurrence per occupied pod.
4. **Recover every original column** of the year's subset (`year`, `presence`, `gbifID`, `basisOfRecord`, …) by matching the thinned (longitude, latitude) pair back onto the year's rows.
5. **Stash the result** in a pre-allocated, year-keyed list slot so the per-year tables can be row-bound into a single data frame at the end.

`find_env_resolution()` needs at least four complete cases to estimate a bandwidth, so any year with fewer than five records is passed through unchanged by the safety guard — there is nothing meaningful to thin away in a year with only a handful of records. We pre-allocate the list of results outside the loop (`vector("list", length(years))`) rather than growing it one element at a time, which is the idiomatic R pattern for inside-a-loop accumulation. At the end, `do.call(rbind, thinned_list)` row-binds every year's table into one data frame whose columns are identical to the original `sambar`.


```{r thin-once}
# Build the list of years to iterate over, in chronological order.
years <- sort(unique(sambar$year))

# Pre-allocate a year-named list to hold each year's thinned table.
thinned_list <- vector("list", length(years))
names(thinned_list) <- as.character(years)

for (yr in years) {

  # ---- 1. Pull this year's rows out of the full sambar table -------------
  yr_df <- sambar[sambar$year == yr, ]

  # ---- 2. Coordinate-only slice for bean ---------------------------------
  yr_xy <- yr_df[, c("longitude", "latitude")]

  # ---- Safety guard: skip thinning when there is nothing to thin ---------
  # find_env_resolution() needs at least four complete cases; for smaller
  # years we keep the original rows unchanged and move on.
  if (nrow(yr_xy) < 5) {
    thinned_list[[as.character(yr)]] <- yr_df
    next
  }

  # ---- 3a. prepare_bean(): extract env values + z-score them -------------
  prepped <- prepare_bean(
    data        = yr_xy,                        # this year's coord-only table
    env_rasters = env_ref,                      # the constant period-wide E-space
    longitude   = "longitude",                  # name of the x coordinate column
    latitude    = "latitude",                   # name of the y coordinate column
    transform   = "scale"                       # z-score every env variable
  )

  # ---- 3b. find_env_resolution(): pick a per-variable cell width ---------
  res <- find_env_resolution(
    data     = prepped,                                          # the prepped, scaled table
    env_vars = c("mean_prec", "mean_temp", "mean_NDVI"),         # columns that define the grid axes
    method   = "sheather-jones"                                  # Sheather-Jones plug-in bandwidth
  )

  # ---- 3c. thin_env_nd(): keep one occurrence per occupied pod -----------
  thin_out <- thin_env_nd(
    data            = prepped,                                   # same prepped table
    env_vars        = c("mean_prec", "mean_temp", "mean_NDVI"),  # same three axes
    grid_resolution = res$suggested_resolution                   # per-variable cell widths from 3b
  )

  # ---- 4. Recover the original columns by matching on (lon, lat) ---------
  # paste() joins each row's two coordinates into a single string so we can
  # check membership in one step with %in%.
  keep_key <- paste(thin_out$thinned_data$longitude,
                    thin_out$thinned_data$latitude)
  yr_thinned <- yr_df[paste(yr_df$longitude, yr_df$latitude) %in% keep_key, ]

  # ---- 5. Stash this year's thinned table into the year-keyed slot ------
  thinned_list[[as.character(yr)]] <- yr_thinned
}

# Combine the per-year tables into one data frame, then clear the rbind
# row names (they are meaningless "year.position" composites).
sambar_thinned <- do.call(rbind, thinned_list)
rownames(sambar_thinned) <- NULL

# Before / after counts per year — verifies the year column survived.
# factor(..., levels = ...) on `after` forces table() to include a zero
# for any year where every record was thinned away.
data.frame(
  year   = sort(unique(sambar$year)),
  before = as.integer(table(sambar$year)),
  after  = as.integer(table(factor(sambar_thinned$year,
                                   levels = sort(unique(sambar$year)))))
)

# Use the thinned table for everything that follows in Step 3.
sambar <- sambar_thinned
```

The before/after table reports, for each modelled year, how many records went into the thinning step and how many survived its own pod-by-pod sampling. Years where `after` is much smaller than `before` are the years that carried the most environmental redundancy; years where the two columns are equal are either too small to thin (the safety guard skipped them) or already environmentally well-spread within that single year.

## Spatio-temporal rarefaction

Once the per-year thinning has reduced *environmental* duplication, we still want to reduce *spatial* duplication: keep at most one record per raster pixel **per year**, so a repeated visit to the same place in a later year is preserved as a temporally-independent observation.

```{r rarefy}
# Working directory for everything point-related downstream.
points_dir <- "data/processed/points"
dir.create(points_dir, showWarnings = FALSE, recursive = TRUE)

# Persist the thinned table to disk so spatiotemporal_rarefaction() can
# pick it up by file path.
points_csv <- file.path(points_dir, "sambar_temporal.csv")
write_csv(sambar, points_csv)

# Use the CRS of the reference raster so points and rasters live in the
# same coordinate system throughout the pipeline.
study_crs <- st_crs(rast(ref_raster))

rare_out <- spatiotemporal_rarefaction(
  points_sp = points_csv, # input CSV of presences
  output_dir = file.path(points_dir, "rarefied"), # where rarefied tables are written
  reference_raster = ref_raster, # defines the pixel grid for rarefaction
  time_cols = "year", # column carrying the time stamp
  xcol = "longitude", # column with x coordinate
  ycol = "latitude", # column with y coordinate
  points_crs = study_crs, # CRS of the input points
  output_prefix = "Sambar_annual", # prefix for the written files
  verbose = FALSE # silence per-record progress messages
)

# The three counts below show how aggressive rarefaction was: how many
# records went in, how many survive pure spatial rarefaction, and how
# many survive spatio-temporal rarefaction (one record per pixel per
# year).
rare_out$input_points # n records that entered rarefaction
rare_out$spatial_points # n surviving pure spatial rarefaction (one per pixel)
rare_out$spatiotemporal_points # n surviving spatio-temporal (one per pixel per year)
```

## Match each observation to the environment it experienced

`temporally_explicit_extraction()` walks the aligned raster directory using `variable_patterns`: clean variable names on the left, filename patterns with time placeholders (`YEAR`) on the right. For each surviving occurrence, it pulls the precipitation, temperature, and NDVI values from the raster of the matching year — so a 2014 record is paired with the 2014 climate, not the long-term average.

```{r extract}
# Output directory for the extracted tables and the scaling parameters.
ext_dir <- file.path(points_dir, "extracted")

ext_out <- temporally_explicit_extraction(
  points_sp = rare_out$files_created$spatiotemporal, # rarefied points to extract at
  raster_dir = aligned_dir, # raster source folder
  variable_patterns = c( # variable -> filename pattern map
    "prec" = "prec_YEAR",
    "temp" = "temp_YEAR",
    "NDVI" = "NDVI_YEAR"
  ),
  time_cols = "year", # time-stamp column on the points
  xcol = "X", # rarefied table uses upper-case X / Y
  ycol = "Y",
  points_crs = study_crs, # CRS shared by points and rasters
  output_dir = ext_dir,
  output_prefix = "sambar_extracted",
  save_raw = TRUE, # write the raw (un-scaled) extracted table
  save_scaled = TRUE, # also write a z-scored copy
  save_scaling_params = TRUE, # write the per-variable means + SDs we'll reuse
  verbose = FALSE # silence per-record extraction progress
)

# Inspect the first few rows of the raw (un-scaled) extracted table.
head(ext_out$raw_values)

# The means and SDs used to z-score the points. We will re-use these
# below so the rasters and points share the same transformation.
ext_out$scaling_params
```

## Scale rasters

`scale_rasters()` z-scores every raster in `input_dir` using the per-variable means and SDs from extraction, so that the rasters and the occurrence points share the *same* transformation. This is critical: a GLM or RF coefficient learned on z-scored points only makes sense when applied to z-scored rasters.

```{r scale}
# Destination directory for the scaled (z-scored) per-year rasters.
scaled_dir <- "data/processed/rasters_scaled"
dir.create(scaled_dir, showWarnings = FALSE, recursive = TRUE)

scale_rasters(
  input_dir = aligned_dir, # raw aligned rasters
  output_dir = scaled_dir, # write the z-scored copies here
  scaling_params_file = ext_out$files_created$scaling_params, # reuse the means + SDs from extraction
  variable_patterns = c(
    "prec" = "prec_YEAR",
    "temp" = "temp_YEAR",
    "NDVI" = "NDVI_YEAR"
  ),
  time_cols = "year",
  overwrite = TRUE, # replace any scaled files left from a previous run
  verbose = FALSE # silence per-raster scaling progress
)
```

## Spatiotemporal partitioning

`spatiotemporal_partition()` splits the occurrence + raster history into cross-validation folds in **both** space and time, so a fold left out for testing is novel along both dimensions — a much harder, more honest evaluation than random k-fold CV.

```{r partition}
par(mfrow = c(1, 3))

partition <- spatiotemporal_partition(
  reference_shapefile_path = thailand, # study-area polygon (Thailand-wide)
  points_file_path = ext_out$files_created$scaled, # presence table with extracted env
  xcol = "x", # scaled table uses lower-case x / y
  ycol = "y",
  points_crs = study_crs,
  time_cols = "year",
  n_spatial_folds = 2, # how many blocks to split the country into
  n_temporal_folds = 2, # how many blocks to split the years into
  max_imbalance = 0.15, # max allowed presence imbalance between folds
  create_plot = TRUE, # produce the fold-map diagnostic
  verbose = FALSE # silence per-fold construction messages
)

# Summary table: which fold has which years and which spatial block.
partition$summary

par(mfrow = c(1, 1)) # reset the plot region back to a single panel
```

## Generate pseudoabsences

A binary GLM or RF needs *absences* as well as presences. We generate **fold-stratified pseudoabsences** — random non-presence locations drawn at a 2:1 absence:presence ratio from anywhere outside a 50 km buffer around the presences in the same fold.

```{r absences}
absences <- generate_absences(
  partition_result = partition, # fold structure built above
  reference_shapefile_path = thailand, # absences are drawn inside the Thailand polygon
  raster_dir = scaled_dir, # env values are extracted at each absence
  variable_patterns = c(
    "prec" = "prec_YEAR",
    "temp" = "temp_YEAR",
    "NDVI" = "NDVI_YEAR"
  ),
  method = "buffer", # draw absences outside a buffer around presences
  buffer_distance = 50000, # 50 km buffer around each presence
  ratio = 2, # 2 absences per presence
  time_cols = "year",
  create_plot = TRUE, # produce a diagnostic absence map
  plot_by_fold = TRUE, # one panel per fold
  verbose = FALSE # silence per-fold progress messages
)

# Per-fold counts of presences vs absences.
absences$summary
```

# Modelling

`TemporalModelR` ships four model families, all with the same `partition_result` + (optionally) `pseudoabsence_result` interface so swapping algorithms is a one-chunk change:

- **GLM** — generalised linear model (`build_temporal_glm`). Linear / polynomial responses, easiest to interpret.
- **GAM** — generalised additive model (`build_temporal_gam`, via `mgcv`). Smooth, data-adaptive nonlinear responses while still per-variable interpretable.
- **Random forest** — ensemble of decision trees (`build_temporal_rf`, via `randomForest`). Captures non-linear interactions automatically; less interpretable than GLM/GAM.
- **Hypervolume** — *n*-dimensional kernel-density envelope (`build_temporal_hv`, via `hypervolume`). **Presence-only** — no pseudoabsences needed and no probability threshold.

We will fit all four against the same partition so the per-fold metrics are directly comparable. Three of them (GLM, GAM, RF) consume the same presence/absence table; only the hypervolume omits `pseudoabsence_result`.

## GLM (linear + a quadratic on temperature)

The temperature niche is expected to peak at some intermediate value and fall away on either side, so we add a quadratic term on `temp` to capture that hump. The `tss` threshold method picks the cutoff that maximises the True Skill Statistic on each fold's test set.

```{r glm}
glm_res <- build_temporal_glm(
  partition_result = partition, # CV folds in space + time
  pseudoabsence_result = absences, # presences + absences per fold
  model_formula = ~ temp + I(temp^2) + prec + NDVI, # quadratic on temp, linear on prec / NDVI
  link = "logit", # binary GLM with logit link
  threshold_method = "tss", # pick threshold maximising True Skill Statistic
  output_dir = "outputs/GLM_Models",
  create_plot = FALSE,
  overwrite = TRUE,
  time_cols = "year",
  verbose = FALSE
)

# Out-of-fold E-space test metrics — one row per fold.
glm_res$fold_test_metrics
```

## GAM (smooth nonlinear responses)

A GAM lets each predictor have a *smooth, data-adaptive* response curve without hand-coding polynomials. We wrap every predictor in `s()` (thin-plate regression spline) and let `mgcv` choose the basis dimension. The link function, threshold rule, and fold structure are the same as the GLM, so the two are directly comparable.

```{r gam}
gam_res <- build_temporal_gam(
  partition_result = partition, # CV folds in space + time
  pseudoabsence_result = absences, # presences + absences per fold
  model_formula = ~ s(temp) + s(prec) + s(NDVI), # `s()` = univariate smooth (mgcv thin-plate spline)
  link = "logit", # binary GAM with logit link
  gam_params = list(method = "REML"), # smoothing-parameter selection rule (mgcv default REML)
  threshold_method = "tss", # pick threshold maximising True Skill Statistic
  output_dir = "outputs/GAM_Models",
  create_plot = FALSE, # skip the smooth-response curves to keep output compact
  overwrite = TRUE,
  time_cols = "year",
  verbose = FALSE # silence per-fold smoothing progress
)

# Out-of-fold E-space test metrics — one row per fold.
gam_res$fold_test_metrics
```

## Random forest (flexible ensemble)

A 500-tree random forest. RFs capture non-linear interactions automatically and usually outperform the GLM/GAM at the cost of a less interpretable model.

```{r rf}
rf_res <- build_temporal_rf(
  partition_result = partition, # CV folds in space + time
  pseudoabsence_result = absences, # presences + absences per fold
  model_vars = c("temp", "prec", "NDVI"), # RF needs predictor names, not a formula
  rf_params = list(ntree = 500), # number of decision trees grown per fold
  threshold_method = "tss", # pick threshold maximising True Skill Statistic
  output_dir = "outputs/RF_Models",
  create_plot = FALSE, # skip variable-importance plot
  overwrite = TRUE,
  time_cols = "year",
  verbose = FALSE # silence per-fold tree-growth progress
)

# Out-of-fold E-space test metrics — one row per fold.
rf_res$fold_test_metrics
```

## Hypervolume (presence-only kernel density)

The hypervolume builds an *n*-dimensional kernel-density envelope around the presence points alone — it does not need pseudoabsences and does not produce a probability scale, so there is no `pseudoabsence_result` or `threshold_method` argument. Only sensitivity-based metrics are reported (specificity / AUC / TSS require negative training data).

> **Heads-up.** Hypervolume fitting can be slow (several minutes per fold on real data) because it samples millions of random points per envelope. If you are running the workshop on a laptop, you can comment this chunk out and skip the HV columns in the plots below.

```{r hv}
hv_res <- build_temporal_hv(
  partition_result = partition, # CV folds in space + time (presences only)
  model_vars = c("temp", "prec", "NDVI"), # predictors defining the envelope axes
  method = "gaussian", # Gaussian KDE envelope (alternative: "svm" = one-class SVM)
  hypervolume_params = list(), # accept hypervolume::hypervolume_gaussian() defaults
  output_dir = "outputs/HV_Models",
  create_plot = FALSE, # skip the per-fold pairplots
  verbose = FALSE # silence per-fold KDE-construction progress
)

# Presence-only E-space metrics: sensitivity, envelope volume, per-fold overlap.
hv_res$fold_test_metrics
hv_res$volumes
```

# Project predictions through time

`generate_spatiotemporal_predictions()` walks every requested time step, loads the matching rasters, applies the fitted models, and writes **one fold-vote consensus raster per time step**. Every cell of the output carries a *count* (`0`, `1`, …, *N*) of how many of the *N* folds voted that the pixel was suitable in that time step — these are the package's native fold-vote outputs, which we will visualise with the package's default `terra::plot()` styling.

The call signature is identical for all four model types; `model_type` inside `model_result` tells the function which projection logic to apply. The hypervolume projection simply omits `pseudoabsence_result` because the model was never fit against absences.

Key arguments shared by every call below:

- **`raster_dir`** — the folder of *scaled* per-year predictor rasters written by `scale_rasters()`. Predictions must use the same transformation as the training points.
- **`variable_patterns`** — a named vector mapping each predictor name (left) to its filename pattern (right) with the `YEAR` placeholder, so `temp_YEAR` resolves to `temp_2010.tif` for year 2010 (the first year in our 2010–2025 window).
- **`time_steps`** — a `data.frame` giving every (year, …) combination to project. We use 2010–2025 (16 years) to match the `sambar` filter and the 4 × 4 plot grid below.
- **`output_dir`** — where the per-fold per-time-step prediction rasters get written. Each model writes to its own folder so the post-processing later can address them separately.

```{r predict}
time_steps <- data.frame(year = 2010:2025) # 16 years -> fits a 4 x 4 plot grid

# A single named vector used by every projection call.
var_patterns <- c("prec" = "prec_YEAR",
                  "temp" = "temp_YEAR",
                  "NDVI" = "NDVI_YEAR")

# GLM projection: walk through every (year) time step and apply each fold's
# fitted GLM to the matching scaled rasters, writing one fold-vote raster
# per fold per year into output_dir.
glm_preds <- generate_spatiotemporal_predictions(
  partition_result = partition, # CV fold geometry built earlier
  model_result = glm_res, # the fitted GLM (one model per fold)
  pseudoabsence_result = absences, # tells the function which area each fold "owns"
  raster_dir = scaled_dir, # folder of z-scored per-year predictor rasters
  variable_patterns = var_patterns, # maps predictor name -> filename pattern (with YEAR)
  time_cols = "year", # name of the time column on the points
  time_steps = time_steps, # the data.frame of years (2010:2025) to project
  output_dir = "outputs/glm_predictions", # per-fold per-year rasters land here
  overwrite = TRUE, # replace any rasters left from a prior run
  verbose = FALSE # silence the per-fold progress messages
)

# GAM projection: same inputs as the GLM call above, only `model_result`
# and `output_dir` change. The function reads `model_type` inside `gam_res`
# and automatically uses the correct projection logic for a GAM.
gam_preds <- generate_spatiotemporal_predictions(
  partition_result = partition, # same fold geometry
  model_result = gam_res, # the fitted GAM (one per fold)
  pseudoabsence_result = absences, # same fold-stratified absences
  raster_dir = scaled_dir, # same scaled per-year predictor rasters
  variable_patterns = var_patterns, # same name -> filename map
  time_cols = "year", # same time-column name
  time_steps = time_steps, # same 11-year grid
  output_dir = "outputs/gam_predictions", # write into a *separate* folder per model
  overwrite = TRUE, # replace previous-run rasters
  verbose = FALSE # silence per-fold progress
)

# Random forest projection: identical interface; the function dispatches to
# the RF projection logic via `model_type` stored inside `rf_res`.
rf_preds <- generate_spatiotemporal_predictions(
  partition_result = partition, # same fold geometry
  model_result = rf_res, # the fitted RF (500 trees per fold)
  pseudoabsence_result = absences, # same fold-stratified absences
  raster_dir = scaled_dir, # same scaled per-year predictor rasters
  variable_patterns = var_patterns, # same name -> filename map
  time_cols = "year", # same time-column name
  time_steps = time_steps, # same 11-year grid
  output_dir = "outputs/rf_predictions", # separate output folder for RF
  overwrite = TRUE, # replace previous-run rasters
  verbose = FALSE # silence per-fold progress
)

# Hypervolume projection: NOTE the missing `pseudoabsence_result` argument.
# Hypervolume is presence-only, so the function never needs absences when
# projecting an HV model — passing them in is unnecessary.
hv_preds <- generate_spatiotemporal_predictions(
  partition_result = partition, # same fold geometry
  model_result = hv_res, # the fitted hypervolume (Gaussian KDE per fold)
  raster_dir = scaled_dir, # same scaled per-year predictor rasters
  variable_patterns = var_patterns, # same name -> filename map
  time_cols = "year", # same time-column name
  time_steps = time_steps, # same 11-year grid
  output_dir = "outputs/hv_predictions", # separate output folder for HV
  overwrite = TRUE, # replace previous-run rasters
  verbose = FALSE # silence per-fold progress
)
```

## Visualise the consensus maps

For each model we stack the per-time-step prediction rasters straight from `preds$prediction_files` and pass them to `terra::plot()` — the package's default visualisation. Every raster cell is a **fold-vote count** (0 to *N*) of how many folds agreed the pixel was suitable in that year. With 16 modelled years (2010–2025), each stack lays out as a 4 × 4 grid; the spatial extent is Thailand-wide, so every panel shows the country outline filled with per-pixel vote counts.

```{r plot-glm, fig.width = 9, fig.height = 9}
# Load all per-year GLM prediction rasters into a single multi-layer
# SpatRaster. `terra::rast()` accepts a character vector of file paths
# and stacks them in the order given.
glm_pred_stack <- terra::rast(glm_preds$prediction_files)

# Set human-readable layer names (e.g. "Prediction_..._2010") by stripping
# the `.tif` extension from each filename. These names become the panel
# titles in the plot below.
names(glm_pred_stack) <- sub("\\.tif$", "", basename(glm_preds$prediction_files))

# Draw the 16 yearly fold-vote maps as a 4 x 4 grid.
# Arguments:
#   - First positional argument: the multi-layer SpatRaster to plot.
#   - nr = 4, nc = 4: 4 rows by 4 columns of panels (one per modelled year).
#   - mar = c(1.0, 1.0, 1.5, 3.0): inner margins for each panel
#     (bottom, left, top, right) in lines of text; the wider right margin
#     reserves room for the per-panel legend strip.
#   - legend = FALSE: suppress per-panel legends so the grid stays tidy
#     (we explain the colour scale in the prose below).
terra::plot(glm_pred_stack, nr = 4, nc = 4,
            mar    = c(1.0, 1.0, 1.5, 3.0),
            legend = FALSE)
```

```{r plot-gam, fig.width = 9, fig.height = 9}
# Same three-step pattern as the GLM chunk above: build a multi-layer
# SpatRaster from the GAM prediction files, label each layer, and draw a
# 4 x 4 grid of yearly fold-vote maps.
gam_pred_stack <- terra::rast(gam_preds$prediction_files)
names(gam_pred_stack) <- sub("\\.tif$", "", basename(gam_preds$prediction_files))
terra::plot(gam_pred_stack, nr = 4, nc = 4,
            mar    = c(1.0, 1.0, 1.5, 3.0),
            legend = FALSE)
```

```{r plot-rf, fig.width = 9, fig.height = 9}
# Same pattern, but for the random forest prediction files.
rf_pred_stack <- terra::rast(rf_preds$prediction_files)
names(rf_pred_stack) <- sub("\\.tif$", "", basename(rf_preds$prediction_files))
terra::plot(rf_pred_stack, nr = 4, nc = 4,
            mar    = c(1.0, 1.0, 1.5, 3.0),
            legend = FALSE)
```

```{r plot-hv, fig.width = 9, fig.height = 9}
# Same pattern, but for the hypervolume prediction files.
hv_pred_stack <- terra::rast(hv_preds$prediction_files)
names(hv_pred_stack) <- sub("\\.tif$", "", basename(hv_preds$prediction_files))
terra::plot(hv_pred_stack, nr = 4, nc = 4,
            mar    = c(1.0, 1.0, 1.5, 3.0),
            legend = FALSE)
```

Reading these grids: each panel is one year (2010 in the top-left, 2025 in the bottom-right of each 4 × 4 grid). Brighter pixels mean more folds agreed the pixel was suitable in that year; darker pixels mean fewer or no folds did. Any *shift* in the bright patches across panels is the model's view of how Sambar deer habitat changed across Thailand during 2010–2025. Comparing the four model grids side-by-side highlights where the algorithms agree on the niche and where they disagree.

## Per-model performance diagnostics

`plot_model_assessment()` is the package's default diagnostic plot. It produces per-fold time-series of percent suitable, sensitivity (and, for presence/absence models, specificity), and the cumulative-binomial-probability (CBP) test, overlaid with the overall E-space metrics from `$fold_test_metrics` when `model_result` is supplied. We run it once per model below so the four algorithms are directly comparable.

Key arguments:

- **`predictions`** — the object returned by `generate_spatiotemporal_predictions()` for that model.
- **`time_column`** — the name(s) of the time column(s) in the partition. Single annual axis here, so `"year"`.
- **`model_result`** — the fitted-model object; supplying it overlays the E-space `$fold_test_metrics` as reference lines on the per-time-step series.

```{r assess-glm}
# Draw the GLM's per-fold diagnostic time series.
# - predictions:  the object returned by generate_spatiotemporal_predictions()
#                 for this model; contains the per-time-step G-space metrics.
# - time_column:  name of the time axis used for the x-axis of the panels.
# - model_result: the fitted-model object; supplying it overlays the
#                 overall E-space fold metrics as horizontal reference lines.
plot_model_assessment(
  predictions = glm_preds,
  time_column = "year",
  model_result = glm_res
)
```

```{r assess-gam}
# Same diagnostic for the GAM. Arguments behave identically to the GLM
# call above, only the `predictions` and `model_result` objects differ.
plot_model_assessment(
  predictions = gam_preds,
  time_column = "year",
  model_result = gam_res
)
```

```{r assess-rf}
# Same diagnostic for the random forest.
plot_model_assessment(
  predictions = rf_preds,
  time_column = "year",
  model_result = rf_res
)
```

```{r assess-hv}
# Same diagnostic for the hypervolume. The function detects that the
# model is presence-only and automatically drops the specificity panel
# (no negative training data => no specificity).
plot_model_assessment(
  predictions = hv_preds,
  time_column = "year",
  model_result = hv_res
)
```

The hypervolume diagnostics show only the sensitivity, percent-suitable, and CBP panels — specificity is undefined for a presence-only model.

# Postprocessing

Every model produced its own folder of per-fold per-year prediction rasters during projection (`outputs/glm_predictions/`, `outputs/gam_predictions/`, `outputs/rf_predictions/`, `outputs/hv_predictions/`). The three post-processing functions below — `summarize_raster_outputs()`, `analyze_temporal_patterns()`, and `analyze_trends_by_spatial_unit()` — are model-agnostic: they read a prediction folder and operate on whatever the prediction rasters contain. We therefore run **one call per model** in each step so the consensus surfaces, trajectory classifications, and provincial summaries are directly comparable across GLM / GAM / RF / Hypervolume.

## Consensus surfaces across folds

`summarize_raster_outputs()` reads the per-time-step prediction rasters from `predictions_dir` and applies a **consensus rule** across folds to flag pixels that *at least `consensus` folds* agreed were suitable. Setting `consensus = 2` is a majority rule for a 4-fold partition. It returns a binary `consensus_stack` (one layer per year) plus a `frequency_raster` (the proportion of years each pixel was suitable).

Key arguments:

- **`predictions_dir`** — folder of fold/year prediction rasters from `generate_spatiotemporal_predictions()`.
- **`output_dir`** — where to write the binary stack and frequency raster.
- **`consensus`** — minimum number of folds that must agree before a pixel is called suitable; here, `2` of `4`.

```{r consensus-glm}
# Collapse the GLM's per-fold per-year prediction rasters into two
# products: a binary `consensus_stack` (one layer per year) and a single
# 0-1 `frequency_raster` (proportion of years suitable).
glm_consensus <- summarize_raster_outputs(
  predictions_dir = "outputs/glm_predictions", # source folder of per-fold rasters
  output_dir = "outputs/glm_consensus", # where to write the consensus/frequency outputs
  consensus = 2, # majority rule: >=2 of the 4 folds must agree
  overwrite = TRUE, # replace any outputs from a previous run
  verbose = FALSE # silence progress messages
)
```

```{r consensus-gam}
# Same consensus rule for the GAM. Only the source / target folders change.
gam_consensus <- summarize_raster_outputs(
  predictions_dir = "outputs/gam_predictions", # source folder of per-fold rasters
  output_dir = "outputs/gam_consensus", # GAM-specific output folder
  consensus = 2, # same majority rule
  overwrite = TRUE,
  verbose = FALSE
)
```

```{r consensus-rf}
# Same consensus rule for the random forest.
rf_consensus <- summarize_raster_outputs(
  predictions_dir = "outputs/rf_predictions", # source folder of per-fold rasters
  output_dir = "outputs/rf_consensus", # RF-specific output folder
  consensus = 2, # same majority rule
  overwrite = TRUE,
  verbose = FALSE
)
```

```{r consensus-hv}
# Same consensus rule for the hypervolume.
hv_consensus <- summarize_raster_outputs(
  predictions_dir = "outputs/hv_predictions", # source folder of per-fold rasters
  output_dir = "outputs/hv_consensus", # HV-specific output folder
  consensus = 2, # same majority rule
  overwrite = TRUE,
  verbose = FALSE
)
```

### Per-year binary suitability — the `consensus_stack`

Each `consensus_stack` is a multi-layer `SpatRaster` carrying **one binary layer per year** (yellow = suitable, purple = unsuitable). Plot it the same way we plotted the prediction stacks earlier — a 4 × 4 grid of the 16 modelled years for the model in question.

```{r consensus-stack-glm, fig.width = 9, fig.height = 9}
# Draw the GLM's per-year binary suitability surfaces as a 4 x 4 grid.
# Arguments mirror the prediction-stack plot above:
#   - First positional argument: the binary multi-layer SpatRaster.
#   - nr / nc: 3 rows by 4 columns of yearly panels.
#   - mar:     inner margins per panel (bottom, left, top, right).
#   - legend = FALSE: drop per-panel legends because the values are binary
#     (0 = unsuitable, 1 = suitable) and the prose below covers the
#     interpretation.
terra::plot(glm_consensus$consensus_stack, nr = 4, nc = 4,
            mar = c(1.0, 1.0, 1.5, 3.0),
            legend = FALSE)
```

```{r consensus-stack-gam, fig.width = 9, fig.height = 9}
# Same plotting recipe applied to the GAM's binary consensus stack.
terra::plot(gam_consensus$consensus_stack, nr = 4, nc = 4,
            mar = c(1.0, 1.0, 1.5, 3.0),
            legend = FALSE)
```

```{r consensus-stack-rf, fig.width = 9, fig.height = 9}
# Same plotting recipe applied to the random forest's binary consensus stack.
terra::plot(rf_consensus$consensus_stack, nr = 4, nc = 4,
            mar = c(1.0, 1.0, 1.5, 3.0),
            legend = FALSE)
```

```{r consensus-stack-hv, fig.width = 9, fig.height = 9}
# Same plotting recipe applied to the hypervolume's binary consensus stack.
terra::plot(hv_consensus$consensus_stack, nr = 4, nc = 4,
            mar = c(1.0, 1.0, 1.5, 3.0),
            legend = FALSE)
```

Reading these grids: yellow pixels were called suitable by **at least 2 of the 4 folds** (the `consensus = 2` majority rule), purple pixels were not. Compared to the raw fold-vote maps shown earlier under *Visualise the consensus maps*, these are the **binary** version after the consensus threshold has been applied.

### Frequency across years

The frequency raster collapses each consensus stack across years into a single 0–1 surface — the proportion of years a pixel was called suitable. Drawn per model as a 2 × 2 panel for easy side-by-side comparison.

```{r consensus-plot, fig.width = 10, fig.height = 8}
# Set up a 2 x 2 plot grid so each model's frequency raster occupies one
# panel, making side-by-side comparison straightforward.
par(mfrow = c(2, 2))

# Each $frequency_raster is a single 0-1 SpatRaster (proportion of years
# suitable). plot() draws it with the default viridis palette and `main`
# sets the panel title.
plot(glm_consensus$frequency_raster,
     main = "GLM — proportion of years suitable")

plot(gam_consensus$frequency_raster,
     main = "GAM — proportion of years suitable")

plot(rf_consensus$frequency_raster,
     main = "RF — proportion of years suitable")

plot(hv_consensus$frequency_raster,
     main = "Hypervolume — proportion of years suitable")

# Reset the plot region back to a single panel for anything that follows.
par(mfrow = c(1, 1))
```

Reading the panels: pixels close to 1 (bright) were predicted suitable in every modelled year — the *persistent core* of the niche for that model. Pixels close to 0 (dark) were never suitable. Intermediate values mark areas that gained or lost suitability across 2010–2025. Differences between the four panels show where the algorithms disagree on long-term suitability.

## Pixel-level temporal trajectories (changepoint analysis)

`analyze_temporal_patterns()` runs `fastcpd` per pixel and classifies each pixel's trajectory across years into one of six categories: never-suitable, always-suitable, no-pattern (statistically flat), increasing, decreasing, or fluctuating. The result is a categorical map of **how** habitat is changing rather than just *whether* it is suitable.

Key arguments:

- **`binary_stack`** — the per-time-step binary `consensus_stack` produced above.
- **`summary_raster`** — the matching 0–1 frequency raster; used to short-circuit pixels that are always or never suitable.
- **`time_steps`** — the same `data.frame` of time-step labels used during projection.
- **`spatial_autocorrelation` / `estimate_time`** — both `FALSE` here to keep the workshop fast; enable on real analyses if you need them.

```{r changepoints-glm}
# Run changepoint detection on the GLM's binary consensus stack to
# classify each pixel's trajectory across years (never-, always-,
# increasing, decreasing, fluctuating, no-pattern).
glm_patterns <- analyze_temporal_patterns(
  binary_stack = glm_consensus$consensus_stack, # binary per-year stack from summarize_raster_outputs()
  summary_raster = glm_consensus$frequency_raster, # 0-1 frequency raster, short-circuits always/never pixels
  time_steps = time_steps, # same data.frame of years used everywhere else
  output_dir = "outputs/glm_patterns", # where the pattern raster and CSV go
  spatial_autocorrelation = FALSE, # do NOT add a neighbours covariate (faster)
  estimate_time = FALSE, # skip the runtime-estimation pre-pass (faster)
  overwrite = TRUE, # replace any patterns left from a prior run
  verbose = FALSE # silence per-pixel progress messages
)
```

```{r changepoints-gam}
# Same call, applied to the GAM's consensus stack. Only the input rasters
# and the output folder change.
gam_patterns <- analyze_temporal_patterns(
  binary_stack = gam_consensus$consensus_stack,
  summary_raster = gam_consensus$frequency_raster,
  time_steps = time_steps,
  output_dir = "outputs/gam_patterns",
  spatial_autocorrelation = FALSE,
  estimate_time = FALSE,
  overwrite = TRUE,
  verbose = FALSE
)
```

```{r changepoints-rf}
# Same call, applied to the random forest's consensus stack.
rf_patterns <- analyze_temporal_patterns(
  binary_stack = rf_consensus$consensus_stack,
  summary_raster = rf_consensus$frequency_raster,
  time_steps = time_steps,
  output_dir = "outputs/rf_patterns",
  spatial_autocorrelation = FALSE,
  estimate_time = FALSE,
  overwrite = TRUE,
  verbose = FALSE
)
```

```{r changepoints-hv}
# Same call, applied to the hypervolume's consensus stack.
hv_patterns <- analyze_temporal_patterns(
  binary_stack = hv_consensus$consensus_stack,
  summary_raster = hv_consensus$frequency_raster,
  time_steps = time_steps,
  output_dir = "outputs/hv_patterns",
  spatial_autocorrelation = FALSE,
  estimate_time = FALSE,
  overwrite = TRUE,
  verbose = FALSE
)
```

The four trajectory-class maps, side-by-side:

```{r changepoints-plot, fig.width = 10, fig.height = 8}
# Set up a 2 x 2 plot region so the four models' trajectory maps appear
# side by side.
par(mfrow = c(2, 2))

# Each $pattern is a categorical SpatRaster whose pixel values encode the
# trajectory class (1 = never-suitable, 2 = always, 3 = no pattern,
# 4 = increasing, 5 = decreasing, 6 = fluctuating).
# `main` titles each panel; `legend = FALSE` keeps the grid tidy and we
# describe the categories in the prose below.
plot(glm_patterns$pattern,
     main = "GLM — temporal trajectory class",
     legend = FALSE)

plot(gam_patterns$pattern,
     main = "GAM — temporal trajectory class",
     legend = FALSE)

plot(rf_patterns$pattern,
     main = "RF — temporal trajectory class",
     legend = FALSE)

plot(hv_patterns$pattern,
     main = "Hypervolume — temporal trajectory class",
     legend = FALSE)

# Reset the plot region back to a single panel for anything that follows.
par(mfrow = c(1, 1))
```

Reading the panels: each colour is a categorical trajectory class. Comparing the four maps shows how much each algorithm agrees on the *shape* of change (increasing / decreasing / fluctuating), not just its location.

## Aggregate by spatial unit (DPKY forest complex)

`analyze_trends_by_spatial_unit()` aggregates the pixel trajectories produced by the changepoint analysis to a polygon set of our choosing. Up to this point everything in the workflow has been **Thailand-wide** — the models were fit on the country boundary, the cross-validation folds and pseudoabsences were drawn inside Thailand, and the trajectory rasters cover the whole country. For the final regional summary we narrow the spotlight to the **Dong Phayayen–Khao Yai Forest Complex (DPKY)** and break it down by its **five constituent protected areas**, so the per-unit numbers describe what happened inside each park or sanctuary individually rather than averaging across the whole complex.

The DPKY shapefile we loaded earlier ships with six polygon features, but only five distinct protected areas — two of the features both belong to *Dong Yai Wildlife Sanctuary* (`ดงใหญ่`). We dissolve those two features into a single multi-polygon so the regional aggregation returns exactly **five rows**, one per protected area:

1. **ทับลาน** — Thap Lan National Park.
2. **เขาใหญ่** — Khao Yai National Park (the first Thai national park established 1962, and the namesake of the complex).
3. **ตาพระยา** — Ta Phraya National Park.
4. **ปางสีดา** — Pang Sida National Park.
5. **ดงใหญ่** — Dong Yai Wildlife Sanctuary.

We also drop the Z (elevation) dimension from the geometry, because the shapefile ships as XYZ but the spatial-aggregation function expects plain 2-D geometry. Both clean-ups happen in a single short chunk below.

Key arguments to `analyze_trends_by_spatial_unit()`:

- **`shapefile_path`** — the cleaned DPKY polygon set (5 features after the merge). The function reprojects it internally to match the raster CRS if needed.
- **`name_field`** — the polygon attribute column carrying the human-readable unit name. We use **`Name_th`**, which holds the Thai-language name of each constituent protected area.
- **`binary_stack` / `pattern_raster` / `time_decrease_raster` / `time_increase_raster`** — outputs from the consensus and pattern steps for the model in question.

```{r dpky-unit-field}
# Two small clean-up steps make the DPKY polygon ready for the regional
# aggregation that follows:
#
# 1. sf::st_zm() drops the Z (elevation) dimension from every geometry.
#    The DPKY shapefile is XYZ on disk, but analyze_trends_by_spatial_unit()
#    and the underlying terra calls expect plain 2-D polygons; stripping
#    Z here prevents confusing errors later. `st_zm()` takes the sf object
#    as its only argument and returns an sf object whose CRS, attribute
#    columns, and polygon shapes are otherwise unchanged.
#
# 2. dplyr::group_by(Name_th) followed by dplyr::summarise() with
#    sf::st_union() dissolves any features that share the same Name_th
#    label into a single (multi-)polygon. The DPKY shapefile carries six
#    rows but only five distinct protected-area names, so this collapses
#    the two Dong Yai (ดงใหญ่) features into one row and leaves the
#    other four parks untouched. The argument `.groups = "drop"` removes
#    the leftover grouping metadata so the result is a plain ungrouped
#    sf object the next chunk can use directly.
dpky <- sf::st_zm(dpky) |>
  dplyr::group_by(Name_th) |>
  dplyr::summarise(geometry = sf::st_union(geometry),
                   .groups  = "drop")
```

```{r regional-glm, results = 'hide'}
# Aggregate the GLM's pixel-level trajectories to the DPKY polygon.
# The function overlays the polygon on the trajectory rasters, counts
# the pixels falling in each of the seven trajectory classes, and writes
# the result to a CSV inside `output_dir`.
#
# Arguments:
#   - shapefile_path:       the cleaned DPKY sf object (5 features, one
#                           row per constituent protected area).
#   - name_field:           "Name_th" — the polygon attribute column
#                           carrying the Thai-language park name.
#   - binary_stack:         per-year binary suitability stack from
#                           summarize_raster_outputs() for this model.
#   - pattern_raster:       categorical trajectory class per pixel.
#   - time_decrease_raster: year of first detected decrease per pixel
#                           (NaN where no decrease was detected).
#   - time_increase_raster: year of first detected increase per pixel
#                           (NaN where no increase was detected).
#   - time_steps:           same data.frame of modelled years (2010-2025)
#                           used for projection and changepoint detection.
#   - output_dir:           folder for the per-model summary CSVs and plots.
#   - overwrite:            TRUE replaces previous-run outputs in that folder.
#   - create_plot:          TRUE returns the diagnostic plots in `$plots`.
#   - verbose:              FALSE silences per-polygon progress messages.
glm_regional <- analyze_trends_by_spatial_unit(
  shapefile_path       = dpky,
  name_field           = "Name_th",
  binary_stack         = glm_consensus$consensus_stack,
  pattern_raster       = glm_patterns$pattern,
  time_decrease_raster = glm_patterns$time_decrease,
  time_increase_raster = glm_patterns$time_increase,
  time_steps           = time_steps,
  output_dir           = "outputs/glm_regional",
  overwrite            = TRUE,
  create_plot          = TRUE,
  verbose              = FALSE
)
```

```{r regional-glm-data}
# Pattern composition summary: counts and percentages of pixels in each
# of the seven trajectory classes inside DPKY.
glm_regional$overall_summary
```

```{r regional-gam, results = 'hide'}
# Same aggregation applied to the GAM outputs. Only the input rasters and
# the output folder change from the GLM call above; all other arguments
# are identical.
gam_regional <- analyze_trends_by_spatial_unit(
  shapefile_path       = dpky,
  name_field           = "Name_th",
  binary_stack         = gam_consensus$consensus_stack,
  pattern_raster       = gam_patterns$pattern,
  time_decrease_raster = gam_patterns$time_decrease,
  time_increase_raster = gam_patterns$time_increase,
  time_steps           = time_steps,
  output_dir           = "outputs/gam_regional",
  overwrite            = TRUE,
  create_plot          = TRUE,
  verbose              = FALSE
)
```

```{r regional-gam-data}
# Diagnostic DPKY-level pattern composition summary for the GAM.
gam_regional$overall_summary
```

```{r regional-rf, results = 'hide'}
# Same aggregation applied to the random forest outputs.
rf_regional <- analyze_trends_by_spatial_unit(
  shapefile_path       = dpky,
  name_field           = "Name_th",
  binary_stack         = rf_consensus$consensus_stack,
  pattern_raster       = rf_patterns$pattern,
  time_decrease_raster = rf_patterns$time_decrease,
  time_increase_raster = rf_patterns$time_increase,
  time_steps           = time_steps,
  output_dir           = "outputs/rf_regional",
  overwrite            = TRUE,
  create_plot          = TRUE,
  verbose              = FALSE
)
```

```{r regional-rf-data}
# Diagnostic DPKY-level pattern composition summary for the RF.
rf_regional$overall_summary
```

```{r regional-hv, results = 'hide'}
# Same aggregation applied to the hypervolume outputs.
hv_regional <- analyze_trends_by_spatial_unit(
  shapefile_path       = dpky,
  name_field           = "Name_th",
  binary_stack         = hv_consensus$consensus_stack,
  pattern_raster       = hv_patterns$pattern,
  time_decrease_raster = hv_patterns$time_decrease,
  time_increase_raster = hv_patterns$time_increase,
  time_steps           = time_steps,
  output_dir           = "outputs/hv_regional",
  overwrite            = TRUE,
  create_plot          = TRUE,
  verbose              = FALSE
)
```

```{r regional-hv-data}
# Diagnostic DPKY-level pattern composition summary for the hypervolume.
hv_regional$overall_summary
```

Putting the four `overall_summary` tables side by side tells you, *for each of the five protected areas*, what fraction of pixels each model labels as *always suitable* / *always absent* / *no pattern* / *increasing* / *decreasing* / *fluctuating* / *failed*. Where the four models agree on a park's composition, the algorithms are telling the same conservation story for that protected area; where they disagree — for instance, if GLM and GAM call Khao Yai mostly stable but the random forest detects substantial decreasing pixels there — that is exactly where additional ecological knowledge or field validation is needed before acting on the model.

# Wrap-up

What we covered:

- Read the pre-prepared annual precipitation, temperature, and NDVI rasters directly from `data/processed/temporal_split/`.
- Summarised the per-year aligned rasters as a single **temporal-mean** 3-layer stack (`mean_prec`, `mean_temp`, `mean_NDVI`) and used it as the reference E-space for thinning.
- Applied **`bean`** once against that temporal-mean E-space (Sheather–Jones bandwidth, stochastic per-pod thinning) to reduce environmental sampling bias across the full Sambar deer occurrence table before any temporal modelling.
- Built a fully temporally-explicit pipeline: time-stamped occurrences ↔ time-stamped rasters via `temporally_explicit_extraction()`.
- Used spatio-temporal cross-validation folds and fold-stratified pseudoabsences for an honest evaluation.
- Fitted **all four `TemporalModelR` algorithms** — GLM, GAM, random forest, and hypervolume — against the same fold structure and compared them on the same held-out test sets.
- Projected every model across 2010–2025 with `generate_spatiotemporal_predictions()` and visualised the resulting Thailand-wide fold-vote consensus rasters using the package's default `terra::plot()` styling (4 × 4 grid per model, one panel per year).
- Inspected per-fold E-space and per-time-step G-space performance with the package's default `plot_model_assessment()` diagnostic for each of the four models.
- Ran `summarize_raster_outputs()` and `analyze_temporal_patterns()` **once per model** on the Thailand-wide rasters so the consensus frequency surfaces and pixel-level trajectory classifications are directly comparable across GLM / GAM / RF / Hypervolume.
- Restricted only the *final* `analyze_trends_by_spatial_unit()` summary to the DPKY polygon, so the per-model pattern-composition tables describe what happened inside the Dong Phayayen–Khao Yai Forest Complex — the conservation-relevant management unit — without throwing away the country-wide niche estimate the four models were built on.

# References

- Castaneda-Guzman M, Hughes C, Paansri P, Cobos M (2026). *nicheR: Ellipsoid-Based Virtual Niches and Visualization*. R package version 0.1.0, <https://github.com/castanedaM/nicheR>.
- Chamberlain S, Barve V, Mcglinn D, Oldoni D, Desmet P, Geffert L, Ram K (2026). *rgbif: Interface to the Global Biodiversity Information Facility API*. R package version 3.8.5.2, <https://CRAN.R-project.org/package=rgbif>.
- Hijmans R (2026). *geodata: Access Geographic Data*. R package version 0.6-10, <https://rspatial.github.io/geodata/>.
- Hughes C, Castaneda-Guzman M, E. Escobar L (2026). *TemporalModelR: Temporally Explicit Species Distribution Modelling in R*. R package version 0.2.0, <https://github.com/CJHughes926/TemporalModelR>.
- Paansri P, Escobar L (2026). *bean: Data Thinning of Species Occurrences in Environmental Space*. R package version 0.2.1, <https://github.com/paanwaris/bean>.
