---
title: "Step 2 — Environmental thinning with **bean**"
subtitle: "ENM-Thailand workshop · Sambar deer (*Rusa unicolor*)"
author: "Workshop materials"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: cosmo
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  dpi       = 100
)
```

# Overview

> **Workshop navigation**
>
> - [**Pre-workshop** — Downloading the workshop data.](https://paanwaris.github.io/ENM-Thailand/articles/Pre-workshop.html) Get the Thailand boundary, WorldClim climate layers, and GBIF Sambar records before the workshop starts.
> - [**Step 1: nicheR** — Virtual species with nicheR.](https://paanwaris.github.io/ENM-Thailand/articles/Step1_nicheR.html) Build a virtual species niche and map it.
> - **Step 2: bean — Environmental thinning with bean.** Reduce environmental sampling bias before fitting a real niche.
> - [**Step 3: TemporalModelR** — Temporally-explicit models with TemporalModelR.](https://paanwaris.github.io/ENM-Thailand/articles/Step3_TemporalModelR.html) Link species occurrences with environmental data through time.

Some occurrence datasets are heavily **biased**: field collecting sample in easily accessible places (roads, protected areas, research stations) far more often than remote interiors, so the records cluster on the map. The classical fix is to **thin in geographic space (G-space)**, keep at most one point per spatial grid cell. That removes spatial duplication, but it has a hidden cost: two records that sit close together on the map can capture very different environmental conditions (one on a ridge, one in a valley), and G-space thinning discards that information.

`bean` solves the problem differently. It thins **in environmental space (E-space)**, keeping at most one point per cell of an environmental grid. The intuition is that two occurrences with nearly identical climate values are *environmentally redundant*, no matter how far apart they are on the map. Thinning per environmental cell (one "pod") therefore produces a more even coverage of the species' niche and a less biased niche estimate.

In this Step we will:

1.  Load the cleaned GBIF occurrences for Sambar deer in Thailand (written by `Pre-workshop.Rmd`).
2.  Prepare the occurrences for `bean` and visualise the bias in E-space.
3.  Find an **objective** environmental grid resolution.
4.  Thin the points with both methods (`thin_env_nd` stochastic, `thin_env_center` deterministic).
5.  Fit an ellipsoid niche on the thinned points and map suitability back to G-space.

All function names and arguments below follow the official package documentation: <https://paanwaris.github.io/bean/>

# Install and load packages

```{r install, eval = FALSE}
# ---- CRAN dependencies -----------------------------------------------------
cran_pkgs <- c(
  "terra",
  "sf",
  "dplyr",
  "readr",
  "rmarkdown",
  "knitr",
  "remotes",
  "rgl",
  "bean"
)
```

```{r packages}
library(bean)     # environmental thinning in E-space
library(terra)    # modern raster + vector spatial backend
library(sf)       # tidy vector spatial data ("simple features")
library(dplyr)    # tidy data manipulation
library(readr)    # fast, friendly CSV reading and writing
library(rgl)      # interactive 3-D rendering for the ellipsoid plots

# rgl::setupKnitr() lets knitr embed any rgl/WebGL plot as a rotatable
# widget directly in the rendered HTML page.
rgl::setupKnitr(autoprint = TRUE)

# Make every random draw in the rest of the file reproducible.
set.seed(2026)
```

```{r load-data}
# Inputs prepared by the Pre-workshop notebook.
bioclim <- rast("data/processed/bioclim/bioclim_thailand.tif")
thailand <- st_read("data/processed/boundary/thailand_boundary.gpkg",
                    quiet = TRUE)
sambar <- read_csv("data/processed/gbif/sambar_thailand.csv",
                     show_col_types = FALSE)

head(sambar)
```

For this exercise we use three bioclim variables, which map the Sambar deer habitat (warm, wet, evergreen and mixed forests).

```{r pick-vars}
# Subset the 19-layer bioclim stack down to the three variables that will
# define the ellipsoid axes.
env <- bioclim[[c("bio1", "bio12", "bio15")]]
names(env) <- c("bio1", "bio12", "bio15")
```

# Prepare occurrences

`prepare_bean()` drops rows with missing coordinates, extracts environmental values at each occurrence, optionally rescales the variables, and returns a tidy data frame that is ready for thinning.

```{r prepare}
# Slim the full GBIF table down to just the coordinate pair. The two
# arguments to dplyr::select() are *renames*: `longitude = decimalLongitude`
# reads as "create a new column called `longitude` whose values come from
# GBIF's `decimalLongitude` column", and likewise for latitude. The renaming
# step is important because prepare_bean() expects the columns to be called
# exactly `longitude` and `latitude`.
sambar_xy <- sambar %>%
  dplyr::select(longitude = decimalLongitude,
                latitude  = decimalLatitude)

# prepare_bean() does three jobs in one call:
#   1) drops any row whose longitude/latitude is missing;
#   2) extracts the value of every environmental layer at each remaining
#      occurrence pixel;
#   3) optionally rescales those environmental values.
# It returns a tidy data frame with one row per occurrence and one column
# per environmental variable, ready to feed into the thinning step.
prepped <- prepare_bean(
  data        = sambar_xy,    # the input table of (longitude, latitude) pairs
  env_rasters = env,          # the SpatRaster stack we will extract values from
  longitude   = "longitude",  # name of the column in `data` carrying x coords
  latitude    = "latitude",   # name of the column in `data` carrying y coords
  transform   = "scale"       # convert every env variable to a z-score
                              # (mean 0, sd 1) so BIO1 (degrees), BIO12 (mm),
                              # and BIO15 (CV) share a comparable numeric range
)

# Print the number of usable occurrences left after prepare_bean() dropped
# rows that fell outside the Thailand raster mask or had missing coordinates.
nrow(prepped)
```

Inspect the raw points in G-space and in E-space side by side. Notice the clustering in E-space, that is the environmental over-sampling `bean` will reduce.

```{r raw-bias, fig.width = 9, fig.height = 4.5}
# Set up a one-row, two-column plotting region so the next two plot() calls
# appear side by side.
par(mfrow = c(1, 2))

# LEFT panel: the Thailand outline only.
# `st_geometry(thailand)` strips the attribute table off the sf object so we
# get just the polygon shape (no fill). `border = "grey20"` draws the outline
# in dark grey, and `main` sets the panel title.
plot(st_geometry(thailand), border = "grey20", main = "G-space")

# Overlay the occurrence points on top of the polygon.
# `prepped[, c("longitude", "latitude")]` selects just the coordinate columns.
# `pch = 21` is a filled circle with a separate border colour; `bg = "tomato"`
# fills the circle red-orange; `col = "white"` draws a thin white border
# around each dot so individual points stay visible even where they overlap;
# `cex = 0.7` shrinks the marker to 70% of the default size.
points(prepped[, c("longitude", "latitude")],
       pch = 21, bg = "tomato", col = "white", cex = 0.7)

# RIGHT panel: the same occurrences plotted in environmental space.
# The first two arguments to plot() are the x and y vectors — here BIO1
# (annual mean temperature) on the x-axis and BIO12 (annual precipitation)
# on the y-axis, both already z-scored by prepare_bean().
# `xlab` / `ylab` set readable axis labels.
# `pch = 19` is a solid filled circle (no separate border).
# `adjustcolor("tomato", alpha.f = 0.5)` takes the tomato red colour and
# makes it 50% transparent so overlapping points show as darker patches —
# a quick visual diagnostic of environmental over-sampling.
# `main` sets the panel title.
plot(prepped$bio1, prepped$bio12,
     xlab = "BIO1 — annual mean T (z-score)",
     ylab = "BIO12 — annual precip. (z-score)",
     pch  = 19,
     col  = adjustcolor("tomato", alpha.f = 0.5),
     main = "E-space")

# Reset the plotting region back to a single panel for any subsequent plot.
par(mfrow = c(1, 1))
```

Reading the panels: **left**, every red dot is one Sambar deer GBIF record on the map of Thailand — the dots concentrate on the western mountains and the central plains because that is where surveys happen most. **Right**, the same records replotted in the environmental plane defined by BIO1 (annual mean temperature) and BIO12 (annual precipitation).

# Choose an objective environmental resolution

`find_env_resolution()` derives a grid resolution for thinning directly from the data. It does so **per variable**, meaning it returns *one* suggested cell width for each environmental axis (so BIO1 gets its own width, BIO12 gets a different width, BIO15 gets yet another), each in the natural units of that variable. The widths come from a **kernel-density bandwidth** of each variable's marginal distribution: at the bandwidth scale, two occurrences carry essentially the same environmental information, so it is a natural cell width below which extra records are redundant.

Three classic bandwidth selectors are supported:

- `"sheather-jones"` *(default)* — Sheather & Jones (1991) plug-in selector; robust to non-Gaussian shapes and the modern recommended default.
- `"silverman"` — Silverman's rule of thumb; fast and stable when the distribution is roughly Gaussian.
- `"scott"` — Scott's rule; the Gaussian-optimal variant of Silverman.

If Sheather–Jones fails (this can happen with strongly tied data) the function automatically falls back to Silverman's rule for that variable and emits a message.

We will run all three selectors and compare them, then pick one to feed into the thinning step.

```{r find-res-three}
# Call find_env_resolution() three times with the same input data and the
# same set of environmental variables, varying only the bandwidth selector
# (`method`). Each call returns a suggested cell width *per variable*
# stored in `$suggested_resolution`.
# Arguments shared by all three calls:
#   - First positional argument (`prepped`): the data frame produced by
#     prepare_bean() — bean reads the environmental columns from here.
#   - `env_vars`: a character vector naming which columns of `prepped` to
#     use as the environmental axes; here our three bioclim variables.
#   - `method`: the bandwidth-selection rule (see prose above each).
res_sj <- find_env_resolution(prepped,
                              env_vars = c("bio1", "bio12", "bio15"),
                              method = "sheather-jones")
res_silverman <- find_env_resolution(prepped,
                                     env_vars = c("bio1", "bio12", "bio15"),
                                     method = "silverman")
res_scott <- find_env_resolution(prepped,
                                 env_vars = c("bio1", "bio12", "bio15"),
                                 method = "scott")

# Build a side-by-side comparison table.
# Each `$suggested_resolution` is a named numeric vector (one entry per
# variable). cbind() glues those three vectors together as columns, and
# the `name = value` syntax inside cbind() labels each column with a clean
# selector name. Read each row to compare what the three selectors picked
# for the same bioclim variable, in z-score units.
cbind(
  sheather_jones = res_sj$suggested_resolution,
  silverman = res_silverman$suggested_resolution,
  scott = res_scott$suggested_resolution
)
```

Read across each row to compare. Silverman and Scott are simple plug-ins that assume a roughly Gaussian distribution, so they usually return *similar*, slightly *larger* bandwidths, coarser cells, fewer pods, more aggressive thinning. Sheather–Jones is data-adaptive and is usually *smaller*, finer cells, more pods, gentler thinning. For real occurrence data, which is almost never Gaussian, we recommend **`"sheather-jones"`**: it preserves more environmental nuance and is the safest default for the workshop.

Each method's diagnostic plot. Each panel below is the kernel-density estimate (KDE) of one bioclim variable; the red horizontal scale bar marks the bandwidth that the selector chose — i.e. the suggested cell width.

```{r find-res-plots-sj, fig.height = 4}
plot(res_sj)
```

```{r find-res-plots-silverman, fig.height = 4}
plot(res_silverman)
```

```{r find-res-plots-scott, fig.height = 4}
plot(res_scott)
```

Reading the panels: the **black curve** is the kernel-density estimate of how the occurrence points are distributed along that environmental axis; the **shaded area** under the curve helps the eye see where the bulk of the density sits; the **red horizontal bar** is the suggested cell width (= the bandwidth) centred on the mode of the density. A narrower bar means a finer grid and gentler thinning; a wider bar means a coarser grid and more aggressive thinning.

We will use Sheather–Jones for the rest of the analysis:

```{r chosen-res}
# Pull the Sheather-Jones suggested cell widths out of the result object and
# store them in `chosen_res`. This is the named numeric vector (one entry per
# variable) we will feed into the thinning functions below as the
# `grid_resolution` argument.
chosen_res <- res_sj$suggested_resolution
chosen_res
```

# Apply thinning

`bean` ships two thinning strategies:

- **`thin_env_nd()` — stochastic.** Builds the grid, finds every occupied pod, and *randomly retains one of the original occurrences* in each pod. Real coordinates and GBIF metadata are preserved.
- **`thin_env_center()` — deterministic.** Builds the grid and *replaces the points in each occupied pod with a single synthetic point at the centre of the pod*. Output is reproducible without a seed, but the points are no longer real observations.

We will run both for comparison.

```{r thin}
# Stochastic thinning: build an environmental grid over the env_vars axes
# and *randomly retain one of the original occurrence points* in each
# occupied grid cell ("pod"). Real coordinates and GBIF metadata are kept.
thin_nd <- thin_env_nd(
  data = prepped, # output of prepare_bean(): one row per occurrence
  env_vars = c("bio1", "bio12", "bio15"), # which columns of `data` define the env grid axes
  grid_resolution = chosen_res # named vector of per-variable cell widths (from Sheather-Jones)
)

# Deterministic thinning: build the same kind of grid, but instead of
# keeping a random real point per pod, REPLACE every occupied pod with a
# single synthetic point sitting at the centre of that pod. Output is
# perfectly reproducible without a seed, at the cost of losing the link
# back to the original GBIF records.
# We deliberately use a fixed, coarse resolution (0.5 z-score units on
# every axis). Because the data were scaled in prepare_bean(), 0.5 means
# half a standard deviation per axis — a sensible "moderate smoothing"
# choice that is easy to interpret and visualise alongside the stochastic
# result above.
thin_center <- thin_env_center(
  data = prepped, # same prepare_bean() output as above
  env_vars = c("bio1", "bio12", "bio15"), # same three axes
  grid_resolution = c(0.5, 0.5, 0.5) # fixed 0.5-SD-wide cells on every axis
)

# Printing a `thin_*` object shows a one-line summary: the number of input
# occurrences, the number of occupied pods, and the retained sample size.
# The actual thinned table of points lives in `thin_*$thinned_data`.
thin_nd
thin_center
```

```{r thin-vs-raw, fig.width = 9, fig.height = 4.5}
# plot_bean() draws an E-space scatter that overlays the *original* cloud
# of points (in faded grey) with the points retained by a thinning call
# (highlighted), so the visual reduction in density is obvious at a glance.
#
# Arguments used by both calls below:
#   - original_data:  the prepare_bean() data frame (the full, un-thinned set)
#   - thinned_object: the object returned by a thin_env_*() call
#   - env_vars:       which columns of the data frame to use as plot axes;
#                     here our three bioclim variables
plot_bean(
  original_data = prepped, # full, un-thinned table
  thinned_object = thin_nd, # stochastic thinning result
  env_vars = c("bio1", "bio12", "bio15") # axes of the scatter
)

plot_bean(
  original_data = prepped, # same full table
  thinned_object = thin_center, # deterministic thinning result
  env_vars = c("bio1", "bio12", "bio15")
)
```

Reading the plots: the **pale grey dots** are the original occurrences (the same cloud you saw in the E-space panel earlier); the **highlighted markers** are the points that survive thinning. With `thin_env_nd` the highlighted markers sit *on top of* the original cloud (because real points are kept); with `thin_env_center` they sit on the regular grid points at the centres of occupied pods (synthetic locations). Notice that both methods break up the dense cluster that dominated the raw E-space plot.

# Fit the ellipsoid niche

`fit_ellipsoid()` formalises the niche by fitting a multivariate ellipsoid around a set of points. We fit two ellipsoids, one on the raw occurrences and one on the thinned ones, to see how thinning shifts the niche estimate.

```{r fit-ellipsoid}
# fit_ellipsoid() draws a multivariate ellipsoid around the supplied points
# in environmental space. The shape (size, orientation, elongation) is the
# package's mathematical representation of the niche.
#
# Arguments shared by both calls:
#   - First positional argument: the data frame of points to fit on.
#   - env_vars: which columns of that data frame become the ellipsoid axes.
#   - method = "covmat": estimate the ellipsoid from the *covariance matrix*
#                        of the points (standard, robust choice).
#   - level   = 0.95:    enclose 95% of the multivariate distribution within
#                        the ellipsoid boundary (the conventional confidence
#                        level for niche modelling).

# Raw ellipsoid: fit on the full, un-thinned set of occurrences. This fit
# is biased toward the densest sampled environments because clustered
# points carry more weight than isolated ones in the covariance estimate.
ell_raw <- fit_ellipsoid(
  prepped, # un-thinned data
  env_vars = c("bio1", "bio12", "bio15"),
  method = "covmat",
  level = 0.95
)

# Thinned ellipsoid: fit on the table inside thin_nd$thinned_data — i.e.
# the stochastically thinned points (one per pod). Because each pod
# contributes at most one record, no environmental region is over-weighted.
ell_thinned <- fit_ellipsoid(
  thin_nd$thinned_data, # thinned data (one point per pod)
  env_vars = c("bio1", "bio12", "bio15"),
  method = "covmat",
  level = 0.95
)
```

Visualise the **raw** ellipsoid first, both in 2-D (a static projection onto BIO1 × BIO12) and in 3-D (a rotatable WebGL widget):

```{r ell-3d-1, webgl = TRUE, fig.width = 7, fig.height = 6}
# Static 2-D view of the raw ellipsoid.
# `dims = c(1, 2)` tells the plot method to draw the projection onto the
# first two environmental axes (here BIO1 on x, BIO12 on y) and to ignore
# BIO15 — useful as a quick, printable cross-section of the niche.
plot(ell_raw, dims = c(1, 2))

# Interactive 3-D view of the same raw ellipsoid.
# `dims = c(1, 2, 3)` asks for all three axes (BIO1, BIO12, BIO15) so the
# full ellipsoidal surface is drawn. Because the chunk option `webgl = TRUE`
# is set in the chunk header, this renders as a rotatable WebGL widget you
# can spin with the mouse to inspect the niche from any angle.
plot(ell_raw, dims = c(1, 2, 3))
```

Reading the plots: in 2-D, the **ellipse** is the 95% confidence boundary of the niche projected onto BIO1 × BIO12; the **dots** are the occurrence points used to fit the niche. In 3-D, the **translucent mesh** is the full ellipsoidal surface in BIO1 × BIO12 × BIO15 space. Drag the 3-D plot with the mouse to rotate it and inspect the niche from any angle.

Now compare with the **thinned** ellipsoid:

```{r ell-3d-2, webgl = TRUE, fig.width = 7, fig.height = 6}
# Same two views, but for the *thinned* ellipsoid. Comparing this 2-D
# projection with the raw one above shows how the niche shape changes once
# the dense E-space cluster has been broken up by bean.
plot(ell_thinned, dims = c(1, 2))

# 3-D WebGL widget of the thinned ellipsoid. Rotate alongside the raw 3-D
# plot from the previous chunk to see how thinning typically *widens* the
# fit — the dense cluster no longer drags the ellipsoid in toward itself.
plot(ell_thinned, dims = c(1, 2, 3))
```

The thinned ellipsoid is typically a little *wider* than the raw one, because removing the dense cluster of redundant points lets the fit reach further into the niche's true extent rather than collapsing around the over-sampled region.

# Project the niche back to G-space

Projection is delegated to the **nicheR** package, which has the same ellipsoid representation under the hood. We attach `nicheR` on the fly, then call `nicheR::predict()` on the scaled bioclim raster.

```{r predict}
# Make sure nicheR is available — it provides the ellipsoid -> raster
# projection used here. requireNamespace() returns TRUE if the package is
# already installed and FALSE if not; the `!` flips that, so we only run
# install_github() when nicheR is missing.
if (!requireNamespace("nicheR", quietly = TRUE)) {
  remotes::install_github("castanedaM/nicheR", upgrade = "never")
}
library(nicheR)

# Project each ellipsoid back onto the geographic landscape.
# Arguments to predict():
#   - First positional argument: the fitted ellipsoid object (ell_raw or
#     ell_thinned). nicheR's predict() method dispatches on its class.
#   - Second positional argument: the environmental raster to project onto.
#     We wrap it in scale() because both ellipsoids were fit on z-scored
#     points (we passed `transform = "scale"` to prepare_bean()), and the
#     raster and the points MUST share the same transformation for the
#     projection to be meaningful.
# The return value contains a `$suitability` SpatRaster layer that we plot
# in the next step.
pred_raw <- predict(ell_raw, scale(env))
pred_thinned <- predict(ell_thinned, scale(env))

# Two side-by-side suitability maps so the raw vs. thinned comparison is
# visually direct.
par(mfrow = c(1, 2))

# LEFT panel: suitability projected from the *raw* ellipsoid.
# `pred_raw$suitability` is a 0-1 SpatRaster; plot() draws it with terra's
# default viridis-like palette; `main` sets the panel title.
plot(pred_raw$suitability, main = "Suitability — raw data")

# RIGHT panel: same projection, but from the *thinned* ellipsoid.
plot(pred_thinned$suitability, main = "Suitability — thinned niche")

# Reset the plotting region back to a single panel for whatever comes next.
par(mfrow = c(1, 1))
```

Reading the plots: both panels show suitability as a 0–1 surface (yellow = high, dark = low). The **left** panel uses the biased, unthinned ellipsoid; the **right** uses the thinned one. The thinned suitability map is typically broader and reaches further into climatically appropriate but under-sampled areas — that is exactly the bias correction `bean` is designed to deliver.

```{r save}
# Write the thinned occurrence table to disk so Step 3 can re-use it.
# Arguments to write.csv():
#   - First positional argument: the data frame to save. ell_thinned
#     carries a slot called `$all_points_used` which is the table of
#     occurrences that actually fed into the thinned ellipsoid fit.
#   - Second positional argument: the output file path. Saving under
#     data/processed/gbif/ keeps it next to the other GBIF outputs.
#   - row.names = FALSE: suppress the default row-number column R would
#     otherwise prepend to the CSV. Without this you get an extra unnamed
#     index column that downstream readers (Step 3) would mistake for data.
write.csv(ell_thinned$all_points_used,
          "data/processed/gbif/sambar_thailand_thinned.csv",
          row.names = FALSE)
```

# Wrap-up

Take-homes:

- Spatial bias in occurrence data **propagates into E-space** as dense clusters.
- `bean` thins *in environmental space* so the fitted niche is not pulled toward the densest sampled environments.
- `find_env_resolution()` picks the grid cell width objectively from a KDE bandwidth — Sheather–Jones is the recommended default for real occurrence data.
- The two thinning methods trade randomness for fidelity to the original records:
  - `thin_env_nd` (stochastic) **keeps real coordinates and GBIF metadata** (dates, recorder IDs, `gbifID`, etc.). Set the random seed for reproducibility.
  - `thin_env_center` (deterministic) replaces real records with synthetic pod centres. Reproducible without a seed, but you lose every original attribute.
- **For Sambar deer in this workshop we recommend `thin_env_nd`.** The GBIF metadata (especially the `year` column) is needed for the temporally-explicit modelling in **Step 3: TemporalModelR**.

# References

- Castaneda-Guzman M, Hughes C, Paansri P, Cobos M (2026). *nicheR: Ellipsoid-Based Virtual Niches and Visualization*. R package version 0.1.0, <https://github.com/castanedaM/nicheR>.
- Chamberlain S, Barve V, Mcglinn D, Oldoni D, Desmet P, Geffert L, Ram K (2026). *rgbif: Interface to the Global Biodiversity Information Facility API*. R package version 3.8.5.2, <https://CRAN.R-project.org/package=rgbif>.
- Hijmans R (2026). *geodata: Access Geographic Data*. R package version 0.6-10, <https://rspatial.github.io/geodata/>.
- Hughes C, Castaneda-Guzman M, E. Escobar L (2026). *TemporalModelR: Temporally Explicit Species Distribution Modelling in R*. R package version 0.2.0, <https://github.com/CJHughes926/TemporalModelR>.
- Paansri P, Escobar L (2026). *bean: Data Thinning of Species Occurrences in Environmental Space*. R package version 0.2.1, <https://github.com/paanwaris/bean>.
