---
title: "Step 3 — Temporally-explicit SDMs with **TemporalModelR**"
subtitle: "ENM-Thailand workshop · Sambar deer (*Rusa unicolor*)"
author: "Workshop materials"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: cosmo
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  dpi       = 100
)
```

# Overview

> **Workshop navigation**
>
> - [**Pre-workshop** — Downloading the workshop data.](https://paanwaris.github.io/ENM-Thailand/articles/Pre-workshop.html) Get the Thailand boundary, WorldClim climate layers, and GBIF Sambar records before the workshop starts.
> - [**Step 1: nicheR** — Virtual species with nicheR.](https://paanwaris.github.io/ENM-Thailand/articles/Step1_nicheR.html) Build a virtual species niche and map it.
> - [**Step 2: bean** — Environmental thinning with bean.](https://paanwaris.github.io/ENM-Thailand/articles/Step2_bean.html) Reduce environmental sampling bias before fitting a real niche.
> - **Step 3: TemporalModelR — Temporally-explicit models with TemporalModelR.** Link species occurrences with environmental data through time.

`TemporalModelR` builds **temporally-explicit** species distribution models. Each occurrence is paired with the environmental conditions *at the time it was recorded*, rather than against a long-term average. Models can then project predictions as time-series surfaces and detect pixel-level temporal trajectories (always-suitable, never-suitable, increasing, decreasing, or fluctuating).

> **Data.** Step 3 uses the pre-prepared annual rasters that live under `data/processed/temporal_split/` — one precipitation (`prec_YYYY.tif`), one temperature (`temp_YYYY.tif`, derived from land-surface temperature), and one NDVI (`NDVI_YYYY.tif`) file per year. **No new climate data is downloaded here** — we simply read the local files and stack them. The `Pre-workshop.Rmd` notebook produced the Sambar deer GBIF table and the Thailand polygon we use alongside them.

In this Step we will:

1.  Read the per-year precipitation, temperature, and NDVI rasters and align them to a common grid.
2.  Build a temporal-mean E-space, thin the GBIF Sambar deer occurrences **once** with `bean` against it, then rarefy with `spatiotemporal_rarefaction()`, and pair each surviving record with the environment it actually experienced using `temporally_explicit_extraction()`.
3.  Z-score the rasters, summarise their temporal mean, build spatiotemporal CV folds, and create fold-stratified pseudoabsences.
4.  Fit `build_temporal_glm()` and `build_temporal_rf()` models and project predictions across years with `generate_spatiotemporal_predictions()`.
5.  Postprocess with `summarize_raster_outputs()`, `analyze_temporal_patterns()`, and `analyze_trends_by_spatial_unit()`.

All function names and arguments below follow the official package documentation: <https://cjhughes926.github.io/TemporalModelR/>

# Install and load packages

```{r install, eval = FALSE}
# ---- CRAN dependencies -----------------------------------------------------
cran_pkgs <- c(
  "terra",
  "sf",
  "dplyr",
  "readr",
  "viridis",
  "randomForest",
  "fastcpd",
  "rmarkdown",
  "knitr",
  "remotes",
  "geodata",
  "bean"        
)
to_install <- cran_pkgs[!cran_pkgs %in% rownames(installed.packages())]
if (length(to_install)) install.packages(to_install, dependencies = TRUE)

# ---- TemporalModelR from GitHub --------------------------------------------
if (!requireNamespace("TemporalModelR", quietly = TRUE)) {
  remotes::install_github("CJHughes926/TemporalModelR")
}
```

```{r packages}
library(TemporalModelR)  # temporally-explicit SDM pipeline
library(bean)            # per-year environmental thinning
library(terra)           # modern raster + vector spatial backend
library(sf)              # tidy vector spatial data ("simple features")
library(dplyr)           # tidy data manipulation
library(readr)           # fast, friendly CSV reading and writing

# Make every random draw in the rest of the file reproducible.
set.seed(2026)
```

# Inputs

```{r load-base}
# Thailand polygon (from the Pre-workshop). `quiet = TRUE` suppresses the
# usual GDAL chatter so the rendered page stays clean.
thailand <- st_read("data/processed/boundary/thailand_boundary.gpkg",
                    quiet = TRUE)

# Clean GBIF Sambar deer table from the Pre-workshop. We:
#   * restrict to 2010-2025 (the range covered by every per-year raster);
#   * rename the GBIF coordinate columns to the conventional `longitude` /
#     `latitude` so they match the names bean and TemporalModelR expect;
#   * add a `presence = 1L` column so the table is ready to be combined
#     with pseudoabsences (presence = 0) later on.
sambar <- read_csv("data/processed/gbif/sambar_thailand.csv",
                   show_col_types = FALSE) %>%
  filter(year >= 2010, year <= 2025) %>%       # keep modelled years only
  rename(longitude = decimalLongitude,         # GBIF -> conventional name
         latitude  = decimalLatitude) %>%      # GBIF -> conventional name
  mutate(presence = 1L)                        # every GBIF record is a presence

# How many records per year survive that filter?
table(sambar$year)
```

# Preprocessing

## Align rasters

`raster_align()` reprojects and resamples every raster found in `input_dir` against a reference, then writes the aligned copies into `output_dir`. Because the three variables come from different sources, they do not share an extent, resolution, or CRS out of the box — we use one of the per-year NDVI files as the reference grid for everything.

```{r raster-align}
# Quick visual sanity-check of the three "current" layers.
prec_ex <- rast("data/processed/temporal_split/prec_2025.tif")
temp_ex <- rast("data/processed/temporal_split/temp_2025.tif")
NDVI_ex <- rast("data/processed/temporal_split/NDVI_2025.tif")

par(mfrow = c(1, 3))
plot(prec_ex, main = "Precipitation 2025")
plot(temp_ex, main = "Temperature 2025")
plot(NDVI_ex, main = "NDVI 2025")
par(mfrow = c(1, 1))     # reset the plot region back to a single panel

# Source directory of the per-variable per-year split rasters, and target
# directory where the aligned copies will be written.
split_dir   <- "data/processed/temporal_split"
aligned_dir <- "data/processed/rasters_aligned"
dir.create(aligned_dir, showWarnings = FALSE, recursive = TRUE)

# Use NDVI as the reference grid — every per-year raster will be reprojected
# / resampled to match its extent, resolution, and CRS.
ref_raster <- NDVI_ex

raster_align(
  input_dir        = split_dir,    # where the per-year rasters currently live
  output_dir       = aligned_dir,  # where the aligned copies will be written
  reference_raster = ref_raster,   # target grid (extent + resolution + CRS)
  overwrite        = TRUE          # replace any aligned files left from a prior run
)
```

## Build a temporal-mean E-space for thinning

Before any thinning, we build a single **temporal-mean** reference E-space that summarises the environment Sambar deer experienced across the entire 2010–2025 window. We compute the per-pixel mean across years for each of the three aligned variables, then stack the three mean surfaces into one 3-layer `SpatRaster`.

This temporal-mean stack will be the **reference environmental space** that `bean` uses to thin every year. Using a single, period-wide E-space (rather than the year-specific raster of each occurrence) keeps the thinning grid *consistent across years*: a 2014 record and a 2020 record that fall in the same temporal-mean pod are treated as environmentally redundant, even though their individual years' rasters may differ slightly.

```{r temporal-mean-env}
# List every aligned per-year raster, then split by variable name.
aligned_files <- list.files(aligned_dir, pattern = "\\.tif$", full.names = TRUE)

# Group the file list by variable name (prec_ / temp_ / NDVI_) so each
# group becomes a multi-layer SpatRaster spanning every year.
prec_files <- file.path(aligned_dir,
                        grep("prec_",  basename(aligned_files), value = TRUE))
temp_files <- file.path(aligned_dir,
                        grep("temp_",  basename(aligned_files), value = TRUE))
NDVI_files <- file.path(aligned_dir,
                        grep("NDVI_",  basename(aligned_files), value = TRUE))

# Per-pixel mean across years for each variable — terra::mean() over a
# multi-layer SpatRaster collapses the time dimension into a single layer.
mean_prec <- terra::mean(rast(prec_files), na.rm = TRUE)
mean_temp <- terra::mean(rast(temp_files), na.rm = TRUE)
mean_NDVI <- terra::mean(rast(NDVI_files), na.rm = TRUE)

# Stack the three mean surfaces into one 3-layer reference E-space, and
# rename the layers to the variable names bean will look up.
env_ref <- c(mean_prec, mean_temp, mean_NDVI)
names(env_ref) <- c("mean_prec", "mean_temp", "mean_NDVI")

par(mfrow = c(1, 3))
plot(mean_prec, main = "Mean precipitation (2010–2025)")
plot(mean_temp, main = "Mean temperature (2010–2025)")
plot(mean_NDVI, main = "Mean NDVI (2010–2025)")
par(mfrow = c(1, 1))     # reset the plot region back to a single panel
```

Reading the panels: each map is the per-pixel average of the corresponding variable across the 16 modelled years, in the variable's natural units (mm for precipitation, °C-equivalent units for land-surface-temperature-derived `temp`, and a unitless 0–1 index for NDVI). Bright patches stand for places that were *consistently* wet, warm, or vegetated; dark patches stand for places that were *consistently* dry, cool, or sparse. This is the environmental landscape `bean` will use to define the thinning grid.

## Environmental thinning with `bean`

We reduce **environmental sampling bias** in the Sambar deer occurrences using `bean`, with the temporal-mean stack `env_ref` from the previous section as the reference E-space. Because the reference E-space is a single, period-wide summary (not year-specific), we run `bean` **once on the full occurrence table** rather than year by year — every record is judged against the same environmental landscape, so two records that fall in the same pod are environmentally redundant regardless of which year they were collected in.

We use the **`"sheather-jones"`** bandwidth selector inside `find_env_resolution()`. Sheather–Jones is a data-adaptive, plug-in selector (Sheather & Jones, 1991) that is robust to the non-Gaussian shapes typical of real occurrence data — it is the modern recommended default and is what the workshop uses everywhere bean appears. We pair it with the stochastic `thin_env_nd()` so each pod retains one of the original GBIF coordinates rather than a synthetic centroid.

Every column of the original `sambar` table is preserved on the output. After thinning bean returns only the longitude/latitude pair, so we recover the full original attributes (including the `year` column the temporal modelling needs) by matching on the coordinate pair.

```{r thin-once}
# Pass only the coordinate pair into bean; the full attribute table is
# recovered below by joining the thinned coordinates back onto `sambar`.
sambar_xy <- sambar[, c("longitude", "latitude")]

# Extract the three temporal-mean env values at every occurrence, and
# z-score them so all three axes share the same numeric scale.
prepped <- prepare_bean(
  data        = sambar_xy,
  env_rasters = env_ref,        # temporal-mean 3-layer stack
  longitude   = "longitude",
  latitude    = "latitude",
  transform   = "scale"
)

# Sheather-Jones plug-in bandwidth as the cell width (one width per
# variable, in z-score units because prepare_bean() scaled the data).
res <- find_env_resolution(
  data     = prepped,
  env_vars = c("mean_prec", "mean_temp", "mean_NDVI"),
  method   = "sheather-jones"
)

# Build the environmental grid using `res$suggested_resolution` and
# randomly retain one of the original occurrences in every occupied pod.
thin_out <- thin_env_nd(
  data            = prepped,
  env_vars        = c("mean_prec", "mean_temp", "mean_NDVI"),
  grid_resolution = res$suggested_resolution
)

# Recover every original column by matching on the coordinate pair.
keep_key <- paste(thin_out$thinned_data$longitude,
                  thin_out$thinned_data$latitude)
sambar_thinned <- sambar[
  paste(sambar$longitude, sambar$latitude) %in% keep_key, , drop = FALSE
]

# Before / after counts per year — verifies that the year column survived.
data.frame(
  year   = sort(unique(sambar$year)),
  before = as.integer(table(sambar$year)),
  after  = as.integer(table(factor(sambar_thinned$year,
                                   levels = sort(unique(sambar$year)))))
)

# Use the thinned table for everything that follows.
sambar <- sambar_thinned
```

Every pod of the temporal-mean E-space now contributes at most one Sambar deer occurrence, so the modelling table no longer over-represents the heavily-surveyed parts of the environment.

## Spatio-temporal rarefaction

Once the per-year thinning has reduced *environmental* duplication, we still want to reduce *spatial* duplication: keep at most one record per raster pixel **per year**, so a repeated visit to the same place in a later year is preserved as a temporally-independent observation.

```{r rarefy}
# Working directory for everything point-related downstream.
points_dir <- "data/processed/points"
dir.create(points_dir, showWarnings = FALSE, recursive = TRUE)

# Persist the thinned table to disk so spatiotemporal_rarefaction() can
# pick it up by file path.
points_csv <- file.path(points_dir, "sambar_temporal.csv")
write_csv(sambar, points_csv)

# Use the CRS of the reference raster so points and rasters live in the
# same coordinate system throughout the pipeline.
study_crs <- st_crs(rast(ref_raster))

rare_out <- spatiotemporal_rarefaction(
  points_sp        = points_csv, # input CSV of presences
  output_dir       = file.path(points_dir, "rarefied"), # where rarefied tables are written
  reference_raster = ref_raster, # defines the pixel grid for rarefaction
  time_cols        = "year", # column carrying the time stamp
  xcol             = "longitude", # column with x coordinate
  ycol             = "latitude", # column with y coordinate
  points_crs       = study_crs, # CRS of the input points
  output_prefix    = "Sambar_annual" # prefix for the written files
)

# The three counts below show how aggressive rarefaction was: how many
# records went in, how many survive pure spatial rarefaction, and how
# many survive spatio-temporal rarefaction (one record per pixel per
# year).
rare_out$input_points # n records that entered rarefaction
rare_out$spatial_points # n surviving pure spatial rarefaction (one per pixel)
rare_out$spatiotemporal_points # n surviving spatio-temporal (one per pixel per year)
```

## Match each observation to the environment it experienced

`temporally_explicit_extraction()` walks the aligned raster directory using `variable_patterns`: clean variable names on the left, filename patterns with time placeholders (`YEAR`) on the right. For each surviving occurrence, it pulls the precipitation, temperature, and NDVI values from the raster of the matching year — so a 2014 record is paired with the 2014 climate, not the long-term average.

```{r extract}
# Output directory for the extracted tables and the scaling parameters.
ext_dir <- file.path(points_dir, "extracted")

ext_out <- temporally_explicit_extraction(
  points_sp           = rare_out$files_created$spatiotemporal, # rarefied points to extract at
  