---
title: "Step 2 — Environmental thinning with **bean**"
subtitle: "ENM-Thailand workshop · Sambar deer (*Rusa unicolor*)"
author: "Workshop materials"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: cosmo
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  dpi       = 100
)
```

# Overview

> **Workshop navigation**
>
> - [**Pre-workshop** — Downloading the workshop data.](https://paanwaris.github.io/ENM-Thailand/articles/Pre-workshop.html) Get the Thailand boundary, WorldClim climate layers, and GBIF Sambar records before the workshop starts.
> - [**Step 1: nicheR** — Virtual species with nicheR.](https://paanwaris.github.io/ENM-Thailand/articles/Step1_nicheR.html) Build a virtual species niche and map it.
> - **Step 2: bean — Environmental thinning with bean.** Reduce environmental sampling bias before fitting a real niche.
> - [**Step 3: TemporalModelR** — Temporally-explicit models with TemporalModelR.](https://paanwaris.github.io/ENM-Thailand/articles/Step3_TemporalModelR.html) Link species occurrences with environmental data through time.

Some occurrence datasets are heavily **biased**: field collecting sample in easily accessible places (roads, protected areas, research stations) far more often than remote interiors, so the records cluster on the map. The classical fix is to **thin in geographic space (G-space)**, keep at most one point per spatial grid cell. That removes spatial duplication, but it has a hidden cost: two records that sit close together on the map can capture very different environmental conditions (one on a ridge, one in a valley), and G-space thinning discards that information.

`bean` solves the problem differently. It thins **in environmental space (E-space)**, keeping at most one point per cell of an environmental grid. The intuition is that two occurrences with nearly identical climate values are *environmentally redundant*, no matter how far apart they are on the map. Thinning per environmental cell (one "pod") therefore produces a more even coverage of the species' niche and a less biased niche estimate.

In this Step we will:

1.  Load the cleaned GBIF occurrences for Sambar deer in Thailand (written by `Pre-workshop.Rmd`).
2.  Prepare the occurrences for `bean` and visualise the bias in E-space.
3.  Find an **objective** environmental grid resolution.
4.  Thin the points with both methods (`thin_env_nd` stochastic, `thin_env_center` deterministic).
5.  Fit an ellipsoid niche on the thinned points and map suitability back to G-space.

All function names and arguments below follow the official package documentation: <https://paanwaris.github.io/bean/>

# Install and load packages

```{r install, eval = FALSE}
# ---- CRAN dependencies -----------------------------------------------------
cran_pkgs <- c(
  "terra",
  "sf",
  "dplyr",
  "readr",
  "rmarkdown",
  "knitr",
  "remotes",
  "rgl",
  "bean"
)
```

```{r packages}
library(bean)     # environmental thinning in E-space
library(terra)    # modern raster + vector spatial backend
library(sf)       # tidy vector spatial data ("simple features")
library(dplyr)    # tidy data manipulation
library(readr)    # fast, friendly CSV reading and writing
library(rgl)      # interactive 3-D rendering for the ellipsoid plots

# rgl::setupKnitr() lets knitr embed any rgl/WebGL plot as a rotatable
# widget directly in the rendered HTML page.
rgl::setupKnitr(autoprint = TRUE)

# Make every random draw in the rest of the file reproducible.
set.seed(2026)
```

```{r load-data}
# Inputs prepared by the Pre-workshop notebook.
bioclim  <- rast("data/processed/bioclim/bioclim_thailand.tif")
thailand <- st_read("data/processed/boundary/thailand_boundary.gpkg",
                    quiet = TRUE)
sambar   <- read_csv("data/processed/gbif/sambar_thailand.csv",
                     show_col_types = FALSE)

head(sambar)
```

For this exercise we use three bioclim variables, which map the Sambar deer habitat (warm, wet, evergreen and mixed forests).

```{r pick-vars}
# Subset the 19-layer bioclim stack down to the three variables that will
# define the ellipsoid axes.
env <- bioclim[[c("bio1", "bio12", "bio15")]]
names(env) <- c("bio1", "bio12", "bio15")
```

# Prepare occurrences

`prepare_bean()` drops rows with missing coordinates, extracts environmental values at each occurrence, optionally rescales the variables, and returns a tidy data frame that is ready for thinning.

```{r prepare}
# Slim down the GBIF table to just longitude/latitude. We rename the GBIF
# coordinate columns at the same time so they match prepare_bean()'s
# expected argument names.
sambar_xy <- sambar %>%
  dplyr::select(longitude = decimalLongitude,
                latitude  = decimalLatitude)

# prepare_bean() returns one row per occurrence with the bioclim values
# extracted at that pixel. With transform = "scale", every environmental
# column is converted to a z-score (mean 0, sd 1) — this puts BIO1, BIO12,
# and BIO15 on the same numerical scale so the thinning grid is fair.
prepped <- prepare_bean(
  data        = sambar_xy,             # the table of coordinates
  env_rasters = env,                   # the SpatRaster to extract from
  longitude   = "longitude",           # name of the longitude column
  latitude    = "latitude",            # name of the latitude column
  transform   = "scale"                # z-score every env variable
)

# Number of usable occurrences after dropping rows that fell outside the
# raster mask or had missing coordinates.
nrow(prepped)
```

Inspect the raw points in G-space and in E-space side by side. Notice the clustering in E-space, that is the environmental over-sampling `bean` will reduce.

```{r raw-bias, fig.width = 9, fig.height = 4.5}
# Two side-by-side panels in one row.
par(mfrow = c(1, 2))

plot(st_geometry(thailand), border = "grey20", main = "G-space")
points(prepped[, c("longitude", "latitude")],
       pch = 21, bg = "tomato", col = "white", cex = 0.7)

plot(prepped$bio1, prepped$bio12,
     xlab = "BIO1 — annual mean T (z-score)",
     ylab = "BIO12 — annual precip. (z-score)",
     pch  = 19,
     col  = adjustcolor("tomato", alpha.f = 0.5),
     main = "E-space")

par(mfrow = c(1, 1))
```

Reading the panels: **left**, every red dot is one Sambar deer GBIF record on the map of Thailand — the dots concentrate on the western mountains and the central plains because that is where surveys happen most. **Right**, the same records replotted in the environmental plane defined by BIO1 (annual mean temperature) and BIO12 (annual precipitation).

# Choose an objective environmental resolution

`find_env_resolution()` derives a grid resolution for thinning directly from the data. It does so **per variable**, meaning it returns *one* suggested cell width for each environmental axis (so BIO1 gets its own width, BIO12 gets a different width, BIO15 gets yet another), each in the natural units of that variable. The widths come from a **kernel-density bandwidth** of each variable's marginal distribution: at the bandwidth scale, two occurrences carry essentially the same environmental information, so it is a natural cell width below which extra records are redundant.

Three classic bandwidth selectors are supported:

- `"sheather-jones"` *(default)* — Sheather & Jones (1991) plug-in selector; robust to non-Gaussian shapes and the modern recommended default.
- `"silverman"` — Silverman's rule of thumb; fast and stable when the distribution is roughly Gaussian.
- `"scott"` — Scott's rule; the Gaussian-optimal variant of Silverman.

If Sheather–Jones fails (this can happen with strongly tied data) the function automatically falls back to Silverman's rule for that variable and emits a message.

We will run all three selectors and compare them, then pick one to feed into the thinning step.

```{r find-res-three}
res_sj       <- find_env_resolution(prepped,
                                    env_vars = c("bio1", "bio12", "bio15"),
                                    method   = "sheather-jones")
res_silverman <- find_env_resolution(prepped,
                                    env_vars = c("bio1", "bio12", "bio15"),
                                    method   = "silverman")
res_scott    <- find_env_resolution(prepped,
                                    env_vars = c("bio1", "bio12", "bio15"),
                                    method   = "scott")

# Side-by-side comparison: one column per selector, one row per variable,
# values in z-score units (we scaled the data in prepare_bean()).
cbind(
  sheather_jones = res_sj$suggested_resolution,
  silverman      = res_silverman$suggested_resolution,
  scott          = res_scott$suggested_resolution
)
```

Read across each row to compare. Silverman and Scott are simple plug-ins that assume a roughly Gaussian distribution, so they usually return *similar*, slightly *larger* bandwidths, coarser cells, fewer pods, more aggressive thinning. Sheather–Jones is data-adaptive and is usually *smaller*, finer cells, more pods, gentler thinning. For real occurrence data, which is almost never Gaussian, we recommend **`"sheather-jones"`**: it preserves more environmental nuance and is the safest default for the workshop.

Each method's diagnostic plot. Each panel below is the kernel-density estimate (KDE) of one bioclim variable; the red horizontal scale bar marks the bandwidth that the selector chose — i.e. the suggested cell width.

```{r find-res-plots-sj, fig.height = 4}
plot(res_sj)
```

```{r find-res-plots-silverman, fig.height = 4}
plot(res_silverman)
```

```{r find-res-plots-scott, fig.height = 4}
plot(res_scott)
```

Reading the panels: the **black curve** is the kernel-density estimate of how the occurrence points are distributed along that environmental axis; the **shaded area** under the curve helps the eye see where the bulk of the density sits; the **red horizontal bar** is the suggested cell width (= the bandwidth) centred on the mode of the density. A narrower bar means a finer grid and gentler thinning; a wider bar means a coarser grid and more aggressive thinning.

We will use Sheather–Jones for the rest of the analysis:

```{r chosen-res}
chosen_res <- res_sj$suggested_resolution
chosen_res
```

# Apply thinning

`bean` ships two thinning strategies:

- **`thin_env_nd()` — stochastic.** Builds the grid, finds every occupied pod, and *randomly retains one of the original occurrences* in each pod. Real coordinates and GBIF metadata are preserved.
- **`thin_env_center()` — deterministic.** Builds the grid and *replaces the points in each occupied pod with