---
title: "Step 1 — Virtual species with **nicheR**"
subtitle: "ENM-Thailand workshop"
author: "Workshop materials"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: cosmo
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  dpi       = 100
)
```

# Overview

> **Workshop navigation**
>
> - [Pre-workshop — Downloading the workshop data.](https://paanwaris.github.io/ENM-Thailand/articles/Pre-workshop.html) Get the Thailand boundary, WorldClim climate layers, and GBIF Sambar records before the workshop starts.
> - **Step 1: nicheR — Virtual species with nicheR.** Build a virtual species niche and map it.
> - [**Step 2: bean** — Environmental thinning with bean.](https://paanwaris.github.io/ENM-Thailand/articles/Step2_bean.html) Reduce environmental sampling bias before fitting a real niche.
> - [**Step 3: TemporalModelR** — Temporally-explicit models with TemporalModelR.](https://paanwaris.github.io/ENM-Thailand/articles/Step3_TemporalModelR.html) Link species occurrences with environmental data through time.

`nicheR` is an R package for building **virtual species' ecological niches** in environmental space (E-space) and projecting them onto geographic space (G-space). An ellipsoid is the simplest closed shape that can summarise a multivariate niche: it has a *centroid* (the preferred conditions), *axes* (the directions of tolerance), and a *volume* (the breadth of the niche).

In this Step we build the niche of a **virtual species** — a hypothetical organism whose tolerances we choose ourselves. Working with a virtual species is the ideal way to learn ENM mechanics because the "true" niche is known by construction, so we can see exactly how each property of the niche manifests on the map:

- **Niche position** (where the centroid sits in E-space) controls *where* on the map the species can occur.
- **Niche breadth** (how wide the axes are) controls *how widely* it can occur.
- **Niche shape** (the orientation and correlation between axes) controls *which combinations* of conditions matter.
- **Environmental availability** (the cloud of conditions Thailand actually offers) decides whether a niche can be expressed at all — a perfectly defined niche over conditions that do not exist in Thailand will simply produce an empty map.

In this Step we will:

1.  Explore the environmental space of Thailand in **three dimensions** using BIO1, BIO12, and BIO15.
2.  Build an ellipsoid from a user-defined range of bioclimatic tolerances.
3.  Project the ellipsoid to Thailand to obtain *suitability* and *Mahalanobis distance* maps.
4.  Add a sampling-bias layer that mimics uneven survey effort.
5.  Generate virtual occurrence points under three sampling strategies (`centroid`, `random`, `edge`) plus a bias-weighted draw.

All function names and arguments below follow the official package documentation: <https://castanedam.github.io/nicheR/>

# Install and load packages

The chunk below installs everything Step 1 needs the first time you run it. Re-running is safe — already-installed packages are skipped.

```{r install, eval = FALSE}
# ---- CRAN dependencies -----------------------------------------------------
cran_pkgs <- c(
  "terra",
  "sf",
  "plotly",
  "viridis",
  "rmarkdown",
  "knitr",
  "remotes"
)
to_install <- cran_pkgs[!cran_pkgs %in% rownames(installed.packages())]
if (length(to_install)) install.packages(to_install, dependencies = TRUE)

# ---- nicheR from GitHub ----------------------------------------------------
if (!requireNamespace("nicheR", quietly = TRUE)) {
  remotes::install_github("castanedaM/nicheR", upgrade = "never")
}
```

```{r packages}
library(nicheR)   # build / project / sample ellipsoid niches
library(terra)    # modern raster + vector spatial backend
library(sf)       # tidy vector spatial data ("simple features")
library(plotly)   # interactive 3-D plots for the environmental space

# Make every random draw in the rest of the file reproducible.
set.seed(2026)
```

# Data

For Step 1 we work with the bioclim rasters and the Thailand polygon that we produced in the Pre-workshop. **Please knit `Pre-workshop.Rmd` first** so these files exist under `data/processed/`.

The two file extensions you will see in this folder are:

- **`.tif` (GeoTIFF)** — a raster format storing pixel grids of continuous values (e.g. temperature, precipitation). Our `bioclim_thailand.tif` is a single multi-band GeoTIFF that bundles all 19 bioclim layers together.
- **`.gpkg` (GeoPackage)** — a modern, single-file format for vector data (points, lines, polygons). Our `thailand_boundary.gpkg` carries the country polygon and its attribute table.

```{r load-data}
bio_path <- "data/processed/bioclim/bioclim_thailand.tif"
tha_path <- "data/processed/boundary/thailand_boundary.gpkg"

# rast()    reads a multi-band GeoTIFF as a SpatRaster (one object, many layers).
# st_read() reads a vector file as an sf object (a data frame whose last
#           column is a geometry column).
bioclim  <- rast(bio_path)
thailand <- st_read(tha_path, quiet = TRUE)

# Draw just the country outline — st_geometry() strips the attribute table
# so we get the polygon without any colour fill.
plot(st_geometry(thailand))

# 19 layer names: bio1, bio2, ..., bio19.
names(bioclim)

# Per-layer statistics. Each column is one bioclim layer; each row is a
# summary statistic:
#   Min / Max   — full range of pixel values in raw units (°C for bio1,
#                 mm for bio12, etc.)
#   1st Qu / Median / 3rd Qu — the 25th, 50th, and 75th percentiles, i.e.
#                 how the values are distributed across pixels
#   Mean        — arithmetic average across all pixels
#   NA's        — number of pixels masked out (outside Thailand)
summary(bioclim)
```

For this exercise we will use only three of the 19 bioclim variables that will draw the 3-D environmental space:

- **bio1** — annual mean temperature (°C)
- **bio12** — annual precipitation (mm)
- **bio15** — precipitation seasonality (coefficient of variation, CV)

```{r pick-vars}
# Subset the 19-layer raster down to the three variables we care about.
# `env_space` is a SpatRaster (used for spatial projection later on);
# `env_space_df` is a plain data frame of the *valid* pixel values
# (na.rm = TRUE drops pixels masked outside Thailand) — handy as the
# background cloud in 2-D and 3-D scatterplots.
env_space    <- bioclim[[c("bio1", "bio12", "bio15")]]
env_space_df <- as.data.frame(env_space, na.rm = TRUE)

plot(env_space)
```

# Explore Thailand's environmental space in 3D

Before defining a niche, we will visualize the *available* environmental conditions of Thailand in three dimensions: `BIO15` (precipitation seasonality, CV) captures within-year rainfall variability and adds an important third axis on top of mean temperature (`BIO1`) and total annual precipitation (`BIO12`). Each grey dot in the 3-D cloud below is one pixel of the country — together they outline the *envelope* of conditions Thailand offers a species.

```{r env-3d}
plotly::plot_ly(
  data   = env_space_df,
  x      = ~bio1, y = ~bio12, z = ~bio15,
  type   = "scatter3d", mode = "markers",
  marker = list(size = 2, opacity = 0.5)
) |>
  plotly::layout(scene = list(
    xaxis = list(title = "BIO1 (°C)"),
    yaxis = list(title = "BIO12 (mm)"),
    zaxis = list(title = "BIO15 (CV)")
  ))
```

# Define a virtual species' ecological niche

A `nicheR` ellipsoid is built from a **range data frame** that lists the lowest and highest value the species tolerates along each environmental axis. From these ranges, `build_ellipsoid()` derives every other geometric property of the niche: the centroid (the midpoint of the ranges), a covariance matrix (how axes co-vary), the semi-axis lengths (the half-width of the ellipsoid in each direction), and the niche volume.

For our virtual tropical mammal we set:

- **bio1 (mean annual T)** between **25 °C and 28 °C** — a warm, narrow thermal niche.
- **bio12 (annual precip.)** between **1200 mm and 2500 mm** — wet but not extreme.
- **bio15 (precip. seasonality, CV)** between **70 and 90** — moderately to strongly seasonal rainfall.

Together these three windows describe a small, well-defined ellipsoid sitting in the lowland, seasonally-wet corner of Thailand's environmental space.

```{r build-ellipsoid}
range_df <- data.frame(
  bio1  = c(25, 28),       # tolerated range of mean annual temperature  (°C)
  bio12 = c(1200, 2500),   # tolerated range of annual precipitation     (mm)
  bio15 = c(70, 90)        # tolerated range of precipitation seasonality (CV)
)

# build_ellipsoid() converts the range table into a full ellipsoid object:
# centroid, covariance matrix, semi-axes, volume.
ell <- build_ellipsoid(range = range_df)
ell
```

Printing `ell` shows a compact summary of the niche. Reading the output line by line:

- **Dimensions: 3D.** Three environmental axes, so the ellipsoid is a true 3-D shape (an "egg" in BIO1 × BIO12 × BIO15 space).
- **Chi-square cutoff: 11.345.** The 95% probability boundary of the niche, taken from the chi-square distribution with 3 degrees of freedom. Pixels with a squared Mahalanobis distance below this cutoff fall *inside* the niche; the rest are *outside*.
- **Centroid (mu): (26.5, 1850, 80).** The geometric centre of the ranges — the *most preferred* environmental conditions.
- **Covariance matrix.** A diagonal matrix because the three ranges were specified independently — each axis varies on its own and the cross-axis covariances are zero.
- **Ellipsoid semi-axis lengths: 729.78, 11.227, 1.684.** The half-widths of the ellipsoid in each direction. Notice how stretched the bio12 axis is (precipitation values span hundreds of mm) compared with bio15 (CV values span tens) — the ellipsoid is *very* elongated along precipitation.
- **Ellipsoid axis endpoints.** The two vertex coordinates at the ends of each axis — useful for sanity-checking that the ellipsoid actually reaches the ranges you specified.
- **Ellipsoid volume: 57 800.1.** The total amount of E-space the niche occupies, in mixed units. Useful for comparing niche breadths between species.

We can visualise the ellipsoid in E-space against the cloud of environmental conditions Thailand actually offers — the *background point cloud*. Every grey dot is one pixel of the country plotted in BIO1 × BIO12 (the third axis, BIO15, is collapsed in this 2-D view). A niche that sits inside the cloud can be realised geographically; a niche that sits outside the cloud will produce an empty suitability map.

```{r plot-ellipsoid}
plot_ellipsoid(ell, background = env_space_df)
```

The ellipse in the plot is the 2-D projection of the niche onto the BIO1 × BIO12 plane. The grey dots are the background point cloud. Pixels that fall *inside* the ellipse correspond to places on the map where the virtual species can occur; pixels outside the ellipse are environmentally unsuitable.

# Project the niche to geography (suitability + Mahalanobis distance)

`predict()` projects the ellipsoid from E-space onto a `SpatRaster` of environmental conditions (here, our Thailand `env_space` stack) and returns two surfaces, layer by layer:

- **Suitability.** A 0–1 score that peaks at 1 at the centroid and falls off smoothly toward the boundary. When `suitability_truncated = TRUE`, pixels outside the niche are clamped to 0, giving a cleaner "in or out" answer in a layer called `suitability_trunc`.
- **Mahalanobis distance.** A multivariate distance from the centroid that takes the shape and correlations of the niche into account. Low = environmentally similar to the centre of the niche; high = far from it. A truncated version (`Mahal_trunc`) is also produced.

In short, *suitability* is the probability-like score you would plot on a map, and *Mahalanobis distance* is the raw geometric distance used to compute it.

```{r predict}
pred <- predict(
  ell,                                # the ellipsoid we just built
  newdata               = env_space,  # SpatRaster to project onto
  include_suitability   = TRUE,       # return the suitability surface
  suitability_truncated = TRUE,       # also return a clamped (in/out) version
  include_mahalanobis   = TRUE,       # return the Mahalanobis-distance surface
  mahalanobis_truncated = TRUE,       # also return a clamped version
  keep_data             = TRUE        # keep the original env layers in `pred`
)

# pred is now a multi-layer SpatRaster carrying: the three predictors
# (bio1, bio12, bio15), the two suitability layers (full + truncated), and
# the two Mahalanobis layers (full + truncated).
pred
```

The printout describes `pred` as a `SpatRaster` with 89 × 50 pixels and 7 layers. The `min values` / `max values` rows give the range of each layer across all pixels.

```{r predict-plot}
plot(pred, col = viridis::viridis(100))

# A "binary potential distribution" is just the truncated suitability
# turned into 1 / 0: 1 wherever the niche is realised, 0 elsewhere.
binary_rast <- (pred$suitability_trunc > 0) * 1

plot(binary_rast,
     axes = FALSE,                              # hide longitude/latitude tick labels
     box  = TRUE,                               # keep the surrounding frame
     col  = c("#d4d4d4", "#0004d5"),            # grey = unsuitable, blue = suitable
     main = "Potential Distribution (Binary)")
```

# Add a sampling-bias layer

Real-world surveys are uneven: roads, cities, protected areas, and research stations get visited far more often than remote interiors. We can simulate this unevenness with a **bias surface** and combine it with our suitability to obtain a *realised* sampling surface, what we would actually expect to record if surveys happened in proportion to where field effort is concentrated.

For this lesson we use the two proxy rasters we wrote in the Pre-workshop:

- **Species richness** — the number of recorded species per pixel. Many recorded species in a pixel implies that many surveys have been done there. We treat richness as a **direct** proxy: higher richness → more sampling effort.
- **Night-time lights** — a satellite measure of artificial light. High light values mark cities and roads. Because remote Sambar deer habitat is *dark*, we treat night-time lights as an **inverse** proxy: more light → *less* relevant sampling effort for our forest-dwelling virtual species.

```{r bias-layer}
sp_rich_layer   <- rast("data/processed/bias/sp_richness_TH.tif")
nighttime_layer <- rast("data/processed/bias/nighttime_TH.tif")

# Stack the two proxies into a 2-layer SpatRaster and rename for clarity.
bias_layer <- c(sp_rich_layer, nighttime_layer)
names(bias_layer) <- c("sp_richness", "nighttime_light")
plot(bias_layer)

# prepare_bias() rescales each proxy to 0-1, optionally inverts proxies whose
# *high* values represent *low* sampling effort, and combines the two into a
# single composite surface. `effect_direction = c("direct", "inverse")`
# tells it to use richness directly but invert night-time lights.
bias_prep <- prepare_bias(
  bias_surface     = bias_layer,
  effect_direction = c("direct", "inverse")
)

# The composite surface is a 0-1 map of "how likely a survey would happen
# here": high values mark accessible, biodiversity-rich pixels; low values
# mark remote, brightly-lit-but-empty pixels (essentially, urban areas).
plot(bias_prep$composite_surface)
```

We then **multiply** the suitability by the bias surface to get a realised sampling probability. Pixels that are both suitable for the species *and* easy to survey light up; pixels that are biologically suitable but in remote areas dim down.

```{r apply-bias, fig.height = 4.5}
biased_pred <- apply_bias(
  prepared_bias    = bias_prep,                 # the composite bias surface from prepare_bias()
  prediction       = pred,                      # the projected ellipsoid from predict()
  prediction_layer = "suitability",             # which layer of `pred` to multiply against bias
  effect_direction = "direct"                   # high bias → high realised sampling
)

par(mfrow = c(1, 2))
plot(pred$suitability,
     main = "Unbiased suitability")
plot(biased_pred$suitability_biased,
     main = "Suitability × bias = realised sampling surface")
par(mfrow = c(1, 1))
```

The biological niche is identical in both maps, but the realised-sampling surface concentrates around accessible, biodiversity-rich pixels of Thailand — only those parts of the niche would realistically generate records in a biased field survey.

# Generate virtual occurrences under three sampling strategies

So far we have a *niche* but no occurrences. To compare ENM methods later in **Step 2: bean** and **Step 3: TemporalModelR**, we draw virtual records from the niche. Generating virtual occurrences is essential to ENM teaching because — unlike with real species — we know the underlying truth and can measure how well any method recovers it.

`nicheR::sample_data()` lets us choose **where in the niche** the simulated points are drawn from:

- `"centroid"` — points weighted toward the niche centroid (preferred conditions). Mimics surveys that find a species near the *heart* of its range.
- `"random"` — points drawn uniformly from inside the niche. The neutral reference.
- `"edge"` — points weighted toward the *periphery* of the niche (marginal conditions). Mimics surveys at the species' range edges.

To ensure every simulated point falls **inside** the ellipsoid, we sample from the *truncated* suitability layer (`suitability_trunc`), which is 0 outside the niche by construction.

```{r sampling-three}
pts_centroid <- sample_data(
  n_occ            = 50,                  # number of points to draw
  prediction       = pred,                # SpatRaster we projected onto
  sampling         = "centroid",          # weight toward the centre of the niche
  prediction_layer = "suitability_trunc", # truncated → points fall *inside* the ellipsoid only
  seed             = 123                  # reproducible random draws
)

pts_random   <- sample_data(
  n_occ            = 50,
  prediction       = pred,
  sampling         = "random",            # uniform sampling inside the niche
  prediction_layer = "suitability_trunc",
  seed             = 123
)

pts_edge     <- sample_data(
  n_occ            = 50,
  prediction       = pred,
  sampling         = "edge",              # weight toward the periphery of the niche
  prediction_layer = "suitability_trunc",
  seed             = 123
)

# Each returned table carries the (x, y) coordinate of the simulated point
# and the environmental values extracted at that pixel.
head(pts_centroid)
head(pts_random)
head(pts_edge)
```

A **bias-weighted** draw uses `sample_biased_data()` together with the realised sampling surface from above. The result mimics what a *biased* field survey would record — samples cluster where the species can occur **and** where surveys actually happen.

```{r biased-sample}
biased_pts <- sample_biased_data(
  n_occ            = 50,                            # number of points to draw
  prediction       = biased_pred,                   # SpatRaster from apply_bias()
  prediction_layer = "suitability_biased_direct",   # the layer apply_bias() produced
  seed             = 123
)

head(biased_pts)

# Pull the BIO1 / BIO12 / BIO15 values at each biased point so we can
# overlay them in E-space alongside the other three samples below.
occ_bias_env <- terra::extract(env_space, biased_pts[, c("x", "y")])
```

## Geographic view — where each strategy puts the points

```{r plot-samples, fig.height = 6, fig.width = 9}
par(mfrow = c(2, 2))

plot(pred[["suitability"]],
     main = "Centroid sample (n = 50)")
points(pts_centroid[, c("x", "y")], pch = 19, cex = 0.6, col = "red")

plot(pred[["suitability"]],
     main = "Random sample (n = 50)")
points(pts_random[, c("x", "y")], pch = 19, cex = 0.6, col = "red")

plot(pred[["suitability"]],
     main = "Edge sample (n = 50)")
points(pts_edge[, c("x", "y")], pch = 19, cex = 0.6, col = "red")

plot(biased_pred$suitability_biased,
     main = "Bias-weighted sample (n = 50)")
points(biased_pts[, c("x", "y")], pch = 19, cex = 0.6, col = "red")

par(mfrow = c(1, 1))
```

Each panel shows the projected suitability surface in viridis (yellow = high suitability, dark = low), with the simulated 50 occurrences in **red**. Reading across the panels:

- The **centroid** sample clusters around the *core* of the niche on the map.
- The **random** sample is scattered evenly *inside* the suitable area.
- The **edge** sample drifts toward the *boundary* of the suitable area.
- The **bias-weighted** sample collapses onto the *accessible* part of the niche — the parts of the country where surveys would realistically happen.

## Environmental-space view — same points, different colours

This time we draw **one panel per strategy** so the contrast between sampling rules is obvious. Each panel shows the same ellipsoid (the niche) and the same grey background cloud (Thailand's available conditions), with the 50 simulated points of that strategy overlaid as coloured markers.

```{r samples-in-espace, fig.height = 7, fig.width = 9}
par(mfrow = c(2, 2))

plot_ellipsoid(ell, background = env_space_df)
title(main = "Centroid sample")
points(pts_centroid[, c("bio1", "bio12")],
       col = "blue", pch = 19, cex = 0.6)

plot_ellipsoid(ell, background = env_space_df)
title(main = "Random sample")
points(pts_random[, c("bio1", "bio12")],
       col = "darkgreen", pch = 17, cex = 0.6)

plot_ellipsoid(ell, background = env_space_df)
title(main = "Edge sample")
points(pts_edge[, c("bio1", "bio12")],
       col = "purple", pch = 15, cex = 0.6)

plot_ellipsoid(ell, background = env_space_df)
title(main = "Bias-weighted sample")
points(occ_bias_env[, c("bio1", "bio12")],
       col = "red", pch = 4, cex = 0.6)

par(mfrow = c(1, 1))
```

Reading the panels: the ellipse is the 2-D projection of the niche; the grey cloud is Thailand's available environmental conditions; the coloured markers are the 50 simulated records of each strategy. The centroid points hug the centre of the ellipse, the random points scatter evenly inside it, the edge points fringe the boundary, and the bias-weighted points concentrate on the side of the niche that overlaps the accessible, biodiversity-rich part of the country.

# Wrap-up

What we just did:

- Defined a niche *a priori* as a 3-D ellipsoid in BIO1 × BIO12 × BIO15 space.
- Explored Thailand's available environmental space in 3D.
- Projected the ellipsoid back to geography and inspected the suitability and Mahalanobis-distance surfaces.
- Built a sampling-bias surface from species richness and night-time lights and combined it with suitability to obtain a realised sampling surface.
- Compared three sampling strategies (`centroid`, `random`, `edge`) plus a bias-weighted draw, both geographically and in environmental space.

# References

- Castaneda-Guzman M, Hughes C, Paansri P, Cobos M (2026). *nicheR: Ellipsoid-Based Virtual Niches and Visualization*. R package version 0.1.0, <https://github.com/castanedaM/nicheR>.
- Chamberlain S, Barve V, Mcglinn D, Oldoni D, Desmet P, Geffert L, Ram K (2026). *rgbif: Interface to the Global Biodiversity Information Facility API*. R package version 3.8.5.2, <https://CRAN.R-project.org/package=rgbif>.
- Hijmans R (2026). *geodata: Access Geographic Data*. R package version 0.6-10, <https://rspatial.github.io/geodata/>.
- Hughes C, Castaneda-Guzman M, E. Escobar L (2026). *TemporalModelR: Temporally Explicit Species Distribution Modelling in R*. R package version 0.2.0, <https://github.com/CJHughes926/TemporalModelR>.
- Paansri P, Escobar L (2026). *bean: Data Thinning of Species Occurrences in Environmental Space*. R package version 0.2.1, <https://github.com/paanwaris/bean>.
