---
title: "Day 2a — Downloading the workshop data"
subtitle: "ENM-Thailand workshop · Sambar deer (*Rusa unicolor*)"
author: "Workshop materials"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
    theme: cosmo
    df_print: paged
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo      = TRUE,
  message   = FALSE,
  warning   = FALSE,
  fig.align = "center",
  fig.width = 7,
  fig.height = 5,
  dpi       = 100
)

# This Rmd lives in notebooks/, but every path inside it is written relative
# to the repository root. Set the knit working directory to the project root
# so the file knits cleanly whether you click "Knit" in RStudio, render it
# from R, or build the pkgdown site.
if (requireNamespace("rprojroot", quietly = TRUE)) {
  knitr::opts_knit$set(
    root.dir = rprojroot::find_root(rprojroot::has_file("DESCRIPTION"))
  )
} else {
  knitr::opts_knit$set(root.dir = normalizePath(".."))
}

# Day 2a is the data-fetching notebook itself. When you build the public
# pkgdown site we don't actually want to re-hit GADM / WorldClim / GBIF; we
# just want the code visible. Detect that mode and skip evaluation.
.fetch_live <- !nzchar(Sys.getenv("PKGDOWN_BUILD"))
knitr::opts_chunk$set(eval = .fetch_live)
```

```{r data-status, echo = FALSE, results = "asis", eval = TRUE}
if (!.fetch_live) {
  cat(
    '<div class="alert alert-info" style="padding:1em;border-left:4px solid #1B6E3F;background:#E8F5E9;">',
    '<strong>Build mode:</strong> the live download chunks are <em>shown but ',
    'not executed</em> on the published website. Knit this file locally with ',
    'your GBIF credentials in <code>.Renviron</code> to actually populate ',
    '<code>data/processed/</code>.',
    '</div>\n\n'
  )
}
```

# Overview

> **Workshop navigation —** part 2 of 4 in the [ENM-Thailand 2026 workshop](https://paanwaris.github.io/ENM-Thailand/).
>
> - [Day 1 — Virtual species with nicheR.](https://paanwaris.github.io/ENM-Thailand/articles/Day1_nicheR.html) Build an ellipsoid niche by hand and project it.
> - **Day 2a (this page) — Downloading the workshop data.** Fetch Thailand, WorldClim and GBIF Sambar records.
> - [Day 2b — Environmental thinning with bean.](https://paanwaris.github.io/ENM-Thailand/articles/Day2b_Bean_Processing.html) Reduce E-space sampling bias before fitting a real niche.
> - [Day 3 — Temporally-explicit SDMs with TemporalModelR.](https://paanwaris.github.io/ENM-Thailand/articles/Day3_TemporalModelR.html) Pair each occurrence with the environment it experienced.

This notebook gathers every input the rest of the workshop needs. Knit it once at the start of Day 2 — everything else (Day 1, Day 2b, Day 3) reads from `data/processed/`.

What we will download / prepare:

1. **Thailand national boundary** (GADM v4.1).
2. **WorldClim v2.1 bioclim** (10 arc-min, 19 layers, cropped + masked to Thailand).
3. **GBIF occurrence records** for Sambar deer (*Rusa unicolor*) in Thailand, strictly filtered to `basisOfRecord %in% c("HUMAN_OBSERVATION", "OCCURRENCE", "OBSERVATION")`.
4. Two bias proxies (**species richness**, **night-time lights**) used by Day 1.

# Install and load packages

```{r install, eval = FALSE}
cran_pkgs <- c(
  "terra",
  "sf",
  "geodata",
  "rgbif",
  "dplyr",
  "readr",
  "rmarkdown",
  "knitr",
  "remotes"
)
to_install <- cran_pkgs[!cran_pkgs %in% rownames(installed.packages())]
if (length(to_install)) install.packages(to_install, dependencies = TRUE)
```

```{r packages}
library(terra)
library(sf)
library(geodata)
library(rgbif)
library(dplyr)
library(readr)
```

```{r paths}
raw_dir       <- "data/raw"
processed_dir <- "data/processed"

# data/processed/ is organised into role-based subfolders.
boundary_dir <- file.path(processed_dir, "boundary")   # Thailand polygon
bioclim_dir  <- file.path(processed_dir, "bioclim")    # WorldClim layers
bias_dir     <- file.path(processed_dir, "bias")       # sampling-bias proxies (Day 1)
gbif_dir     <- file.path(processed_dir, "gbif")       # GBIF Sambar records

for (d in c(raw_dir, processed_dir,
            boundary_dir, bioclim_dir, bias_dir, gbif_dir)) {
  dir.create(d, recursive = TRUE, showWarnings = FALSE)
}
```

# Step 1 — Thailand boundary (GADM)

```{r tha-boundary}
tha <- gadm(country = "THA", level = 0, path = raw_dir)
tha_sf <- st_as_sf(tha)

st_write(tha_sf,
             file.path(boundary_dir, "thailand_boundary.gpkg"),
             delete_dsn = TRUE, quiet = TRUE)

plot(st_geometry(tha_sf), main = "Thailand boundary (GADM level 0)")
```

# Step 2 — WorldClim bioclim (10 arc-min)

```{r bioclim}
bio_global <- worldclim_global(var = "bio", res = 10, path = raw_dir)

bio_tha <- crop(bio_global, terra::ext(tha))
bio_tha <- mask(bio_tha, tha)

# Tidy names: wc2.1_10m_bio_1 -> bio1
names(bio_tha) <- paste0("bio", seq_len(terra::nlyr(bio_tha)))

writeRaster(
  bio_tha,
  filename  = file.path(bioclim_dir, "bioclim_thailand.tif"),
  overwrite = TRUE
)

bio_tha <- rast(file.path(bioclim_dir, "bioclim_thailand.tif"))
plot(bio_tha[[c("bio1", "bio12", "bio15")]],
     main = c("BIO1 — mean T (°C)",
              "BIO12 — annual precip. (mm)",
              "BIO15 — precip. seasonality (CV)"))
```

# Step 3 — GBIF occurrences for Sambar deer

We pull a fresh GBIF download of *Rusa unicolor* in Thailand. The pipeline is **strictly filtered** so only the following `basisOfRecord` values are retained:

- `HUMAN_OBSERVATION`
- `OCCURRENCE`
- `OBSERVATION`

These are the field-observation types most relevant for ENM/SDM. Fossil, machine, material-sample, and preserved-specimen records are excluded.

> **Credentials.** Never commit real passwords to git. The chunk below reads
> from environment variables; populate them once in your user `.Renviron`
> (`usethis::edit_r_environ()`):
>
> ```
> GBIF_USER  = "your_gbif_username"
> GBIF_EMAIL = "your_email@example.com"
> GBIF_PWD   = "your_gbif_password"
> ```
>
> Restart R so the values are loaded. If any variable is missing, the chunk
> falls back to obvious placeholder strings so the source you commit to GitHub
> never reveals your credentials.

```{r gbif-creds}
options(
  gbif_user  = Sys.getenv("GBIF_USER",  unset = "[YOUR_GBIF_USERNAME]"),
  gbif_email = Sys.getenv("GBIF_EMAIL", unset = "[YOUR_EMAIL]"),
  gbif_pwd   = Sys.getenv("GBIF_PWD",   unset = "[YOUR_GBIF_PASSWORD]")
)

# If no real credentials are present (e.g. fresh clone, GitHub Actions),
# skip the GBIF download chunks so the document still builds. The code is
# still displayed.
.have_gbif_creds <- nzchar(Sys.getenv("GBIF_USER")) &&
                    nzchar(Sys.getenv("GBIF_EMAIL")) &&
                    nzchar(Sys.getenv("GBIF_PWD"))
```

```{r gbif-download, eval = .have_gbif_creds}
key <- name_backbone(name = "Rusa unicolor")$usageKey

allowed_basis <- c("HUMAN_OBSERVATION", "OCCURRENCE", "OBSERVATION")

raw_list <- lapply(allowed_basis, function(b) {
  rgbif::occ_search(
    taxonKey            = key,
    country             = "TH",
    basisOfRecord       = b,
    hasCoordinate       = TRUE,
    hasGeospatialIssue  = FALSE,
    limit               = 5000
  )$data
})

raw <- dplyr::bind_rows(raw_list)

# Belt-and-braces: enforce the filter again in case the API loosened it
raw <- raw %>% filter(basisOfRecord %in% allowed_basis)

write_csv(raw, file.path(gbif_dir, "sambar_gbif.csv"))

nrow(raw)
```

## Clean and write the analysis-ready table

```{r gbif-clean, eval = .have_gbif_creds}
sambar <- raw %>%
  select(species,
         decimalLongitude, decimalLatitude,
         year, month,
         basisOfRecord, gbifID) %>%
  filter(!is.na(decimalLongitude),
                !is.na(decimalLatitude),
                basisOfRecord %in% allowed_basis)

sambar_sf <- st_as_sf(sambar,
                          coords = c("decimalLongitude", "decimalLatitude"),
                          crs    = 4326, remove = FALSE)

# Keep only points that actually fall inside Thailand
inside <- st_within(sambar_sf, tha_sf, sparse = FALSE)[, 1]
sambar <- sambar[inside, ]

write_csv(sambar, file.path(gbif_dir, "sambar_thailand.csv"))

plot(st_geometry(tha_sf), border = "grey20",
     main = "Sambar deer occurrences")
points(sambar$decimalLongitude, sambar$decimalLatitude,
       pch = 21, bg = "tomato", col = "white", cex = 0.7)
```

Everything Day 1, Day 2b, and Day 3 need now lives in `data/processed/`. Day 3 additionally relies on the local `temporal_rasters/` folder, which holds pre-prepared annual precipitation and temperature rasters for 2000–2025 (this folder is provided with the workshop materials).